### 2025-01-23 manuscript.tex
rename and add action

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{orcidlink} 
\usepackage[super,comma,sort&compress]  
   {natbib}\bibliographystyle{numbered}
% \usepackage[right]{lineno} \linenumbers

\makeatletter
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
  \textbf{\@title}

  \@author
\end{flushleft}\egroup}
\makeatother

%%%  Insert title below; leave date empty.

\title{Title is a maximum of 145 characters, including spaces, and avoids the use of jargon and uncommon abbreviations}
\date{}

%%%  Author first and last names should be spelled 
%%%  out in their entirety (do not abbreviate "J.H. 
%%%  Watson" unless this is how the author's name 
%%%  always appears). Middle initials are OK. Do 
%%%  not include titles, positions, or degrees.

%%%  Use numbered footnotes to indicate institutional 
%%%  affiliations. Authors may have multiple 
%%%  institutional affiliations, and affiliations 
%%%  may be shared among multiple authors.

%%%  After the institutional affiliations, numbered 
%%%  footnotes may be used to indicate an author's 
%%%  present address, equal contribution status, 
%%%  and/or senior author status. Corresponding 
%%%  authors may additionally include Twitter (X) 
%%%  handles as a means of contact.

%%%  The final numbered footnote should indicate 
%%%  which author is the lead contact (required). 
%%%  One author must be designated as the lead contact. 
%%%  There can be no more than one lead contact.

%%%  Corresponding authors should be indicated with 
%%%  asterisks (*). Use 2 asterisks (**) for the 
%%%  second-listed corresponding author, 3 (***) for 
%%%  the third-listed, and so on. The lead contact 
%%%  must be a corresponding author. Additional 
%%%  authors may also serve as corresponding authors.

\author[1,5,\orcidlink{0000-0000-0000-0000}]{First Last}
\author[1,2,5]{First Middle Last}
\author[2,3]{First M.M. Last}
\author[3,*]{First Last, Jr.}
\author[3,4,6,7,8,**]{F. Middle Last}

%%%  Institutional affiliations should contain the 
%%%  following information at minimum: department(s)/
%%%  subunit(s), institution, city, state/region (if 
%%%  applicable), and country.

\affil[1]{University A, London SW7 2AZ, UK}
\affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
\affil[3]{Division C, Department C, University C, Cambridge, MA, USA}

\affil[4]{Present address: 123 Sesame Street, New York, NY, USA}
\affil[5]{These authors contributed equally}
\affil[6]{Senior author}
\affil[7]{Twitter: @CorrespondingAuthor}
\affil[8]{Lead contact}

%%%  List only one email address per corresponding author.

\affil[*]{Correspondence: first.last.jr@university.edu}
\affil[**]{Correspondence: f.middle.last@university.edu}

\begin{document}

\maketitle

\section*{SUMMARY}

The summary (abstract) should consist of a single paragraph of 150 words or fewer.

\section*{KEYWORDS}

%%%  Include up to 10 keywords, separated by commas. 
%%%  Keywords entered in EM are not carried over; only 
%%%  keywords included in the main text will be used 
%%%  in the final article metadata. Please note that 
%%%  for some journals, keywords are chosen by the editors.

cell, press, LaTeX, template

\section*{INTRODUCTION}

No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.

\section*{RESULTS}

The results may be divided with subheadings. In our view, good subheadings convey information about the findings, so we encourage you to be specific. For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony. Please do not use numbered subheadings. For some formats, ``results and discussion" may be combined under a single heading.

\subsection*{Figures}

For initial submissions, figures and their titles and legends can be run-in with the text. For final submissions, the image files must be uploaded as separate files, one complete file per figure. The figure titles and legends should be placed toward the end of the main text, after the declaration of interests. Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.

\subsection*{Citations}

To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest. Please note that we do not number sections or subsections.

To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."

\subsection*{Equations}

Simple formulae should appear in line with the text whenever possible. You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\). You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.

Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:

\[ x^n + y^n = z^n \]

\begin{displaymath}
\sqrt{x^2+1}
\end{displaymath}

If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.

\begin{equation}
f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
\end{equation}

\section*{DISCUSSION}

The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.

\subsection*{Limitations of the study}

A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.

\newpage

\noindent{The following Cell Press journals require most research articles to include their methods in a \textbf{METHODS} section, which should appear after the discussion: \textit{Cell Biomaterials}, \textit{Chem}, \textit{Chem Catalysis}, \textit{Cell Reports Physical Science}, \textit{Cell Reports Sustainability}, \textit{Device}, \textit{Joule}, \textit{Matter}, \textit{Newton}, \textit{One Earth}, and \textit{Patterns}. For all other Cell Press journals, please \textbf{delete} the contents of this page from the template and refer instead to \textbf{STAR Methods} on page 9.}

\section*{METHODS}

%%%  The methods (also/formerly called "experimental 
%%%  procedures") should appear 
%%%  immediately after the discussion. Subheadings 
%%%  may be customized. Please consult your handling 
%%%  editor if you have questions about what content 
%%%  should appear here.

\subsection*{Custom methods subheading 1}
\subsection*{Custom methods subheading 2}
\subsection*{Custom methods subheading 3}
\subsection*{Custom methods subheading 4}

\newpage

%%%  The following components should appear after the 
%%%  methods. 
%%%  For journals using STAR Methods, these components 
%%%  should appear immediately after the discussion 
%%%  (after any "limitations" or "conclusions" subsection 
%%%  within the discussion).

\section*{RESOURCE AVAILABILITY}

%%%  The resource availability section is required 
%%%  for all research articles. This component 
%%%  has 3 subsections: "lead contact," "materials 
%%%  availability," and "data and code availability." 
%%%  All 3 subsections must be included, even if no 
%%%  unique materials were generated in the study. 
%%%  Do not edit or change the names of the subheadings. 
%%%  No other subheadings or text are allowed in the 
%%%  resource availability section.

\subsection*{Lead contact}

%%%  Authors are required to designate a lead contact, 
%%%  who will be responsible for communication with 
%%%  the journal before and after publication and is 
%%%  the arbiter of disputes, including concerns 
%%%  related to reagents or resource sharing. Only 
%%%  one author can be named the lead contact, and 
%%%  only the lead contact’s information may be 
%%%  provided in this section.

Requests for further information and resources should be directed to and will be fulfilled by the lead contact, Sally White (s.white@university.edu).

\subsection*{Materials availability}

%%%  This subsection must include a statement describing 
%%%  the availability of newly generated materials 
%%%  associated with the paper, including any conditions 
%%%  for access. If there are no newly generated materials 
%%%  associated with the paper, the statement should 
%%%  state this, e.g.: This study did not generate new 
%%%  materials.

Plasmids generated in this study have been deposited to [Addgene, name and catalog number or unique identifier].

\subsection*{Data and code availability}

%%%  All original research papers must include a 
%%%  comprehensive and accurate “data and code 
%%%  availability” statement within the “resource 
%%%  availability” component of the paper before it 
%%%  is accepted for publication. These statements 
%%%  are structured and consist of three bulleted 
%%%  components. Each component must be present.

\begin{itemize}
    \item Small-molecule crystallography data have been deposited at CCDC under the database identifier ABCDEF and are publicly available as of the date of publication. 1H NMR spectra have been deposited at Mendeley under the DOI 10.12345/a1b2c3d4e5f6g7.1 and are publicly available as of the date of publication. All other data reported in this paper will be shared by the lead contact upon request.
    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.\cite{georgios_rizos_2023_10253149}
    \item Any additional information required to reanalyze the data reported in this paper is available from the lead contact upon request.    
\end{itemize}

\section*{ACKNOWLEDGMENTS}

%%%  Use this section to acknowledge contributions 
%%%  from non-authors and list funding sources, 
%%%  including grant numbers.

This work was funded by [FUNDER] via grant [GRANT NO.]. The authors thank all members of the lab for their support.

\section*{AUTHOR CONTRIBUTIONS}

%%%  This component is required for most research papers. 
%%%  Mention each individual author with a statement 
%%%  outlining the contribution of each author to the work.

Conceptualization, S.C.P. and S.Y.W.; methodology, A.B., S.C.P., and S.Y.W.; investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.; writing-–original draft, S.C.P. and S.Y.W.; writing-–review \& editing, S.C.P. and S.Y.W.; funding acquisition, S.C.P. and S.Y.W.; resources, M.E.V and C.K.B.; supervision, A.B., N.L.W., and A.A.D.

\section*{DECLARATION OF INTERESTS}

%%%  This component is required for all articles, even 
%%%  if the authors have no competing interests; if 
%%%  this is the case, insert "The authors declare no 
%%%  competing interests." Please refer to the 
%%%  declaration of interests policy: 
%%%  https://www.cell.com/declaration-of-interests

S.Y.W. is an employee and shareholder of COMPANY. M.E.V. is a founder of COMPANY and a member of its scientific advisory board.

\section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}

%%%  If generative AI and AI-assisted technologies 
%%%  were used in the writing process, this must 
%%%  be disclosed in the paper. This declaration 
%%%  does not apply to the use of basic tools for 
%%%  checking grammar, spelling, references, etc. 
%%%  If you have nothing to disclose, please do not 
%%%  include this component.

During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON]. After using this tool or service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.

\section*{SUPPLEMENTAL INFORMATION INDEX}

%%%  Supplemental information must be uploaded as 
%%%  separate files. In the main text, please list the 
%%%  files to be included in a brief index. For details, 
%%%  please review the supplemental information guidelines:

%%%  Journals with a general "methods" section: 
%%%  http://www.cell.com/supplemental-information

%%%  Journals with STAR Methods: 
%%%  https://www.cell.com/STAR-supplemental-information

\begin{description}
  \item Figures S1-S5 and their legends in a PDF
  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
  \item Table S2. Another descriptive title for a different Excel file
  \item Data S1. Raw data on x, y, and z
\end{description}

\newpage

\section*{MAIN FIGURE TITLES AND LEGENDS}

%%%  At final submission, figure files MUST be 
%%%  provided separately as high-resolution image 
%%%  files. All of the panels for a figure should 
%%%  be in the same file. Figures should have 
%%%  clear labels/file names (Figure 1, Figure 2, 
%%%  etc.).

%%%  Figure titles and legends should be placed 
%%%  at the end of the main text. You do not 
%%%  need to place the figures, nor their titles 
%%%  and legends, within the main text. While 
%%%  typesetting your article, our team will 
%%%  place each figure in the best location 
%%%  based on the final layout and on your 
%%%  figure citations, e.g., (Figures 1A and 1B).

%%%  Please review the figure guidelines before 
%%%  submitting your final materials: 
%%%  https://www.cell.com/figureguidelines.

\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}

\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}

The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
\newline
(A) Or each panel or group of panels can be described separately, as shown here and below.
\newline
(B) Graph of X, Y, and Z.
\newline
(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
\newline
Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

\newpage

\section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}

%%%  Whenever possible, we encourage you to submit 
%%%  your main-text tables as Microsoft Word documents, 
%%%  using Word's table function. This will ensure 
%%%  the best results during conversion. Tables 
%%%  should be numbered Table 1, Table 2, etc. and 
%%%  should not include subpanels (do not use Table 1A, 
%%%  1B, etc.). Give each table a brief descriptive 
%%%  title. Table legends are optional but encouraged. 
%%%  Footnotes (superscript lowercase letters) should 
%%%  be used where necessary to indicate some feature 
%%%  of the data; please do not use bold, italic, 
%%%  colored text, or shading for this purpose. Use 
%%%  separate cells, not line breaks or spaces, for 
%%%  all discrete data elements. Small embedded 
%%%  graphics with color are OK.

%%%  Like figures, all tables must be cited within 
%%%  the main text, and our typesetting team will 
%%%  place the tables within the typeset paper at 
%%%  the appropriate locations.

\subsection*{Table 1. A table with clear organization of data}

\begin{tabular}{|l | l | l | l|} 
 \hline
 \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
 \hline
 Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
 \hline
 Row B & 7 & 78 & 5,415 \\ [1ex] 
 \hline
 Row C & 545 & 778 & 7507 \\ [1ex] 
 \hline
 Row D & 545 & 18,744 & 7,560 \\ [1ex] 
 \hline
 Row E & 88 & 788 & 6,344 \\ [1ex] 
 \hline
\end{tabular}

\bigskip

The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.

\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.

\textsuperscript{b}More footnotes

\newpage

%%%  REFERENCES: As of 2023, all Cell Press journals 
%%%  use Numbered (AMA) style. We recommend placing 
%%%  your references in the included "references.bib" 
%%%  file.

\bibliography{references}

\bigskip

%%%  In your References, please include only articles 
%%%  that are published (online publication and 
%%%  preprint servers are OK). Unpublished data, 
%%%  submitted and/or accepted manuscripts, abstracts, 
%%%  and personal communications should be cited within 
%%%  the text only ("unpublished data," "data not 
%%%  shown," "Alice Smith, personal communication") 
%%%  and not included in the references list. Personal 
%%%  communication should be documented by a letter 
%%%  of permission. Whenever possible, please make 
%%%  sure your .bib file has the complete author lists 
%%%  for each item (at minimum, the first 11 authors 
%%%  listed).

\newpage

All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods. Other Cell Press journals use a non-structured \textbf{methods} format; for those journals, please \textbf{delete} the contents of this page from the template and refer instead to the \textbf{methods} template on page 3.

If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
\begin{itemize}
\item \textit{Cell} authors should download the guide for \textit{Cell} authors \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf}{available here}
\item All other journal authors should use \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_general-1678470557763.pdf}{this version of the guide}
\end{itemize}

\section*{STAR METHODS}

%%%  The STAR Methods should appear in your main 
%%%  manuscript file after the figure legends, main 
%%%  table(s) and table legend(s).

\subsection*{Key resources table}

\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}

\subsection*{Experimental model and study participant details}

%%%  Please list here under separate headings 
%%%  all the experimental models/study participants 
%%%  (animals, human participants, plants, microbe 
%%%  strains, cell lines, primary cell cultures) 
%%%  used in the study. For each model, provide 
%%%  information related to their species/strain, 
%%%  genotype, age/developmental stage, sex (and 
%%%  gender, ancestry, race, and ethnicity if 
%%%  reported for human studies), maintenance, 
%%%  and care, including institutional permission 
%%%  and oversight information for the studies 
%%%  the experimental animal/human participant 
%%%  study. The influence (or association) of sex, 
%%%  gender, or both on the results of the study 
%%%  must be reported. In cases where it cannot, 
%%%  authors should discuss this as a limitation 
%%%  to their research’s generalizability.

%%%  Please omit this component if your study does 
%%%  not use experimental models typical in the 
%%%  life sciences (e.g., if your study is 
%%%  computational or physical science research).

\subsection*{Method details}

%%%  Please provide precise details of all the 
%%%  procedures in the paper (behavioral task, 
%%%  generation of reagents, biological assays, 
%%%  modeling, etc.) such that it is clear how, 
%%%  when, where, and why procedures were 
%%%  performed. We encourage authors to provide 
%%%  information related to the experimental 
%%%  design as suggested by NIH and ARRIVE 
%%%  guidelines (e.g., information about 
%%%  replicates, randomization, blinding, sample 
%%%  size estimation, and the criteria for 
%%%  inclusion and exclusion of any data or 
%%%  subjects).

\subsection*{Quantification and statistical analysis}

%%%  Please describe here all statistical analysis 
%%%  and software used. We ask authors to indicate 
%%%  in this component where all of the statistical 
%%%  details of experiments can be found (e.g., in 
%%%  the figure legends, figures, results, etc.), 
%%%  including the statistical tests used, exact 
%%%  value of n, what n represents (e.g., number 
%%%  of animals, number of cells, etc.), definition 
%%%  of center, and dispersion and precision 
%%%  measures (e.g., mean, median, SD, SEM, confidence 
%%%  intervals). Also, please summarize in this 
%%%  component how significance was defined, the 
%%%  statistical methods used to determine strategies 
%%%  for randomization and/or stratification, sample 
%%%  size estimation, and inclusion and exclusion 
%%%  of any data or subjects, as well as any methods 
%%%  used to determine whether the data met 
%%%  assumptions of the statistical approach.

\subsection*{Additional resources}

%%%  Please provide links to websites that provide 
%%%  further information relevant to the study 
%%%  (e.g., protocol download, troubleshooting 
%%%  forum, etc.). Clinical trial registry numbers 
%%%  and links should also be placed here. Please 
%%%  briefly describe the resource and its 
%%%  relevance for the paper. Please report this 
%%%  information as: “Description: URL.”

Cell.com homepage: https://www.cell.com
\newline Templates for Cell Press authors: https://www.cell.com/article-templates

%%%  ADDITIONAL MANUSCRIPT COMPONENTS:

%%%  Depending on the journal and the article 
%%%  type, you may be asked to upload the 
%%%  following as separate files: graphical 
%%%  abstract, highlights, eTOC blurb (In 
%%%  Brief), and/or other article components 
%%%  such as a "bigger picture" statement. 
%%%  These items are typically not required for 
%%%  initial submissions. Please refer to the 
%%%  journal's website, your acceptance 
%%%  letter, and/or the Final Files 
%%%  Requirements checklist to see if any of
%%%  these items are required (at any stage).

\end{document}

### 2025-01-24 manuscript.tex
result section 1 done.

@@
 
-\title{Title is a maximum of 145 characters, including spaces, and avoids the use of jargon and uncommon abbreviations}
+\title{Deciphering enhancer rules underlying early human neuronal and neural crest development.}
 \date{}
@@
 
-\author[1,5,\orcidlink{0000-0000-0000-0000}]{First Last}
-\author[1,2,5]{First Middle Last}
-\author[2,3]{First M.M. Last}
-\author[3,*]{First Last, Jr.}
-\author[3,4,6,7,8,**]{F. Middle Last}
+\author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
+\author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}
 
@@
 
-\affil[1]{University A, London SW7 2AZ, UK}
-\affil[2]{Department B, University B, Toronto, ON M5S 3H6, Canada}
-\affil[3]{Division C, Department C, University C, Cambridge, MA, USA}
-
-\affil[4]{Present address: 123 Sesame Street, New York, NY, USA}
-\affil[5]{These authors contributed equally}
-\affil[6]{Senior author}
-\affil[7]{Twitter: @CorrespondingAuthor}
-\affil[8]{Lead contact}
+\affil[1]{VIB Center of AI \& computational Biology, Leuven, Belgium}
+\affil[2]{VIB-KU Leuven Center for Brain and Disease Research, Leuven, Belgium}
+\affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
 
@@
 
-\affil[*]{Correspondence: first.last.jr@university.edu}
-\affil[**]{Correspondence: f.middle.last@university.edu}
+\affil[*]{Correspondence: stein.aerts@kuleuven.be}
 
@@
 
-The summary (abstract) should consist of a single paragraph of 150 words or fewer.
+%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
 
@@
 
-cell, press, LaTeX, template
-
 \section*{INTRODUCTION}
 
-No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be defined when first used in the text.
+%%% No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be
+%%% defined when first used in the text.
 
@@
 
-The results may be divided with subheadings.
-In our view, good subheadings convey information about the findings, so we encourage you to be specific.
-For example, "Factor X requires factor Y to function in process Z" is better than "Analysis of factors X and Y by approach Q." We recommend that you use similar language in your figure titles for clarity and structural harmony.
-Please do not use numbered subheadings.
-For some formats, ``results and discussion" may be combined under a single heading.
+\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
+We made use of human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
+end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem
+cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. These were
+incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3
+until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
+neural crest states (REF). In order to characterize the organoid-derived cell types to a greater level of detail at both
+transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
+scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-110; \textbf{Fig. 1a}),
+resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
+states were obtained from all three cell lines (\textbf{FIGURE SUPPLEMENT}) demonstrating the robustness of the
+organoid system.\par
+To evaluate the biological relevance of the cell types and states obtained from the organoid, we performed 10x single
+cell multiome on the head of a single four weeks post conception human embryo, resulting in 25,785 high quality cells (
+\textbf{Fig. 1c}). We annotated both the organoid and embryo cells based on the expression of marker genes
+(\textbf{TABLE SUPPLEMENT}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
+neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
+Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and in the
+embryo only we could identify an endothelial, perivascular macrophage (PVM) / microglia (MIC) and otic placode
+population (\textbf{Fig. 1c}).\par
+Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). Both
+organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning as
+is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal ARX, NKX2-2, PAX6, IRX3, ZIC1 and
+GDF7 (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
+A cluster that still expresses the neuronal progenitor marker SOX2, cells that express the early neuronal induction
+marker ASCL1 and distinct clusters marked by the expression of GAD1, UNCX and GATA2 (\textbf{Fig. 1e}). Finally,
+focusing on the neural crest cells, in both the organoid and the embryo we obtained a population of pre-migratory neural
+crest (GRHL2 expression and absence of ZEB2 expression), migratory neural crest (SOX10 and ZEB2 expression), facial
+mesenchyme cells (TWIST1 expression) and early differentiation peripheral neurons (NEUROD1 expression). In the embryo,
+but not in the organoid, we observed an additional population of cells marked by the expression of SOX10 and PAX2
+reflecting the otic placode (\textbf{Fig. 1f}).\par
+Given that we performed single-cell multiome on the entire head of the four weeks post conception embryo we also
+assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
+(SIX3) or anterior spinal cord (HOXB3) identity, we observed a gradient of anterior-to-posterior identities in the
+embryo cells marked by the expression of anterior-posterior marker TF (REF): from anterior to posterior SIX3, EMX2,
+FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par
+Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
+both organoid and embryo cells for the expression of marker genes of the G1, G2M and S phase of the cell cycle and
+quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
+clear that both early differentiation peripheral neurons and to a greater extend early differentiating neurons of the
+central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
+Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo-derived cell types
+and states on a chromatin accessibility level we identified sets of co-accessible regions in the organoid model using
+topic modeling. We focused on topics (i.e., sets of co-accessible regions) that were enriched for marker TF motifs and
+calculated the average expression of marker genes for each cell group where these regions were co-accessible, for both
+organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
+the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of \textbf{X.X}
+across identical states (\textbf{FIGURE SUPPLEMENT}).
 
-\subsection*{Figures}
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
 
-For initial submissions, figures and their titles and legends can be run-in with the text.
-For final submissions, the image files must be uploaded as separate files, one complete file per figure.
-The figure titles and legends should be placed toward the end of the main text, after the declaration of interests.
-Our typesetters will place the figures and their titles and legends appropriately based on your in-text citations.
+\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
 
-\subsection*{Citations}
+\textbf{a}, Schematic. \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
+(\textbf{c}) cells colored by cell type identify. \textbf{d-f}, UMAP based on gene expression profiles of organoid (top)
+and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons
+(\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
+\textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
+axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
+color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
+cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
+type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
+$log_{2}(CPM)$
+(dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
+cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
+number of regions enriched for the motif (dot size) per topic (x-axis) and motif (y-axis; bottom). PVM/MIC: perivascular
+microphage/microglia.
 
-To cite/link to display items (figures and tables) and/or sections of the manuscript, simply write, for example, ``(Figure 1)" or ``see discussion." Our team will do the rest.
-Please note that we do not number sections or subsections.
+\subsection*{
+  DeepNeuralTube: a sequence-to-function model revealing the enhancer code underlying early neuronal and neural crest 
+  development
+}
 
-To cite references, you may use the cite command, e.g., ``Recent articles in \textit{Matter} and \textit{Cell} \cite{wang2020conformational,king2002,cates1984statics}  have shown ..." or  ``Many interesting discoveries have been reported, \cite{radzihovsky1995new,kapoor2022leakage,hantke2018my,alcock2006} which ..."
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
 
-\subsection*{Equations}
-
-Simple formulae should appear in line with the text whenever possible.
-You can write inline math by enclosing it between \verb|\(| and \verb|\)|, as in this example: \(x^2 + y^2 = z^2\).
-You can also enclose it between dollar signs (\texttt{\$}), as in this example: $E=mc^2$.
-
-Larger, more complex formulae may appear on a new line, either by enclosing them between \verb|\[| and \verb|\]| or by using the \verb|displaymath| environment:
-
-\[ x^n + y^n = z^n \]
-
-\begin{displaymath}
-\sqrt{x^2+1}
-\end{displaymath}
-
-If any equations or formulae need to be referred to or cited again later in the text, use the \verb|equation| environment to number them. Later, you can cite these as "Equation 1," "Equation 2," etc.
-
-\begin{equation}
-f(x) = \sum_{i=0}^{n} \frac{a_i}{1+x}
-\end{equation}
+\subsection*{Figure 2. }
 
@@
 
-The discussion should explain the significance of the results and place them into a broader context. Subheadings are permitted.
+%%% The discussion should explain the significance of the results and place them into a broader context.
+%%% Subheadings are permitted.
 
@@
 
-A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some journals and some article formats.
+%%% A "limitations" or "limitations of the study" subsection in the discussion is encouraged and may be required for some
+%%% journals and some article formats.
 
@@
 %%%  https://www.cell.com/figureguidelines.
-
-\noindent\includegraphics[width=0.85\linewidth]{Figure1.jpg}
-
-\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
-
-The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
-\newline
-(A) Or each panel or group of panels can be described separately, as shown here and below.
-\newline
-(B) Graph of X, Y, and Z.
-\newline
-(C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
-\newline
-Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}.
-Example figure created by Cassie Comeau, Cell Press.
 

### 2025-01-27 manuscript.tex
section 2, sans figure legend.

@@
 (SIX3) or anterior spinal cord (HOXB3) identity, we observed a gradient of anterior-to-posterior identities in the
-embryo cells marked by the expression of anterior-posterior marker TF (REF): from anterior to posterior SIX3, EMX2,
-FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par
+embryo (REF): from anterior to posterior expression of SIX3, EMX2, FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par
 Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
@@
 
-\subsection*{Figure 1. A brief title that describes the entire figure without citing specific panels}
+\subsection*{Figure 1. Human neural tube organoids recapitulate early neuronal and neural crest development}
 
@@
 \subsection*{
-  DeepNeuralTube: a sequence-to-function model revealing the enhancer code underlying early neuronal and neural crest 
+  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
   development
 }
+
+To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube, neural crest
+and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
+"$DeepNeuralTube_{organoid}$" and "$DeepNeuralTube_{embryo}$". The models use DNA sequence, of genomic regions, as input and
+classify the regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class
+fashion; \textbf(Fig. 2a)). The classes of the models are defined based on topic modeling. To get a fine-grained
+definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest
+(incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both the organoid cells (to
+train $DeepNeuralTube_{organoid}$) and embryo cells (to train $DeepNeuralTube_{embryo}$). To validate the ability of the models to
+predict cell type-specific enhancer \textit{activity}, we tested their performance on enhancers from the VISTA enhancer
+browser (\textbf{REF}). Relevant for this study, this database contains enhancers active in the neural tube (n=349) and
+facial mesenchyme (n=143). Neural tube enhancers could be classified with an area under the receiver-operator curve
+(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ (\textbf{Fig. 2b}; and
+area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{FIGURE SUPPLEMENT})), facial mesenchyme
+enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$
+(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{FIGURE SUPPLEMENT})). Of note, a fraction of neural tube
+enhancers (x.x \%) are predicted to be specific to pre-migratory neural crest by both models.\par
+Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
+TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic and
+calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
+TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs used by the models. For both known motifs and \textit{de novo}
+motifs we calculated pairwise similarities using TOMTOM (\textbf{REF}) followed by clustering, resulting in 56 motif
+clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
+(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{SUPPLEMENTAL TABLE}) and cluster
+26 which is embryo specific (CG-rich motif; \textbf{SUPPLEMENTAL TABLE}). Most clusters (\textbf{Fig. 2f-g} and
+\textbf{SUPPLEMENTAL FIGURE}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
+Thus, $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ use known-motifs as features to classify regions based
+on cell type-specific chromatin accessibility.\par
+As an illustrative example, we investigated an enhancer located in one of the introns of GLI3 (hs111.0; \textbf{SUPPLEMENTAL
+FIGURE}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both
+$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ explain this region similarly (Pearson correlation of x) and
+highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). A version of this enhancer where 5\% of the
+nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both
+models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par
+Finally, to also investigate the performance of the models on migratory neural crest, a cell type for which there are no
+validated enhancers in the Vista enhancer browser, we selected a genomic region that is specifically accessible in the
+migratory neural crest cell type of the organoid (\textbf{SUPPLEMENTAL FIGURE}). Both models predict this region
+to be specific to the migratory neural crest class (\textbf{SUPPLEMENTAL FIGURE}) and nucleotide importance scores
+reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). We tested the enhancer activity of this region
+using a reporter assay in chicken embryos. Indeed the enhancer was specifically active in migratory neural crest 24
+hours after injection in an HH4 embryo (\textbf{Fig. 2i}).
 

### 2025-03-18 manuscript.tex
Section on figure 3 done!

@@
 
-\subsection*{Figure 2. }
+\subsection*{Figure 2. DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer
+code.}
+\textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
+$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
+using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
+colored by: cluster identity (\textbf{c}), model of origin (\textbf{d}; only showing \textit{de novo} motifs in this
+case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
+\textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
+\textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
+for $DeepNeuralTube_{organoid}$ (top two panels) and $DeepNeuralTube_{embryo}$ (bottom two panels) and enhancer-reporter
+assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
+$DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
+enhancer (chr ...; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
+constitutive enhancer that serves as electroporation control.
+
+\subsection*{
+  Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors
+}
+Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
+identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
+54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
+two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
+\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
+SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
+belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
+enhancers using DeepNeuralTube reveals two FOX (one partial), one SOX and two RFX (one partial) binding sites in SFPE1
+and two FOX and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par
+
+To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
+contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$.
+We then used TF-MoDISco\cite{shrikumar2020tfmodisco} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG).
+Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig.
+3e}).
+Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
+underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
+models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
+and RFX instances is important (\textbf{Fig.
+3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA2}, \textit{TEAD1} and \textit{RFX4} in both the organoid and embryo cells (\textbf{Fig.
+3g}).
+In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-2} expression (\textbf{(Fig.
+3e- g)}); PAX instances, correlating with \textit{PAX6} expression (\textbf{Fig.
+3e-g}) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
+3e-g}).
+For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2} expression.
+Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig.
+3e}).
+Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
+3h}).
+FOXA2, TEAD1 and RFX4 instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig.
+3h}).\par
+
+It has been reported that FOXA2 is a pioneer factor. Indeed, when centering on FOXA2 instances in both the organoid and
+embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
+asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
+TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
+peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOXA2 instances. Those floorplate peaks
+that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
+did not we classified as floorplate specific. Next, we quantified both the number of FOXA2 instances per region and the
+affinity of each individual instance (making use of the FOXA2 motif score calculated using
+FIMO\cite{grant2011fimo,schreiber2025tangermeme}) hypothesizing that the overall affinity of regions that are generally
+bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
+the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
+have a single FOXA2 instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
+Furthermore, in case there is but a single FOXA2 instance in a general FOXA2 target the affinity of this instance tends
+to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
+have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
+\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).
+
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
+
+\subsection*{
+  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
+}
+\textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics.
+\textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells).
+\textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38.
+SFPE2 and SFPE1 is highlighted.
+\textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom).
+\textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic.
+Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
+Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+\textbf{f}, Heatmap showing the maximum normalized contribution score for each region (y-axis) and pattern (x-axis).
+\textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics.
+\textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+\textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance.
+Signal is centered on the FOXA2 instance start position.
+\textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top).
+\textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grant2011fimo,schreiber2025tangermeme} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
+\textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
+
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
+
+\subsection*{Figure 4. }
+
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
+
+\subsection*{Figure 5. }
 

### 2025-03-19 manuscript.tex
Section figure 4 done!

@@
 the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of \textbf{X.X}
-across identical states (\textbf{FIGURE SUPPLEMENT}).
+across identical states (\textbf{FIGURE SUPPLEMENT}).\par
 
@@
 using a reporter assay in chicken embryos. Indeed the enhancer was specifically active in migratory neural crest 24
-hours after injection in an HH4 embryo (\textbf{Fig. 2i}).
+hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par
 
@@
 have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
-\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).
+\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
 
@@
 
+\subsection*{
+  Anterior-posterior patterning in neuronal progenitors is regulated by different homeodomain transcription factors that
+  each recognize a distinct binding site.
+}
+
+Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
+organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
+identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
+Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
+from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as in those in
+which DV patterning is encoded? For this purpose, we calculated the pairwise correlation between region-topic
+probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
+as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
+that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-2}; \textbf{Fig. 4b})\par
+
+$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
+together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
+explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
+homeodomain TF binding motifs, one for each domain (\textbf{Fig.
+4e}).
+From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA} corresponding with the expression of respectively \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1} and \textit{HOXB3} [ADD GENE EXPRESSION HEATMAP!].
+Finally, SOX2 instances are present in regions of all AP domains (\textbf{Fig.
+4e}).\par
+
+Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
+accessibility of those domains.
+Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}.
+We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains.
+To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig.
+4f}).
+Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig.
+4f}).
+Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiran2024synthetic, kempynck2025crested} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig.
+4g}).
+[SWITCH PANELS].
+Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig.
+4g}).
+Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class.
+This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig.
+4h}).
+In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated.
+This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig.
+4h}).\par
+
+From these results it seems that a relatively simplistic code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains.
+Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig.
+4i}).
+Surprisingly, this very simple classifier reached high performance with area under the ROC of X and area under the precision recall curve of X (\textbf{Fig.
+4i}).\par
+
 \noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
 
-\subsection*{Figure 4. }
+\subsection*{
+Figure 4. A simple enhancer code underlies anterior-posterior patterning in neuronal progenitors.
+}
+\textbf{a}, UMAP of embryo neuronal progenitors colored by cell-topic probabilities in RGB color scales for
+anterior-posterior topics.
+\textbf{b}, Heatmap showing pairwise pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics.
+\textbf{c}, Heatmap showing region-topic probabilities for the top 3,000 regions per topic (y-axis) and each anterior-posterior topic (x-axis).
+\textbf{d}, Heatmap showing prediction scores of $DeepNeuralTube_{embryo}$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis).\textbf{e}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topics (x-axis).\textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31.
+\textbf{g}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 61 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 31 (bottom); n = 200.
+\textbf{h}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 31 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 61 (bottom); n = 200.
+.\textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
 

### 2025-03-19 manuscript.tex
Section figure 5 done!

@@
 Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
-\textbf{f}, Heatmap showing the maximum normalized contribution score for each region (y-axis) and pattern (x-axis).
+\textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics.
 \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics.
@@
 
+\subsection*{
+  Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
+}
+
+Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
+crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
+$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
+topic and generated a code table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
+$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
+these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par
+
+Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig.
+5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig.
+5d}).
+Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig.
+5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig.
+3e}) and neurons (\textbf{Fig.
+6b}).
+The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
+5d}).
+This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiran2024synthetic}.\par
+
+The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94).
+Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig.
+5b-c}).
+We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances.
+The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig.
+5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}.
+The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothstein2020heterodmerization}.
+After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig.
+5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig.
+5d}).\par
+
+Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
+pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
+importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
+expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
+shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
+facial mesenchyme\cite{kim2024dnaguided}.\par
+
+Next, we investigated which patterns co-occur in the same genomic regions. For this purpose, we calculated the Jaccard
+index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
+method we found three sets of regions that are predicted to be contain instances of different TFs. One set where
+instances of TEAD4, GRHL1/2 and ZEB1/2 co-occur, one where instances of Nuclear receptor, TFAP2A/B/C and ZIC1-5
+instances co-occur and one set where instances of FOXC2 and TWIST1/ALX4 co-occur.\par
+
+Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
+that are associated with facial variation\cite{white2021insights}. Given that regions surrounding these SNPs were
+enriched for enhancer activity in cranial neural crest cells\cite{white2021insights} we asked if, using
+$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
+suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score).
+At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models.
+Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig.
+5f}).
+Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG).
+As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}.
+The G>A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig.
+5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state.
+This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp upsteam of a partial homeobox site potentially creating a new binding site for TWIST1-ALX4 dimers (\textbf{Fig.
+5h}).\par
+
 \noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
 
-\subsection*{Figure 5. }
+\subsection*{
+  Figure 5. Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
+}
+\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics.
+\textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic.
+Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
+Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+\textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics.
+\textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neural crest related topics.
+\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+\textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{white2021insights} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome.
+Lead SNPs are shown with a bigger dot size.
+\textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red.
+\textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube_{organoid}$ of the references (top) and alternative (bottom) allele of SNP rs1555067.
+The location of the SNP is indicated with a dotted line.
+
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_6.png}
+
+\subsection*{Figure 6. }
+
+\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_7.png}
+
+\subsection*{Figure 7. }
 

### 2025-03-19 manuscript.tex
Section figure 6 done!

@@
 
+\subsection*{
+  Deciphering the enhancer code underlying early neuronal differentiation.
+}
+
+Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
+acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube_{organoid}$ and
+$DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each topic and generated a code table based on the
+identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
+of these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 6b}).\par
+
+For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
+instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
+These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD4/1},
+\textit{FOXA2} and \textit{RFX4} (\textbf{Fig.
+6d}).
+Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig.
+6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig.
+6d}).
+Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig.
+6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1}.
+For subsequent topics SOX2 instances are not important anymore (\textbf{Fig.
+6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24\textbf{Fig.
+6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig.
+6d}) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
+6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}.
+Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2}.
+Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig.
+6b-c}) correlating with the expression of \textit{GATA2}.\par
+
+Next, we analyzed the co-occurence of the identified patterns by calculating the Jaccard index of regions based on the presence of instances of each pattern.
+We find that instances of neuronal progenitor patterns tend to co-occur the most (\textbf{Fig.
+6e}), in line with our results in figure 3.\par
+
 \noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_6.png}
 
-\subsection*{Figure 6. }
+\subsection*{
+  Figure 6. Deciphering the enhancer code of early neuronal differentiation.
+}
+
+\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics.
+\textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic.
+Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
+Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+\textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neuron related topics.
+\textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neuon related topics.
+\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
 

### 2025-03-19 manuscript.tex
Section figure 7 done!

@@
 
-\subsection*{
+\subsection* {
   Figure 6. Deciphering the enhancer code of early neuronal differentiation.
@@
 
+\subsection* {
+  Four to five distinct transcription factor binding sites in a region of around 240 base pairs is sufficient to encode
+  all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
+}
+
+Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
+tube organoids and four weeks post conception embryo head with almost one-to-one correspondence based on TFBS importances and TF expression.
+Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo.
+To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as to belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig.
+3e}, \textbf{Fig.
+4e}, \textbf{Fig.
+5b} and \textbf{Fig.
+6b}).
+The overall performance of these models was acceptably high (\textbf{Fig.
+7a}) with an average area under the ROC of X and area under the precision-recall curve of X.
+Notably, all models for neural crest related states performed best (\textbf{Fig.
+7a}).
+Next, we selected patterns that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one patterns for fifteen to sixteen states plus the additional four AP states (\textbf{Fig.
+7a}).
+Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiran2024synthetic, kempynck2025crested}.
+We used at least three TFBS in the design process for each state.
+In case less than three distinct patterns had a positive coefficient for a state, we implanted multiple copies of each pattern to come to a total of at least four TFBS.
+The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig.
+7a}) - and were cell type specific (\textbf{Fig.
+7b}) - with the prediction score for neural crest related states being again the highest.
+Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig.
+7a}) while other are more specific to certain states.
+\par
+
+To conclude, enhancers specific to DV-patterning, neural crest states and early neuronal differentiation tend to be encoded in
+the human genome using around four to five distinct TFBS (\textbf{Fig.
+7a}) in regions of on average 240-243bp +/- 67-77bp in size (\textbf{Fig.
+7c}).
+AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig.
+7a}).
+
 \noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_7.png}
 
-\subsection*{Figure 7. }
+\subsection*{
+  Figure 7. Neural tube cell states are encoded using four to five distinct transcription factor binding
+  sites
+}.
+\textbf{a}, Area under the receiver-operator curve (outlined dots) and precision-recall curve of logistic regression
+models classifying genomic regions to each topic using motif scores as features (top) and patterns with positive
+coefficient for each logistic regression model along with the number of each patterns used for enhancer design
+represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
+each topic (bottom). \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
+for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
+location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
+for each pattern calculated using FIMO\cite{grant2011fimo, schreiber2025tangermeme} for organoid (top left) and embryo
+(top right) and histograms showing the number of motif instances per 25 bp bins with confidence intervals indicated in
+gray scales (black is +/- 1 standard deviation, gray is +/- 2 standard deviation and white is more than 3 standard
+deviations) for organoid (bottom left) and embryo (bottom right).
 

### 2025-03-20 manuscript.tex
Introduction done

@@
 
-%%% No subheadings, please. Use of abbreviations should be kept to a minimum, and nonstandard abbreviations should be
-%%% defined when first used in the text.
+%%% INTRO TO FIELD
+
+Genomes of complex multi-cellular organisms encode all cell types/states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} of that organism in specific non-coding elements named
+enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
+transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the precise arrangement, copy
+number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}.
+Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}.
+For example, during immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, during human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}.
+Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par
+
+Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
+genome. Especially organoids derived from pluripotent stem cells are interesting in this context, given that they
+demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
+takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore, give insights into how genomes
+interpret both cell intrinsic and extrinsic signals. For example, epigentic
+modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
+perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
+signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
+model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
+example, early human development.\par
+
+%%% What is missing and bridge
+
+Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts, at a transcriptomic level.
+For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
+However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.\par
+
+During vertebrate development both the neural tube and neural crest originate from the neural ectoderm. The neural tube
+consist of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo later giving rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}.
+Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}.
+The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied.
+Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of each individual cell types are still limited.\par
+
+In this study, we will determine the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.
 

### 2025-03-21 manuscript.tex
Part of discussion.

@@
 
+We built an enhancer-code atlas of early human neuronal and neural crest development. Providing enhancer-codes of
+neuronal progenitor DV- and AP-patterning, early neuronal fate acquisition and neural crest induction, migration and
+facial mesenchyme specification.\par
+
+In order to create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes.
+Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo.
+The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance.
+Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in human.\par
+
+% enhancers
+
+% human tfs paper: https://www.sciencedirect.com/science/article/pii/S0092867418301065?via%3Dihub
+
+% https://www.cell.com/fulltext/s0092-8674(12)01496-1: also about dimerization and that motif "flanks" are important for
+% different TFs
+% also: https://www.nature.com/articles/s41588-022-01048-5 about flank
+
+%% Enhancers are clusters of TFBS
+%%%% https://www.cell.com/cell/fulltext/S0092-8674(16)31251-X
+%%%% https://www.cell.com/molecular-cell/fulltext/S1097-2765(22)01215-1
+%%%% https://www.sciencedirect.com/science/article/pii/S0959440X21000993
+
+%% Specificity through dimerization
+%%%% https://www.cell.com/cell/fulltext/S0092-8674(23)01438-1
+%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC6961570/
+
+% single TF
+% https://genome.cshlp.org/content/26/7/882.short
+
+%% pioneer
+% https://www.nature.com/articles/s41588-018-0140-x
+% https://elifesciences.org/articles/73358
+% https://www.cell.com/cell/fulltext/S0092-8674(15)00304-9
+% https://www.biorxiv.org/content/10.1101/2024.05.24.595749v3.abstract
+% https://www.nature.com/articles/s41576-023-00648-z
+% https://www.nature.com/articles/s41586-020-2195-y
+
+%% Low affinity/suboptimization:
+%%%% https://www.nature.com/articles/s41586-023-06922-8
+%%%% https://www.science.org/doi/10.1126/science.aac6948
+%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC11160683/
+%%%% https://www.sciencedirect.com/science/article/abs/pii/S0070215315001842
+
+%% Repression through chromatin closing
+% carmen paper & ibrahim paper
+
+%% what about repression through chromatin opening
+
+%% clusters of low affinity HOX binding sites
+% https://www.cell.com/cell/fulltext/S0092-8674(14)01518-9
+% https://elifesciences.org/articles/28975
+
+% https://www.pnas.org/doi/epub/10.1073/pnas.2411887121
+% https://www.nature.com/articles/s41594-024-01235-4
+
+% another paper of face GWAS: https://www.nature.com/articles/s41588-018-0057-4
+
+% might be interesting
+% https://www.nature.com/articles/s41594-024-01449-6
+% https://www.nature.com/articles/s41586-018-0549-5
+
+% white box models
+% DeepNeuralTube
+% AP patterning (TP53)
+% interpretation of human variation
+
 \subsection*{Limitations of the study}

### 2025-03-24 manuscript.tex
More on discussion.

@@
 \subsection* {
-  Four to five distinct transcription factor binding sites in a region of around 240 base pairs is sufficient to encode
+  Four to five distinct transcription factor binding sites in a region of around 140 base pairs is sufficient to encode
   all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
@@
 
-% enhancers
-
-% human tfs paper: https://www.sciencedirect.com/science/article/pii/S0092867418301065?via%3Dihub
-
-% https://www.cell.com/fulltext/s0092-8674(12)01496-1: also about dimerization and that motif "flanks" are important for
-% different TFs
-% also: https://www.nature.com/articles/s41588-022-01048-5 about flank
-
-%% Enhancers are clusters of TFBS
-%%%% https://www.cell.com/cell/fulltext/S0092-8674(16)31251-X
-%%%% https://www.cell.com/molecular-cell/fulltext/S1097-2765(22)01215-1
-%%%% https://www.sciencedirect.com/science/article/pii/S0959440X21000993
-
-%% Specificity through dimerization
-%%%% https://www.cell.com/cell/fulltext/S0092-8674(23)01438-1
-%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC6961570/
+The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
+compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
+are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
+DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
+TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
+within one nucleosome depleted regions (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
+the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
+model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
+model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
+of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
+TWIST1 and ALX4 heterodimer in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest.
+For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances.
+Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG).
+Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}.
+This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}.
+Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}.
+Similarly, we observed that FOXA2 binding sites that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity.
+The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.
+
+Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancer
+consist out of clusters of 4-5 different TFBS.
+Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain.
+Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}.
+It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}.
+It has been hypothesized that specificity is mainly achieved through cooperative binding with co-factors\cite{mannChapter3Hox2009}.
+Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs.
+In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for TFX and TFY that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites.
+Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity.
+In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that can be important for cell type specific binding.
 
@@
 
-%% pioneer
-% https://www.nature.com/articles/s41588-018-0140-x
-% https://elifesciences.org/articles/73358
-% https://www.cell.com/cell/fulltext/S0092-8674(15)00304-9
-% https://www.biorxiv.org/content/10.1101/2024.05.24.595749v3.abstract
-% https://www.nature.com/articles/s41576-023-00648-z
-% https://www.nature.com/articles/s41586-020-2195-y
-
-%% Low affinity/suboptimization:
-%%%% https://www.nature.com/articles/s41586-023-06922-8
-%%%% https://www.science.org/doi/10.1126/science.aac6948
-%%%% https://pmc.ncbi.nlm.nih.gov/articles/PMC11160683/
-%%%% https://www.sciencedirect.com/science/article/abs/pii/S0070215315001842
+%% position of the binding sites vis a vi nucleosomes
+% https://www.pnas.org/doi/abs/10.1073/pnas.1804663115
 
@@
 
-% https://www.pnas.org/doi/epub/10.1073/pnas.2411887121
-% https://www.nature.com/articles/s41594-024-01235-4
-
 % another paper of face GWAS: https://www.nature.com/articles/s41588-018-0057-4
@@
 % https://www.nature.com/articles/s41586-018-0549-5
+
+% co-factors (limitation)
+% "new" studies
+% https://www.biorxiv.org/content/10.1101/2022.08.26.505496v1.abstract
+% https://www.cell.com/cell/fulltext/S0092-8674(20)31541-5?uuid=uuid%3A51e14242-dce3-432c-8521-14d9a94b0381
+% https://www.cell.com/molecular-cell/fulltext/S1097-2765(21)01070-4?uuid=uuid%3Afbec3fc1-2707-4902-bda8-f196da451272
+% https://www.science.org/doi/full/10.1126/science.adf6149
+% focus on single enhancers and not entire locus
+%
+% from lambert et al
+%%TFs have traditionally been classified as either ‘‘activators’’ or ‘‘repressors’’; however, this notion has been repeatedly questioned.
+Many TFs can recruit multiple cofactors that have opposite effects (Frietze and Farnham, 2011; Rosenfeld et al., 2006;
+% Schmitges et al., 2016), dependent on the local sequence
+%context and availability of cofactors (Meijsing et al., 2009;
+%Wong and Struhl, 2011).
 

### 2025-03-25 manuscript.tex
Discussion finished.

@@
 Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}.
-Similarly, we observed that FOXA2 binding sites that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity.
-The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.
+Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity.
+The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par
 
@@
 It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}.
-It has been hypothesized that specificity is mainly achieved through cooperative binding with co-factors\cite{mannChapter3Hox2009}.
+It has been hypothesized that specificity is mainly achieved through cooperative binding with other TFs\cite{mannChapter3Hox2009}.
 Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs.
@@
 Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity.
-In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that can be important for cell type specific binding.
+In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that could be important for cell type specific binding.\par
+
+In conclusion, we provide an enhancer-code atlas of early human neuronal and neural crest development. We confirm the
+paradigm that enhancers consist out of clusters of TFBS with soft syntactic rules but also provide a counter example to
+this paradigm.
+Furthermore, we validate the biological relevance of neural tube organoids.
+Finally, we provide $DeepNeuralTube$, two sequence-to-chromatin accessibility models, freely to the research community, which can be used to predict, interpret and design enhancers underlying early human neuronal and neural crest development.
 
@@
 
-%% clusters of low affinity HOX binding sites
-% https://www.cell.com/cell/fulltext/S0092-8674(14)01518-9
-% https://elifesciences.org/articles/28975
-
 % another paper of face GWAS: https://www.nature.com/articles/s41588-018-0057-4
-
-% might be interesting
-% https://www.nature.com/articles/s41594-024-01449-6
-% https://www.nature.com/articles/s41586-018-0549-5
-
-% co-factors (limitation)
-% "new" studies
-% https://www.biorxiv.org/content/10.1101/2022.08.26.505496v1.abstract
-% https://www.cell.com/cell/fulltext/S0092-8674(20)31541-5?uuid=uuid%3A51e14242-dce3-432c-8521-14d9a94b0381
-% https://www.cell.com/molecular-cell/fulltext/S1097-2765(21)01070-4?uuid=uuid%3Afbec3fc1-2707-4902-bda8-f196da451272
-% https://www.science.org/doi/full/10.1126/science.adf6149
-% focus on single enhancers and not entire locus
-%
-% from lambert et al
-%%TFs have traditionally been classified as either ‘‘activators’’ or ‘‘repressors’’; however, this notion has been repeatedly questioned.
-Many TFs can recruit multiple cofactors that have opposite effects (Frietze and Farnham, 2011; Rosenfeld et al., 2006;
-% Schmitges et al., 2016), dependent on the local sequence
-%context and availability of cofactors (Meijsing et al., 2009;
-%Wong and Struhl, 2011).
 
@@
 %%% journals and some article formats.
+
+In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. The effect a
+TF has on target gene expression depends on its effector domain and interactions with
+co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
+the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. Large scales studies, in a variety of cell
+states, are now being performed to map TFs to co-factors and downstream effects on gene
+expression\cite{tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
+sotoCompendiumHumanTranscription2022, delrossoLargescaleMappingMutagenesis2023,
+mukundHighthroughputFunctionalCharacterization2023, bellComparativeCofactorScreens2024,
+tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. These studies may especially be
+important to infer repressive roles of TFs on gene expression, an effect that is currently difficult to study with state
+of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
+dewinterModellingDesignTranscriptional2025}. In particular for the neural tube, it is known that repressive interactions
+in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. In this study we could only predict a strong
+repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
+systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024} (SUPPLEMENT FIG).\par
+
+Furthermore, the expression of a gene is in general regulated by multiple enhancers in the gene's locus\cite{perryShadowEnhancersFoster2010,
+martinez-araSystematicAnalysisIntrinsic2022, bergmanCompatibilityRulesHuman2022, zuinNonlinearControlTranscription2022,
+gschwindEncyclopediaEnhancergeneRegulatory2023, martinez-araLargescaleAnalysisIntegration2024,
+loubiereDevelopmentalHousekeepingTranscriptional2024, thomasEnhancerCooperativityCan2025}. In this study we focused on
+the enhancer code of individual enhancers but not how the signal of multiple enhancers is integrated to control target
+gene expression. Currently efforts are ongoing to build large scale models for modeling gene expression making use of the
+code of all enhancers in the gene's locus\cite{avsecEffectiveGeneExpression2021,linderPredictingRNAseqCoverage2025a,
+hingerlScoobyModelingMultimodal2024, hingerlFlashzoiEnhancedBorzoi2024, lalDecodingSequenceDeterminants2025}.\par
+
+To conclude, we suspect that future models will need to take into account both the effector function of TFs through
+modeling co-factor interactions and the entire gene's locus (including all enhancers and promoter(s)) to effectively
+model GRNs underlying the cell's transcriptome.
 

### 2025-03-25 manuscript.tex
Update references.

@@
 contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$.
-We then used TF-MoDISco\cite{shrikumar2020tfmodisco} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG).
+We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG).
 Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig.
@@
 affinity of each individual instance (making use of the FOXA2 motif score calculated using
-FIMO\cite{grant2011fimo,schreiber2025tangermeme}) hypothesizing that the overall affinity of regions that are generally
+FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
 bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
@@
 \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top).
-\textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grant2011fimo,schreiber2025tangermeme} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
+\textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
 \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
@@
 4f}).
-Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiran2024synthetic, kempynck2025crested} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig.
+Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig.
 4g}).
@@
 5d}).
-This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiran2024synthetic}.\par
+This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
 
@@
 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}.
-The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothstein2020heterodmerization}.
+The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
 After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig.
@@
 shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
-facial mesenchyme\cite{kim2024dnaguided}.\par
+facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par
 
@@
 Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
-that are associated with facial variation\cite{white2021insights}. Given that regions surrounding these SNPs were
-enriched for enhancer activity in cranial neural crest cells\cite{white2021insights} we asked if, using
+that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
+enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
 $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
@@
 \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
-\textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{white2021insights} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome.
+\textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome.
 Lead SNPs are shown with a bigger dot size.
@@
 7a}).
-Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiran2024synthetic, kempynck2025crested}.
+Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}.
 We used at least three TFBS in the design process for each state.
@@
 location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
-for each pattern calculated using FIMO\cite{grant2011fimo, schreiber2025tangermeme} for organoid (top left) and embryo
+for each pattern calculated using FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025} for organoid (top left) and embryo
 (top right) and histograms showing the number of motif instances per 25 bp bins with confidence intervals indicated in
@@
 
-\noindent{The following Cell Press journals require most research articles to include their methods in a \textbf{METHODS} section, which should appear after the discussion: \textit{Cell Biomaterials}, \textit{Chem}, \textit{Chem Catalysis}, \textit{Cell Reports Physical Science}, \textit{Cell Reports Sustainability}, \textit{Device}, \textit{Joule}, \textit{Matter}, \textit{Newton}, \textit{One Earth}, and \textit{Patterns}.
-For all other Cell Press journals, please \textbf{delete} the contents of this page from the template and refer instead to \textbf{STAR Methods} on page 9.}
-
-\section*{METHODS}
-
-%%%  The methods (also/formerly called "experimental 
-%%%  procedures") should appear 
-%%%  immediately after the discussion. Subheadings 
-%%%  may be customized. Please consult your handling 
-%%%  editor if you have questions about what content 
-%%%  should appear here.
-
-\subsection*{Custom methods subheading 1}
-\subsection*{Custom methods subheading 2}
-\subsection*{Custom methods subheading 3}
-\subsection*{Custom methods subheading 4}
-
-\newpage
-
 %%%  The following components should appear after the 
@@
 
-Requests for further information and resources should be directed to and will be fulfilled by the lead contact, Sally White (s.white@university.edu).
+Requests for further information and resources should be directed to and will be fulfilled by the lead contact, Stein Aerts (stein.aerts@kuleuven.be).
 
@@
 
-Plasmids generated in this study have been deposited to [Addgene, name and catalog number or unique identifier].
+This study did not generate new materials
 
@@
 All other data reported in this paper will be shared by the lead contact upon request.
-    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.\cite{georgios_rizos_2023_10253149}
+    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.
     \item Any additional information required to reanalyze the data reported in this paper is available from the lead contact upon request.    

### 2025-03-27 manuscript.tex
add pcc for organoid v embryo
make sure figures are placed niceley

@@
 \usepackage{orcidlink}
+\usepackage{afterpage}
+\usepackage{caption}
 \usepackage[super,comma,sort&compress]  
@@
 organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
-the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of \textbf{X.X}
+the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.892
 across identical states (\textbf{FIGURE SUPPLEMENT}).\par
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
-
-\subsection*{Figure 1. Human neural tube organoids recapitulate early neuronal and neural crest development}
-
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_1.png}
+  \caption{
+    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
     \textbf{a}, Schematic. \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
@@
     (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
-\textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
+    (Continues next page)
+  }
+\end{figure}
+\afterpage{\clearpage}
+
+\begin{figure}[b]
+  \ContinuedFloat
+  \caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+    (Continued) \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
     axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
@@
     microphage/microglia.
+  }
+\end{figure}
 
@@
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
-
-\subsection*{Figure 2. DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer
-code.}
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_2.png}
+  \caption{
+    \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
     \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
@@
     using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
-colored by: cluster identity (\textbf{c}), model of origin (\textbf{d}; only showing \textit{de novo} motifs in this
-case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
+    colored by: cluster identity (\textbf{c}), model of origin (Continues next page)
+  }
+\end{figure}
+\afterpage{\clearpage}
+
+\begin{figure}[b]
+  \ContinuedFloat
+  \caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
  \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
@@
     constitutive enhancer that serves as electroporation control.
+  }
+\end{figure}
 
@@
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
-
-\subsection*{
-  Figure 3. Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.
-}
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_3.png}
+  \caption{
+    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.} \\
     \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics.
 \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells).
-\textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38.
+(Continues next page).
+  }
+\end{figure}
+\afterpage{\clearpage}
+
+\begin{figure}[b!]
+  \ContinuedFloat
+  \caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38.
 SFPE2 and SFPE1 is highlighted.
@@
 \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
+  }
+\end{figure}
 
@@
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
-
-\subsection*{
-Figure 4. A simple enhancer code underlies anterior-posterior patterning in neuronal progenitors.
-}
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_4.png}
+  \caption{
+    \textbf{A simple enhancer code underlies anterior-posterior patterning in neuronal progenitors.} \\
   \textbf{a}, UMAP of embryo neuronal progenitors colored by cell-topic probabilities in RGB color scales for
@@
 .\textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
+  }
+\end{figure}
+\afterpage{\clearpage}
 
@@
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
-
-\subsection*{
-  Figure 5. Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
-}
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_5.png}
+  \caption{
+    \textbf{Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.} \\
 \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics.
@@
 The location of the SNP is indicated with a dotted line.
+  }
+\end{figure}
+\afterpage{\clearpage}
 
@@
 
+\begin{figure}[b!]
   \noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_6.png}
-
-\subsection* {
-  Figure 6. Deciphering the enhancer code of early neuronal differentiation.
-}
-
-\textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics.
+  \captionsetup{justification=raggedright}
+  \caption{ \textbf{Deciphering the enhancer code of early neuronal differentiation.}}
+\end{figure}
+
+\begin{figure}[b!]
+  \ContinuedFloat
+  \caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+  (Continued) \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics.
 \textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic.
@@
 \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neuon related topics.
-\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.}
+\end{figure}
 
@@
 
-\noindent\includegraphics[width=1.0\linewidth]{figures/main/Figure_7.png}
-
-\subsection*{
-  Figure 7. Neural tube cell states are encoded using four to five distinct transcription factor binding
-  sites
-}.
+\afterpage{\clearpage}
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/main/Figure_7.png}
+  \caption{
+    \textbf{Neural tube cell states are encoded using four to five distinct transcription factor binding sites} \\
 \textbf{a}, Area under the receiver-operator curve (outlined dots) and precision-recall curve of logistic regression
@@
 represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
-each topic (bottom). \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
+each topic (bottom). (Continues next page)}
+\end{figure}
+\afterpage{\clearpage}
+
+\newpage
+\vspace*{-\topskip}  
+\begin{figure}[t]
+  \ContinuedFloat
+  %\caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+  (Continued)  \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
 for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
@@
 deviations) for organoid (bottom left) and embryo (bottom right).
-
+}
+\end{figure}
+
+\clearpage
 \section*{DISCUSSION}

### 2025-03-27 manuscript.tex
changes throughout text.

@@
 \author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
+\author[1,2,3,\orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
+\author[1,2,3]{Valerie Christiaens}
 \author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}
@@
 For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
-However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.\par
+However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.
+In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par
 
@@
 Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of each individual cell types are still limited.\par
-
-In this study, we will determine the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.
 
@@
 until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
-neural crest states (REF). In order to characterize the organoid-derived cell types to a greater level of detail at both
+neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. In order to characterize the organoid-derived cell types to a greater level of detail at both
 transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
-scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-110; \textbf{Fig. 1a}),
+scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}),
 resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
@@
 organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning as
-is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal ARX, NKX2-2, PAX6, IRX3, ZIC1 and
-GDF7 (\textbf{Fig. 1d}). In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
-A cluster that still expresses the neuronal progenitor marker SOX2, cells that express the early neuronal induction
-marker ASCL1 and distinct clusters marked by the expression of GAD1, UNCX and GATA2 (\textbf{Fig. 1e}). Finally,
+is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig.
+1d}).
+In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
+A cluster that still expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction
+marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). Finally,
 focusing on the neural crest cells, in both the organoid and the embryo we obtained a population of pre-migratory neural
-crest (GRHL2 expression and absence of ZEB2 expression), migratory neural crest (SOX10 and ZEB2 expression), facial
-mesenchyme cells (TWIST1 expression) and early differentiation peripheral neurons (NEUROD1 expression). In the embryo,
-but not in the organoid, we observed an additional population of cells marked by the expression of SOX10 and PAX2
+crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial
+  mesenchyme cells (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). In the embryo,
+  but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2}
 reflecting the otic placode (\textbf{Fig. 1f}).\par
@@
 assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
-(SIX3) or anterior spinal cord (HOXB3) identity, we observed a gradient of anterior-to-posterior identities in the
-embryo (REF): from anterior to posterior expression of SIX3, EMX2, FOXG1, BARHL1, EN1 and HOXB3 (\textbf{Fig. 1g}).\par
+(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, we observed a gradient of anterior-to-posterior identities in the
+embryo (REF): from anterior to posterior expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par
 Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
@@
 quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
-clear that both early differentiation peripheral neurons and to a greater extend early differentiating neurons of the
+clear that both early differentiating peripheral neurons and to a greater extend early differentiating neurons of the
 central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
@@
 across identical states (\textbf{FIGURE SUPPLEMENT}).\par
+
+% !!! calculate this corr coefficient for topic based states !!!
 
@@
 
-To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube, neural crest
+To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, neural crest-
 and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
@@
 Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
-TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic and
+TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic, to find
+known motifs underlying the accessibility of each region. We
+also
 calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
@@
 on cell type-specific chromatin accessibility.\par
-As an illustrative example, we investigated an enhancer located in one of the introns of GLI3 (hs111.0; \textbf{SUPPLEMENTAL
+As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{SUPPLEMENTAL
 FIGURE}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both
@@
 belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
-enhancers using DeepNeuralTube reveals two FOX (one partial), one SOX and two RFX (one partial) binding sites in SFPE1
-and two FOX and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par
+enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
+and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par
 
@@
 We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG).
-Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code table (\textbf{Fig.
+Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig.
 3e}).
@@
 and RFX instances is important (\textbf{Fig.
-3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA2}, \textit{TEAD1} and \textit{RFX4} in both the organoid and embryo cells (\textbf{Fig.
-3g}).
-In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-2} expression (\textbf{(Fig.
-3e- g)}); PAX instances, correlating with \textit{PAX6} expression (\textbf{Fig.
-3e-g}) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
-3e-g}).
-For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2} expression.
+3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig.
+3g} and SUPPLEMENTAL FIG).
+In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig.
+3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig.
+3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
+3e-g} and SUPPLEMENTAL FIG).
+For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig.
+3e-g} and SUPPLEMENTAL FIG).
 Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig.
-3e}).
+3e} and SUPPLEMENTAL FIG).
 Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
 3h}).
-FOXA2, TEAD1 and RFX4 instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig.
+FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig.
 3h}).\par
 
-It has been reported that FOXA2 is a pioneer factor. Indeed, when centering on FOXA2 instances in both the organoid and
+It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
 embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
@@
 TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
-peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOXA2 instances. Those floorplate peaks
+peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
 that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
-did not we classified as floorplate specific. Next, we quantified both the number of FOXA2 instances per region and the
-affinity of each individual instance (making use of the FOXA2 motif score calculated using
+did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
+affinity of each individual instance (making use of the FOX motif score calculated using
 FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
@@
 the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
-have a single FOXA2 instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
-Furthermore, in case there is but a single FOXA2 instance in a general FOXA2 target the affinity of this instance tends
+have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
+Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
 to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
-have a single instance, we observe that FOXA2 instances in regions that are general tend to be of the form
+have a single instance, we observe that FOX instances in regions that are general tend to be of the form
 \textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
@@
 Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
-from the DV program. In other words, is AP patterning encoded in the same or in different genomic regions as in those in
-which DV patterning is encoded? For this purpose, we calculated the pairwise correlation between region-topic
+from the DV program.
+In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning?
+For this purpose, we calculated the pairwise correlation between region-topic
 probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
 as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
-that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-2}; \textbf{Fig. 4b})\par
+that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
 
@@
 4e}).
-From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA} corresponding with the expression of respectively \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1} and \textit{HOXB3} [ADD GENE EXPRESSION HEATMAP!].
-Finally, SOX2 instances are present in regions of all AP domains (\textbf{Fig.
-4e}).\par
+From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
+Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs.
+Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}.
+Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig.
+4d-e} and SUPPLEMENTAL FIG).
+In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs.
+The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig.
+4d-e} and SUPPLEMENTAL FIG).
 
@@
 4g}).
-[SWITCH PANELS].
 Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig.
@@
 Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class.
-This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig.
+This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig.
 4h}).
@@
 
+\newpage
+
 \subsection*{
@@
 Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
-crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
+crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
 $DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
-topic and generated a code table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
+topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
 $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
@@
 Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig.
-5b-c}) the importance of TEAD instances correlates with \textit{TEAD4} expression, GRHL with both \textit{GRHL1} and \textit{GRHL2} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig.
-5d}).
+5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig.
+5d} and SUPPLEMENTAL FIG).
 Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig.
@@
 The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
-5d}).
+5d} and SUPPLEMENTAL FIG).
 This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
@@
 The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig.
-5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10}.
+5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig.
+5d} and SUPPLEMENTAL FIG).
 The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
 After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig.
-5b-c}) that correlate with the expression of \textit{RFX4} (\textbf{Fig.
-5d}).\par
+5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig.
+5d} and SUPPLEMENTAL FIG).\par
 
@@
 pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
-importance of the FOX instances correlates with the expression of \textit{FOXC2} and that of the dimer motif with the
-expression of both \textit{TWIST1} and \textit{ALX4}. The latter is consistent with a recent study where it has been
-shown that TWIST1 and ALX4 cooperatively bind this specific dimer motif regulating the expression of genes involved in
+importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
+expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The latter is consistent with a recent study where it has been
+shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
 facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par
@@
 index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
-method we found three sets of regions that are predicted to be contain instances of different TFs. One set where
-instances of TEAD4, GRHL1/2 and ZEB1/2 co-occur, one where instances of Nuclear receptor, TFAP2A/B/C and ZIC1-5
-instances co-occur and one set where instances of FOXC2 and TWIST1/ALX4 co-occur.\par
+method we found three sets of regions that are predicted to contain instances of different set of TFs. One set where
+instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
+instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
 
@@
 As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}.
-The G>A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig.
+The G \textgreater A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig.
 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state.
-This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp upsteam of a partial homeobox site potentially creating a new binding site for TWIST1-ALX4 dimers (\textbf{Fig.
+This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig.
 5h}).\par
@@
 instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
-These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD4/1},
-\textit{FOXA2} and \textit{RFX4} (\textbf{Fig.
-6d}).
+These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
+\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig.
+6d} and SUPPLEMENTAL FIG).
 Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig.
 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig.
-6d}).
+6d} and SUPPLEMENTAL FIG).
 Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig.
-6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1}.
-For subsequent topics SOX2 instances are not important anymore (\textbf{Fig.
-6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24\textbf{Fig.
+6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}.
+For subsequent topics SOX instances are not important anymore (\textbf{Fig.
+6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig.
 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig.
-6d}) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
+6d} and SUPPLEMENTAL FIG) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}.
-Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2}.
+Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig.
+6d} and SUPPLEMENTAL FIG).
 Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig.
-6b-c}) correlating with the expression of \textit{GATA2}.\par
+6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig.
+6d} and SUPPLEMENTAL FIG).\par
 
@@
   \captionsetup{justification=raggedright}
-  \caption{ \textbf{Deciphering the enhancer code of early neuronal differentiation.}}
-\end{figure}
-
-\begin{figure}[b!]
-  \ContinuedFloat
-  \caption*{\rule{\linewidth}{0.4pt}}
-  \caption{
-  (Continued) \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics.
+  \caption{ \textbf{Deciphering the enhancer code of early neuronal differentiation.} \\
+ \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics.
 \textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic.
@@
 \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neuon related topics.
-\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.}
+\textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+  }
 \end{figure}
@@
 Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
-tube organoids and four weeks post conception embryo head with almost one-to-one correspondence based on TFBS importances and TF expression.
+tube organoids and the head of a four weeks post conception human embryo with almost one-to-one correspondence based on TFBS importances and TF expression.
 Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo.
-To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as to belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig.
+To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig.
 3e}, \textbf{Fig.
@@
 7a}).
-Next, we selected patterns that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one patterns for fifteen to sixteen states plus the additional four AP states (\textbf{Fig.
+Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig.
 7a}).
-Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}.
+Finally, we used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}.
 We used at least three TFBS in the design process for each state.
-In case less than three distinct patterns had a positive coefficient for a state, we implanted multiple copies of each pattern to come to a total of at least four TFBS.
+In case less than three distinct TFBS had a positive coefficient for a state, we implanted multiple copies of each TFBS to come to a total of at least four TFBS.
 The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig.
@@
 
-To conclude, enhancers specific to DV-patterning, neural crest states and early neuronal differentiation tend to be encoded in
-the human genome using around four to five distinct TFBS (\textbf{Fig.
-7a}) in regions of on average 240-243bp +/- 67-77bp in size (\textbf{Fig.
-7c}).
+To conclude, our results indicate that enhancers specific to DV-patterning, neural crest states and early neuronal differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
+7a}) in regions of on average 134 - 154 bp in size (\textbf{Fig.
+7c}), corresponding to the size of one nucleosome depleted region (NDR).
 AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig.
@@
 TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
-within one nucleosome depleted regions (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
+within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
 the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
@@
 of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
-TWIST1 and ALX4 heterodimer in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest.
+TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest.
 For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances.
@@
 It has been hypothesized that specificity is mainly achieved through cooperative binding with other TFs\cite{mannChapter3Hox2009}.
-Indeed, we observe this for EN1 and HOXB3 that are predicted to co-bind with PAX8 and X respectively forming specific dimer motifs.
-In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for TFX and TFY that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites.
+Indeed, we observe this for EN1/2 and HOXB3 that are predicted to co-bind with PAX5/8 and MEIS1/2 or PBX1/3 respectively forming specific dimer motifs.
+In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for respectively LHX2/9, RAX or SIX3/6 and OTX1/2, EMX2, DMBX1 or LMX1A that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites.
 Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity.
@@
 
-\begin{itemize}
-    \item Small-molecule crystallography data have been deposited at CCDC under the database identifier ABCDEF and are publicly available as of the date of publication.
-1H NMR spectra have been deposited at Mendeley under the DOI 10.12345/a1b2c3d4e5f6g7.1 and are publicly available as of the date of publication.
-All other data reported in this paper will be shared by the lead contact upon request.
-    \item All original code has been deposited at Zenodo under the DOI 10.56789/a1b2c3d4e5f6g7.1 and is publicly available as of the date of publication.
-    \item Any additional information required to reanalyze the data reported in this paper is available from the lead contact upon request.    
-\end{itemize}
-
 \section*{ACKNOWLEDGMENTS}
@@
 %%%  including grant numbers.
-
-This work was funded by [FUNDER] via grant [GRANT NO.]. The authors thank all members of the lab for their support.
-
 \section*{AUTHOR CONTRIBUTIONS}
@@
 
-Conceptualization, S.C.P.
-and S.Y.W.; methodology, A.B., S.C.P., and S.Y.W.; investigation, M.E., A.N.V., N.A.V., S.C.P., and S.Y.W.; writing-–original draft, S.C.P.
-and S.Y.W.; writing-–review \& editing, S.C.P.
-and S.Y.W.; funding acquisition, S.C.P.
-and S.Y.W.; resources, M.E.V and C.K.B.; supervision, A.B., N.L.W., and A.A.D.
+\begin{itemize}
+  \item Conceptualization: S.D.W. \& S.A.
+  \item Experiments and sample preparation: S.D.W., V.C. \& C.M.
+  \item Computational analysis: S.D.W. \& C.M.
+  \item Writing – original draft: S.D.W. \& S.A.
+  \item Writing – review \& editing:
+\end{itemize}
 
@@
 %%%  https://www.cell.com/declaration-of-interests
-
-S.Y.W. is an employee and shareholder of COMPANY. M.E.V. is a founder of COMPANY and a member of its scientific advisory board.
 
@@
 
-During the preparation of this work, the author(s) used [NAME OF TOOL OR SERVICE] in order to [REASON].
-After using this tool or service, the author(s) reviewed and edited the content as needed and take(s) full responsibility for the content of the publication.
+During the preparation of this work, the author(s) did not make use of generative AI.
 
@@
 
-\begin{description}
-  \item Figures S1-S5 and their legends in a PDF
-  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
-  \item Table S2. Another descriptive title for a different Excel file
-  \item Data S1. Raw data on x, y, and z
-\end{description}
+%\begin{description}
+%  \item Figures S1-S5 and their legends in a PDF
+%  \item Table S1. A descriptive title for an Excel file that was too large to appear in the PDF
+%  \item Table S2. Another descriptive title for a different Excel file
+%  \item Data S1. Raw data on x, y, and z
+%\end{description}
 
@@
 
-\subsection*{Table 1. A table with clear organization of data}
-
-\begin{tabular}{|l | l | l | l|} 
- \hline
- \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
- \hline
- Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
- \hline
- Row B & 7 & 78 & 5,415 \\ [1ex] 
- \hline
- Row C & 545 & 778 & 7507 \\ [1ex] 
- \hline
- Row D & 545 & 18,744 & 7,560 \\ [1ex] 
- \hline
- Row E & 88 & 788 & 6,344 \\ [1ex] 
- \hline
-\end{tabular}
+%\subsection*{Table 1. A table with clear organization of data}
+
+%\begin{tabular}{|l | l | l | l|} 
+% \hline
+% \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
+% \hline
+% Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
+% \hline
+% Row B & 7 & 78 & 5,415 \\ [1ex] 
+% \hline
+% Row C & 545 & 778 & 7507 \\ [1ex] 
+% \hline
+% Row D & 545 & 18,744 & 7,560 \\ [1ex] 
+% \hline
+% Row E & 88 & 788 & 6,344 \\ [1ex] 
+% \hline
+%\end{tabular}
 
@@
 
-The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.
-
-\textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.
-
-\textsuperscript{b}More footnotes
+%The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.
+
+%textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.
+
+%\textsuperscript{b}More footnotes
 
@@
 
-All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods.
+%All Cell Press \textbf{life and medical science} journals, and the multidisciplinary journal \textbf{iScience}, use the \textbf{\href{https://www.cell.com/star-authors-guide}{STAR Methods}} format for reporting materials and methods.
 Other Cell Press journals use a non-structured \textbf{methods} format; for those journals, please \textbf{delete} the contents of this page from the template and refer instead to the \textbf{methods} template on page 3.
 
-If you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
-\begin{itemize}
-\item \textit{Cell} authors should download the guide for \textit{Cell} authors \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf}{available here}
-\item All other journal authors should use \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_general-1678470557763.pdf}{this version of the guide}
-\end{itemize}
+I%f you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
+%\begin{itemize}
+%\item \textit{Cell} authors should download the guide for \textit{Cell} authors \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_Cell-1678470557763.pdf}{available here}
+%\item All other journal authors should use \href{https://www.cell.com/pb-assets/journals/research/cell/methods/Methods_Guide_general-1678470557763.pdf}{this version of the guide}
+%\end{itemize}
 
@@
 
-\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
+%\textit{To create the KRT, please use the \href{https://star-methods.com}{KRT webform} or the \href{http://www.cell.com/pb-assets/journals/research/cell/methods/table-template1.docx}{Word template} and upload this file separately.}
 
@@
 
-Cell.com homepage: https://www.cell.com
-\newline Templates for Cell Press authors: https://www.cell.com/article-templates
+%Cell.com homepage: https://www.cell.com
+%\newline Templates for Cell Press authors: https://www.cell.com/article-templates
 

### 2025-03-31 scripts/figure_1/plot_figure.py
Add scripts for figures

@@ -1,0 +1,1801 @@
+# /// script
+# requires-python = "==3.11.*"
+# dependencies = [
+#   "matplotlib",
+#   "scanpy",
+#   "numpy",
+#   "pandas",
+#   "tqdm",
+#   "tangermeme",
+#   "torch",
+#   "scipy",
+#   "harmonypy",
+#   "pycistarget @ git+https://github.com/aertslab/pycistarget",
+#   "setuptools",
+#   "tables",
+#   "h5py",
+#   "igraph",
+#   "leidenalg"
+# ]
+# ///
+
+print("importing modules...")
+
+import matplotlib.pyplot as plt
+import matplotlib
+import scanpy as sc
+import numpy as np
+import pandas as pd
+import json
+from pycistarget.input_output import read_hdf5
+import os
+from tqdm import tqdm
+from tangermeme.tools.tomtom import tomtom
+import torch
+from scipy import sparse
+
+
+def smooth_topics_distributions(
+    topic_region_distributions: pd.DataFrame,
+) -> pd.DataFrame:
+    def smooth_topic_distribution(x: np.ndarray) -> np.ndarray:
+       return x * (np.log(x + 1e-100) - np.sum(np.log(x + 1e-100)) / x.shape[0])
+    smoothed_topic_region_distributions = pd.DataFrame(
+        np.apply_along_axis(
+            smooth_topic_distribution,
+            1,
+            topic_region_distributions.values,
+        ),
+        index=topic_region_distributions.index,
+        columns=topic_region_distributions.columns,
+    )
+    return smoothed_topic_region_distributions
+
+
+###############################################################################################################
+#                                             LOAD DATA                                                       #
+###############################################################################################################
+
+print("LOADING DATA")
+
+# load organoid RNA data and subset for ATAC cells
+adata_organoid = sc.read_h5ad("../data_prep_new/organoid_data/RNA/GEX_adata.h5ad")
+
+sample_id_to_num = {
+    s: adata_organoid.obs.query("sample_id == @s").index[0].split("-")[-1]
+    for s in set(adata_organoid.obs.sample_id)
+}
+
+organoid_atac_cell_names = []
+with open("../data_prep_new/organoid_data/ATAC/cell_names.txt") as f:
+    for l in f:
[…]

### 2025-03-31 scripts/figure_2/plot_figure.py

@@ -1,0 +1,2119 @@
+# /// script
+# requires-python = "==3.11.*"
+# dependencies = [
+#   "matplotlib",
+#   "scanpy",
+#   "numpy",
+#   "pandas",
+#   "tqdm",
+#   "tangermeme",
+#   "torch",
+#   "scipy",
+#   "harmonypy",
+#   "pycistarget @ git+https://github.com/aertslab/pycistarget",
+#   "setuptools",
+#   "tables",
+#   "h5py",
+#   "igraph",
+#   "leidenalg",
+#   "crested",
+#   "scikit-learn",
+#   "tensorflow[and-cuda]"
+# ]
+# ///
+
+print("importing modules...")
+from matplotlib import axes as mpl_axes
+import matplotlib.pyplot as plt
+import pandas as pd
+import numpy as np
+import seaborn as sns
+from scipy.spatial import distance
+from scipy.cluster import hierarchy
+import crested
+import pathlib
+import os
+from tqdm import tqdm
+from functools import cache
+import sklearn
+import json
+from pycistarget.input_output import read_hdf5
+from tangermeme.tools.tomtom import tomtom
+import logomaker
+import h5py
+import modiscolite
+import scanpy as sc
+import torch
+from crested.utils._seq_utils import one_hot_encode_sequence
+from crested.tl._explainer_tf import Explainer
+import tensorflow as tf
+import requests
+from PIL import Image
+import io
+
+sc.settings.figdir = os.path.join(os.getcwd(), "tmp_figs")
+
+
+def topic_name_to_model_index_organoid(t: str) -> int:
+    cell_type = t.split("_Topic")[0]
+    topic = int(t.split("_")[-1])
+    if cell_type == "neuron":
+        return topic - 1
+    elif cell_type == "progenitor":
+        return topic - 1 + 25
+    elif cell_type == "neural_crest":
+        return topic - 1 + 55
+    else:
+        raise ValueError(f"{t} unknown.")
+
+
+def model_index_to_topic_name_organoid(i: int) -> str:
+    if i >= 0 and i < 25:
+        cell_type = "neuron"
+        topic_number = i + 1
+    elif i >= 25 and i < 55:
+        cell_type = "progenitor"
+        topic_number = i + 1 - 25
+    elif i >= 55 and i < 65:
+        cell_type = "neural_crest"
+        topic_number = i + 1 - 55
+    elif i >= 65 and i < 75:
+        cell_type = "pluripotent"
[…]

### 2025-03-31 scripts/figure_3/plot_figure.py

@@ -1,0 +1,3114 @@
+import matplotlib
+import matplotlib.pyplot as plt
+import numpy as np
+import scanpy as sc
+import pandas as pd
+from pycisTopic.topic_binarization import binarize_topics
+import gzip
+import os
+import pyBigWig
+import json
+import tensorflow as tf
+import crested
+import pathlib
+from crested.utils._seq_utils import one_hot_encode_sequence
+from crested.tl._explainer_tf import Explainer
+import logomaker
+from dataclasses import dataclass
+from typing import Self
+import h5py
+from tqdm import tqdm
+import pickle
+from tangermeme.tools.fimo import fimo
+from tangermeme.utils import extract_signal
+import torch
+from functools import reduce
+import seaborn as sns
+from scipy.stats import binomtest
+import networkx as nx
+from scipy.stats import mannwhitneyu
+import modiscolite
+from scipy import stats
+from scipy.stats import fisher_exact
+
+color_dict = json.load(open("../color_maps.json"))
+
+###############################################################################################################
+#                                             LOAD DATA                                                       #
+###############################################################################################################
+
+# load organoid RNA data and subset for ATAC cells
+adata_organoid = sc.read_h5ad("../figure_1/adata_organoid.h5ad")
+
+sample_id_to_num = {
+    s: adata_organoid.obs.query("sample_id == @s").index[0].split("-")[-1]
+    for s in set(adata_organoid.obs.sample_id)
+}
+
+adata_organoid_progenitor = sc.read_h5ad("../figure_1/adata_organoid_progenitor.h5ad")
+adata_embryo_progenitor = sc.read_h5ad("../figure_1/adata_embryo_progenitor.h5ad")
+
+## plot topic contr
+
+
+def topic_name_to_model_index_organoid(t: str) -> int:
+    cell_type = t.split("_Topic")[0]
+    topic = int(t.split("_")[-1])
+    if cell_type == "neuron":
+        return topic - 1
+    elif cell_type == "progenitor":
+        return topic - 1 + 25
+    elif cell_type == "neural_crest":
+        return topic - 1 + 55
+    else:
[…]

### 2025-03-31 scripts/figure_4/plot_figure.py

@@ -1,0 +1,1371 @@
+import matplotlib
+import matplotlib.pyplot as plt
+import scanpy as sc
+import pandas as pd
+import numpy as np
+import seaborn as sns
+from crested.utils._seq_utils import one_hot_encode_sequence
+import pysam
+import tensorflow as tf
+import os
+from dataclasses import dataclass
+import h5py
+import logomaker
+from typing import Self
+from tqdm import tqdm
+import torch
+from tangermeme.tools.tomtom import tomtom
+from tangermeme.tools.fimo import fimo
+from scipy.spatial import distance
+from scipy.cluster import hierarchy
+from sklearn.model_selection import train_test_split
+from sklearn.linear_model import LogisticRegression
+from sklearn.metrics import precision_recall_curve, roc_curve
+import pickle
+import random
+from pycisTopic.topic_binarization import binarize_topics
+
+###############################################################################################################
+#                                                                                                             #
+#                                                   FUNCTIONS                                                 #
+#                                                                                                             #
+###############################################################################################################
+
+
+def rgb_scatter_plot(
+    x,
+    y,
+    r_values,
+    g_values,
+    b_values,
+    ax,
+    g_cut=0,
+    e_thr=0.4,
+    r_name="",
+    g_name="",
+    b_name="",
+    r_vmin=None,
+    r_vmax=None,
+    g_vmin=None,
+    g_vmax=None,
+    b_vmin=None,
+    b_vmax=None,
+):
+    def normalize_channel(values, vmin=None, vmax=None):
+        if vmin is None:
+            vmin = values.min()
+        if vmax is None:
+            vmax = values.max()
+        if vmax > vmin:
+            return np.clip((values - vmin) / (vmax - vmin), 0, 1)
+        else:
+            return values
+    ax.set_axis_off()
+    r_normalized = normalize_channel(r_values, r_vmin, r_vmax)
+    g_normalized = normalize_channel(g_values, g_vmin, g_vmax)
+    b_normalized = normalize_channel(b_values, b_vmin, b_vmax)
[…]

### 2025-03-31 scripts/figure_5/plot_figure.py

@@ -1,0 +1,1638 @@
+import os
+
+os.environ["CUDA_VISIBALE_DEVICES"] = "0"
+import matplotlib.pyplot as plt
+import matplotlib
+import scanpy as sc
+import pandas as pd
+import numpy as np
+from dataclasses import dataclass
+import h5py
+from typing import Self
+import tensorflow as tf
+from tqdm import tqdm
+from crested.tl._explainer_tf import Explainer
+from crested.utils._seq_utils import one_hot_encode_sequence
+import pickle
+import logomaker
+from functools import reduce
+from tangermeme.tools.fimo import fimo
+from tangermeme.utils import extract_signal
+import torch
+import seaborn as sns
+from pycisTopic.topic_binarization import binarize_topics
+import networkx as nx
+from scipy.stats import binomtest
+import json
+from bs4 import BeautifulSoup
+import requests
+import time
+
+###############################################################################################################
+#                                                                                                             #
+#                                                   FUNCTIONS                                                 #
+#                                                                                                             #
+###############################################################################################################
+
+
+def rgb_scatter_plot(
+    x,
+    y,
+    r_values,
+    g_values,
+    b_values,
+    ax,
+    g_cut=0,
+    e_thr=0.4,
+    r_name="",
+    g_name="",
+    b_name="",
+    r_vmin=None,
+    r_vmax=None,
+    g_vmin=None,
+    g_vmax=None,
+    b_vmin=None,
+    b_vmax=None,
+):
+    def normalize_channel(values, vmin=None, vmax=None):
+        if vmin is None:
+            vmin = values.min()
+        if vmax is None:
+            vmax = values.max()
+        if vmax > vmin:
+            return np.clip((values - vmin) / (vmax - vmin), 0, 1)
+        else:
+            return values
+    ax.set_axis_off()
+    r_normalized = normalize_channel(r_values, r_vmin, r_vmax)
+    g_normalized = normalize_channel(g_values, g_vmin, g_vmax)
+    b_normalized = normalize_channel(b_values, b_vmin, b_vmax)
[…]

### 2025-03-31 scripts/figure_6/plot_figure.py

@@ -1,0 +1,1159 @@
+import os
+
+os.environ["CUDA_VISIBALE_DEVICES"] = "0"
+import matplotlib.pyplot as plt
+import matplotlib
+import scanpy as sc
+import pandas as pd
+import numpy as np
+from dataclasses import dataclass
+import h5py
+from typing import Self
+import tensorflow as tf
+from tqdm import tqdm
+from crested.tl._explainer_tf import Explainer
+from crested.utils._seq_utils import one_hot_encode_sequence
+import pickle
+import logomaker
+from functools import reduce
+from tangermeme.tools.fimo import fimo
+from tangermeme.utils import extract_signal
+import torch
+import seaborn as sns
+from pycisTopic.topic_binarization import binarize_topics
+import networkx as nx
+from scipy.stats import binomtest
+import json
+import requests
+import time
+
+###############################################################################################################
+#                                                                                                             #
+#                                                   FUNCTIONS                                                 #
+#                                                                                                             #
+###############################################################################################################
+
+
+def rgb_scatter_plot(
+    x,
+    y,
+    r_values,
+    g_values,
+    b_values,
+    ax,
+    g_cut=0,
+    e_thr=0.4,
+    r_name="",
+    g_name="",
+    b_name="",
+    r_vmin=None,
+    r_vmax=None,
+    g_vmin=None,
+    g_vmax=None,
+    b_vmin=None,
+    b_vmax=None,
+):
+    def normalize_channel(values, vmin=None, vmax=None):
+        if vmin is None:
+            vmin = values.min()
+        if vmax is None:
+            vmax = values.max()
+        if vmax > vmin:
+            return np.clip((values - vmin) / (vmax - vmin), 0, 1)
+        else:
+            return values
+    ax.set_axis_off()
+    r_normalized = normalize_channel(r_values, r_vmin, r_vmax)
+    g_normalized = normalize_channel(g_values, g_vmin, g_vmax)
+    b_normalized = normalize_channel(b_values, b_vmin, b_vmax)
[…]

### 2025-03-31 scripts/figure_7/plot_figure.py

@@ -1,0 +1,1618 @@
+import os
+import matplotlib.pyplot as plt
+import matplotlib
+import scanpy as sc
+import pandas as pd
+import numpy as np
+from dataclasses import dataclass
+import h5py
+from typing import Self
+from tqdm import tqdm
+import pickle
+import logomaker
+from functools import reduce
+import torch
+import seaborn as sns
+from scipy.stats import binomtest
+import json
+from matplotlib.lines import Line2D
+import pysam
+from crested.utils._seq_utils import one_hot_encode_sequence
+from sklearn.model_selection import train_test_split
+from sklearn.linear_model import LogisticRegression
+from sklearn.metrics import precision_recall_curve, roc_curve, auc
+from scipy.stats import norm
+
+##########################################################################################
+#
+#
+#                                           FUNCTIONS
+#
+#
+#########################################################################################
+
+
+@dataclass
+class Seqlet:
+    contrib_scores: np.ndarray
+    hypothetical_contrib_scores: np.ndarray
+    ppm: np.ndarray
+    start: int
+    end: int
+    region_name: str
+    region_one_hot: np.ndarray
+    is_revcomp: bool
+
+    def __init__(
+        self,
+        p: h5py._hl.group.Group,
+        seqlet_idx: int,
+        ohs: np.ndarray,
+        region_names: list[str],
+    ):
+        self.contrib_scores = p["contrib_scores"][seqlet_idx]
+        self.hypothetical_contrib_scores = p["hypothetical_contribs"][seqlet_idx]
+        self.ppm = p["sequence"][seqlet_idx]
+        self.start = p["start"][seqlet_idx]
+        self.end = p["end"][seqlet_idx]
+        self.is_revcomp = p["is_revcomp"][seqlet_idx]
+        region_idx = p["example_idx"][seqlet_idx]
+        self.region_name = region_names[region_idx]
+        self.region_one_hot = ohs[region_idx]
+        if (
+            not np.all(self.ppm == self.region_one_hot[self.start : self.end])
+            and not self.is_revcomp
+        ) or (
+            not np.all(
+                self.ppm[::-1, ::-1] == self.region_one_hot[self.start : self.end]
+            )
+            and self.is_revcomp
+        ):
+            raise ValueError(
+                f"ppm does not match onehot\n"
+                + f"region_idx\t{region_idx}\n"
+                + f"start\t\t{self.start}\n"
[…]

### 2025-04-01 manuscript.tex
add numbers

@@
 (\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{FIGURE SUPPLEMENT})). Of note, a fraction of neural tube
-enhancers (x.x \%) are predicted to be specific to pre-migratory neural crest by both models.\par
+enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par
 Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
@@
     $DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
-    enhancer (chr ...; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
+    enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
     constitutive enhancer that serves as electroporation control.
@@
 underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
-models (CORR COEF; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
+models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
 and RFX instances is important (\textbf{Fig.
@@
     \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics.
-\textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = X, $n_{topic 38}$ = X, $n_{topic 36}$ = X, $n_{topic 54}$ = , and $n_{topic 48}$ = for organoid cells; and $n_{topic 34}$ = X, $n_{topic 38}$ = X, $n_{topic 79}$ = X, $n_{topic 88}$ = , and $n_{topic 58}$ = for embryo cells).
-(Continues next page).
+\textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
+    and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells). (Continues next page).
   }
@@
   \caption{
-    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr:...-...; hg38.
+    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38.
 SFPE2 and SFPE1 is highlighted.
@@
 4i}).
-Surprisingly, this very simple classifier reached high performance with area under the ROC of X and area under the precision recall curve of X (\textbf{Fig.
+Surprisingly, this very simple classifier reached high performance with area under the ROC of 0.82 and area under the precision recall curve of 0.80 (\textbf{Fig.
 4i}).\par
@@
 $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
-these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 5b}).\par
+these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par
 
@@
 suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score).
-At an absolute delta prediction score threshold of 0.5 we identified X SNPs that we deem explainable by either of the two models.
-Note that only very few (X out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig.
+At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models.
+Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig.
 5f}).
@@
 As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}.
-The G \textgreater A substitution of this SNP causes a delta prediction score of 0.X for organoid Topic 58 (\textbf{Fig.
+The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig.
 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state.
@@
 identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
-of these patterns is highly correlated across both models (CORR COEF; \textbf{Fig. 6b}).\par
+of these patterns is highly correlated across both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par
 
@@
 The overall performance of these models was acceptably high (\textbf{Fig.
-7a}) with an average area under the ROC of X and area under the precision-recall curve of X.
+7a}) with an average area under the ROC of 0.77 and area under the precision-recall curve of 0.39.
 Notably, all models for neural crest related states performed best (\textbf{Fig.

### 2025-04-01 manuscript.tex
Add supplement.

@@
 resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
-states were obtained from all three cell lines (\textbf{FIGURE SUPPLEMENT}) demonstrating the robustness of the
+states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the
 organoid system.\par
@@
 \textbf{Fig. 1c}). We annotated both the organoid and embryo cells based on the expression of marker genes
-(\textbf{TABLE SUPPLEMENT}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
+(\textbf{Supplementary Table S1}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
 neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
@@
 organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
-the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.892
-across identical states (\textbf{FIGURE SUPPLEMENT}).\par
-
-% !!! calculate this corr coefficient for topic based states !!!
+the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.83
+across identical states (\textbf{Supplementary Fig. S1c-d}).\par
 
@@
 (auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ (\textbf{Fig. 2b}; and
-area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{FIGURE SUPPLEMENT})), facial mesenchyme
+area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})), facial mesenchyme
 enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$
-(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{FIGURE SUPPLEMENT})). Of note, a fraction of neural tube
+(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). Of note, a fraction of neural tube
 enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par
@@
 clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
-(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{SUPPLEMENTAL TABLE}) and cluster
-26 which is embryo specific (CG-rich motif; \textbf{SUPPLEMENTAL TABLE}). Most clusters (\textbf{Fig. 2f-g} and
-\textbf{SUPPLEMENTAL FIGURE}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
+(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
+26 which is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). Most clusters (\textbf{Fig. 2f-g} and
+\textbf{Supplementary Table S2}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
 Thus, $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ use known-motifs as features to classify regions based
 on cell type-specific chromatin accessibility.\par
-As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{SUPPLEMENTAL
-FIGURE}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both
+As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig.
+S2c}).
+This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig.
+2h}).
+Both
 $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ explain this region similarly (Pearson correlation of x) and
@@
 validated enhancers in the Vista enhancer browser, we selected a genomic region that is specifically accessible in the
-migratory neural crest cell type of the organoid (\textbf{SUPPLEMENTAL FIGURE}). Both models predict this region
-to be specific to the migratory neural crest class (\textbf{SUPPLEMENTAL FIGURE}) and nucleotide importance scores
+migratory neural crest cell type of the organoid (\textbf{Supplementary Fig. S2d}). Both models predict this region
+to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e}) and nucleotide importance scores
 reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). We tested the enhancer activity of this region
@@
 contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$.
-We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG).
+We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities.
 Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig.
@@
 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig.
-3g} and SUPPLEMENTAL FIG).
+3g} and Supplementary Fig.
+S3).
 In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig.
 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig.
-3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
-3e-g} and SUPPLEMENTAL FIG).
+3e-g} and Supplementary Fig.
+S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
+3e-g} and Supplementary Fig.
+S3).
 For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig.
-3e-g} and SUPPLEMENTAL FIG).
+3e-g} and Supplementary Fig.
+S3).
 Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig.
-3e} and SUPPLEMENTAL FIG).
+3e} and Supplementary Fig.
+S3).
 Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
@@
 It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
-embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
+embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
 asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
@@
 Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig.
-4d-e} and SUPPLEMENTAL FIG).
+4d-e} and Supplementary Fig.
+S4).
 In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs.
 The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig.
-4d-e} and SUPPLEMENTAL FIG).
+4d-e} and Supplementary Fig.
+S4).
 
@@
 Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
-crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
+crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
 $DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
@@
 5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig.
-5d} and SUPPLEMENTAL FIG).
+5d} and Supplementary Fig.
+S5).
 Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig.
@@
 The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
-5d} and SUPPLEMENTAL FIG).
+5d} and Supplementary Fig.
+S5).
 This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
@@
 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig.
-5d} and SUPPLEMENTAL FIG).
+5d} and Supplementary Fig.
+S5).
 The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
@@
 5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig.
-5d} and SUPPLEMENTAL FIG).\par
+5d} and Supplementary Fig.
+S5).\par
 
@@
 importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
-expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The latter is consistent with a recent study where it has been
+expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). The latter is consistent with a recent study where it has been
 shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
@@
 5f}).
-Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG).
+Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig.
+S6}).
 As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}.
@@
 \textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig.
-6d} and SUPPLEMENTAL FIG).
+6d} and Supplementary Fig.
+S7).
 Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig.
 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig.
-6d} and SUPPLEMENTAL FIG).
+6d} and Supplementary Fig.
+S7).
 Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig.
@@
 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig.
-6d} and SUPPLEMENTAL FIG) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
+6d} and Supplementary Fig.
+S7) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}.
 Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig.
-6d} and SUPPLEMENTAL FIG).
+6d} and Supplementary Fig.
+S7).
 Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig.
 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig.
-6d} and SUPPLEMENTAL FIG).\par
+6d} and Supplementary Fig.
+S7).\par
 
@@
 For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances.
-Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG).
+Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig.
+8}).
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}.
@@
 repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
-systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024} (SUPPLEMENT FIG).\par
+systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
+(\textbf{Supplementary Fig. S9}).\par
 
@@
 
+\subsection*{Supplementary figures}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\setcounter{figure}{0}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_1.png}
+  \caption{
+    \textbf{Ogranoid cell lines and embryo produce similar gene expression profiles}\\
+    \textbf{a}, UMAP based on gene expression profiles of organoid cells colored by cell identity and split by cell
+    line. \textbf{b}, scatter plot showing gene expression (of highly variable genes) of the same cell type across cell lines. Number top left
+    indicates Pearson correlation coefficient. \textbf{c}, scatter plot showing pairwise gene expression (of highly
+    variable genes) across cell types and organoid and embryo. Number top left indicates Pearson correlation
+    coefficient. \textbf{d}, d histogram showing distribution of Pearson correlation coefficients of average gene
+    expression profile per topic, topics representing
+    the same biological state(matching) and other topics (non-matching) across organoid and embryo.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_2.png}
+  \caption{
+    \textbf{Validation of DeepNeuralTube and profiles of selected enhancers}\\
+    \textbf{a-b}, precision recall curves for neural tube enhancers (left) and facial mesenchyme enhancers (right) for
+    organoid (\textbf{a}) and embryo (\textbf{b}). \textbf{c}, Chromatin accessibility profile of dorso-ventral topics for organoid
+    (top) and embryo (bottom) of region chr7:42,058,384-42,246,450; hg38. Location of HS111 enhancer is indicated. Values are
+    CPM normalized and in range [0, 5]. \textbf{d}, Chromatin accessibility profile of neural crest topics for organoid
+    (top) and embryo (bottom) of region chr1:161,183,163-161,218,934; hg38. Location of tested enhancer (chr1:161200798-161201298) is indicated.
+    Values are CPM normalized and in range [0, 5]. \textbf{e}, Prediction score of $DeepNeuralTube_O$ (top) and
+    $DeepNeuralTube_E$ (bottom) for each topic for neural crest enhancer chr1:161200798-161201298; hg38.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_3.png}
+  \caption{
+    \textbf{Linking motif to TF in dorso-ventral neuronal progenitors}.\\
+    \textbf{a-h} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
+    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
+    that is annotated to: HMG/Sox (\textbf{a}), Forkhead (\textbf{b}), TEA (\textbf{c}), Homeodomain; Paired box or
+    Paired box (\textbf{d}), RFX (\textbf{e}), Homeodomain (\textbf{f}), C2H2 ZF (\textbf{g}) or C2H2 ZF; Homeodomain
+    (\textbf{h}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_4.png}
+  \caption{\textbf{Linking motif to TF in anterior-posterior neuronal progenitors}.\\
+    \textbf{a-e} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
+    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
+    that is annotated to: Homeodomain or Homeodomain; Paired box or Paired box (\textbf{a-e}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_5.png}
+  \caption{
+    \textbf{Linking motif to TF in neural crest}.\\
+    \textbf{a-j} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
+    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
+    that is annotated to:
+    Forkhead (\textbf{a}), 
+    HMG/SOX (\textbf{b}), 
+    TEA (\textbf{c}), 
+    AP-2 (\textbf{d}), 
+    C2H2 ZF; Homeodomain (\textbf{e}), 
+    C2H2 ZF (\textbf{f}), 
+    Nuclear receptor (\textbf{g}), 
+    Grainyhead (\textbf{h}), 
+    bHLH (\textbf{i}), 
+    RFX (\textbf{j}).
+    Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+  }
+\end{figure}
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_6.png}
+  \caption{
+    \textbf{Delta prediction score across topics and SNPs}\\
+    \textbf{a}, heatmap showing delta prediction score across topics using $DeepNeuralTube_O$ (left) and
+    $DeepNeuralTube_E$ (right) for SNPs with a minimal absolute delta prediction score of 0.2.
+  }
+\end{figure}
+
+\afterpage{\clearpage}
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_7.png}
+  \caption{
+    \textbf{Linking motif to TF in neurons}.\\
+    \textbf{a-k} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
+    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
+    that is annotated to (Continues next page):
+ }
+\end{figure}
+\afterpage{\clearpage}
+
+\begin{figure}[b]
+  \ContinuedFloat
+  \caption*{\rule{\linewidth}{0.4pt}}
+  \caption{
+    (Continued)
+    TEA (\textbf{a}), 
+    Forkhead (\textbf{b}), 
+    HMG/SOX (\textbf{c}), 
+    CUT; Homeodomain (\textbf{d}), 
+    Nuclear receptor (\textbf{e}), 
+    C2H2 ZF; Homeodomain (\textbf{f}), 
+    bHLH (\textbf{g}), 
+    EBF (\textbf{h}), 
+    Homeodomain (\textbf{i}), 
+    GATA (\textbf{j}), 
+    RFX ({\textbf{k}}).
+    Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_8.png}
+  \caption{
+    \textbf{SOX dimer instances with 4bp spacing are most accessible in migratory neural crest}\\
+    \textbf{a}, \textit{de novo} identified SOX dimer motifs with 3, 4, or 5 bp spacing. 
+    \textbf{b}, Heatmap showing individual SOX monomer and dimer motif instances and scaled and average chromatin
+    accessibility of regions with those instances for neural crest Topic 60 and Topic 59.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_9.png}
+  \caption{
+    \textbf{ZEB2 is a chromatin closing transcription repressor.}\\
+    \textbf{a}, Network with genes (dots) and regions (outlined-dots) inferred using SCENIC+ colored by gene expression
+    and chromatin accessibility in organoid pre-migratory neural crest cells. 
+    \textbf{b}, scatter plot showing ZEB2 target gene expression (top) and target region accessibility (bottom) as a
+    function of ZEB2 expression. Number on the top right indicates Pearson correlation coefficient.
+  }
+\end{figure}
+
 \end{document}

### 2025-04-02 manuscript.tex
Finalize figures.

@@
 
-\begin{figure}[b]
+\begin{figure}[b!]
   \ContinuedFloat
@@
     microphage/microglia.
+    Color scales for RGB scatter plots are as follows (in Log(CPM)). 
+    For organoid:
+    ARX [0, 1.5],
+    NKX2-2 [0, 1.5],
+    PAX6 [0, 1.5],
+    IRX3 [0, 1.0],
+    ZIC1 [0, 1.5],
+    GDF7 [0, 1.5],
+    SIX3 [0, 1.5],
+    EMX2 [0, 1.5],
+    FOXG1 [0, 0.8],
+    BARHL1 [0, 1.0],
+    EN1 [0, 1.5],
+    HOXB3 [0, 1.5],
+    SOX2 [0, 1.5],
+    ASCL1 [0, 1.5],
+    ISL1 [0, 1.5],
+    GAD1 [0, 0.8],
+    UNXC [0, 0.8],
+    GATA2 [0, 0.8],
+    GRHL2 [0, 1.5],
+    ZEB2 [0, 3.0],
+    SOX10 [0, 0.8],
+    NEUROD1 [0, 1.5],
+    PAX2 [0, 1.5],
+    TWIST1 [0, 1.5]
+    For embryo:
+    ARX [0, 1.5],
+    NKX2-2 [0, 1.5],
+    PAX6 [0, 1.5],
+    IRX3 [0, 1.5],
+    ZIC1 [0, 1.5],
+    GDF7 [0, 1.5]
+    SIX3 [0, 1.5],
+    EMX2 [0, 1.5],
+    FOXG1 [0, 0.8],
+    BARHL1 [0, 1.5],
+    EN1 [0, 1.5],
+    HOXB3 [0, 1.5],
+    SOX2 [0, 1.5],
+    ASCL1 [0, 1.5],
+    ISL1 [0, 1.0]
+    GAD1 [0, 0.8],
+    UNXC [0, 0.8],
+    GATA2 [0, 0.8]
+    GRHL2 [0, 1.5],
+    ZEB2 [0, 3.0],
+    SOX10 [0, 0.8]
+    NEUROD1 [0, 1.5],
+    PAX2 [0, 1.5],
+    TWIST1 [0, 2.0]
   }

### 2025-04-03 manuscript.tex
Start M&M.

@@
 \usepackage{caption}
+\usepackage{siunitx}
+\DeclareSIUnit{\molar}{M}
+\DeclareSIUnit{\units}{U}
 \usepackage[super,comma,sort&compress]  
@@
 
+\subsubsection*{Collection of human fetal samples}
+
+The experiments on human embryos in this study fall under category 2 as defined by the International Society for Stem
+Cell Research (ISSCR) 2021 guidelines. All experiments followed relevant guidelines and regulations, including the ISSCR
+2021 guidelines.
+Human fetal samples were collected from routine termination of pregnancies at the Karolinska University Hospital following informed consent of the donors.
+The donors were informed of the purpose of the research and specific procedures the tissues will be used for.
+The use of fetal samples collected from abortions was approved by the Swedish Ethical Review Authority and the National Board of Health and Welfare (Etikprövningsmyndigheten; DNR2020-02074).
+Embryos were not kept in culture and were immediately dissected after collection.
+The embryo was staged according to the Carnegie stages.
+Following the dissection, the samples were transferred to ice-cold Hibernate E medium (ThermoFisher, A1247601) and processed the same day.
+
+\subsubsection*{Generation of human neural tube organoids}
+The generation of human neural tube organoids falls under category 1a as defined by the ISSCR 2021 guidelines. 
+Two induced Pluripotent Stem Cell lines (iPSC): IPSC EPITHELIAL-1 (\href{https://rrid.site/data/record/SCR_013869-1/CVCL_EE38/resolver?q=SIGi001-A&l=SIGi001-A&i=rrid:cvcl_ee38-ebisc-sigi001-a}{RRID:CVCL\_EE38}; Purchased from Sigma-Aldrich iPSC0028 and hereafter referred to as Sigma iPSC), BJ1 iPSC (iPSC generated from BJ fibroblasts (\href{https://rrid.site/data/record/SCR_013869-1/CVCL_3653/resolver?q=CRL-2522&l=CRL-2522&i=rrid:cvcl_3653-atcc-crl-2522}{RRID:CVCL\_3653}) and kindly gifted by professor Catherine Verfaillie (Leuven Stem Cell Institute)) and one human Embryonic Stem Cell (ESC) line H9-ESC (\href{https://rrid.site/data/record/SCR_013869-1/CVCL_9773/resolver?q=NIHhESC-10-0062&l=NIHhESC-10-0062&i=rrid:cvcl_9773-0}{RRID:CVCL\_9773} and kindly gifted by professor Pierre Vanderhaeghen) were used for the \textit{in vitro} generation of human neural tube organoids.
+The identities of the cell lines were validated based on their genomic sequences (see Methods details below).
+Pluripotent stem cells were cultured in flat bottom, TC-treated 6 well plates (Avantor; 734-2323). For culture details see
+Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven (%TODO Number%)  
+. Human neural tube organoids were generated based on Fattah \textit{et al.} 2021 Nature communications\cite{
+abdelfattahActuationEnhancesPatterning2021} and Ranga \textit{et al.} 2016 PNAS\cite{rangaNeuralTubeMorphogenesis2016}.
+At 60-70\% confluency pluripotent stem cells were washed twice using Phosphate Buffered Saline
+(PBS; ThermoFisher, 14190169; 2ml) next a single cell suspension was made by adding 1ml of room temperature TrypLE
+Express Enzyme (ThermoFisher, 12605010) and incubation for 3 minutes at \SI{37}{\degreeCelsius} followed by pipetting up and down using a
+\SI{1}{\milli\litre} pipette.
+Immediately after the TrypLE reaction was inhibited by adding \SI{2}{\milli\liter} of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)).
+Cells were pelleted at 300 RCF for \SI{3}{\minute} and after counting on a LUNA-FL™ Dual Fluorescence Cell Counter resuspended to \SI{3.5e6}{cells\per\milli\litre} in $GM_0$ ( basal GM (N-2 Supplement (ThermoFisher, 17502048) 1x, B-27 Supplement (ThermoFisher, 17504044) 1x, GlutaMax Supplement (ThermoFisher, 35050061) 1x, Sodium Pyruvate (ThermoFisher, 11360070) \SI{1}{\milli\molar}, MEM Non-Essential Amino Acids Solution (ThermoFisher, 11140050) 1x, Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}, Neurobasal Medium (ThermoFisher, 21103049) \SI{50}{\percent} V/V and DMEM/F12 (ThermoFisher, 11320033) \SI{50}{\percent} V/V) with RHO/ROCK inhibitor (Y-27632 (Dihydrochloride), Stemcell technologies, 100-1044) \SI{1}{\micro\molar}).
+Next, the polyethylene glycol A (PEG-A) hydrogel matrix (see Methods detail below) was prepared by first adding Corning Mouse Laminin (Avantor, 734-1098) \SI{0.1}{\milli\gram\per\milli\liter} followed by activated factor 13 (see Methods details below) in that order to a PEG mixture (see Methods details below; PEG \SI{1.5}{\percent} or \SI{2}{\percent} V/V, buffer (TRIS \SI{0.05}{\milli\molar}, $CaCl_2$ \SI{0.05}{\milli\molar}; pH 7.6; filter sterilized) in $GM_0$ medium).
+After mixing the matrix by pipetting up and down, without generating bubbles, for 1-2 minutes \SI{3.5e4}{cells\per\micro\litre} mixture were added.
+The matrix and gell mixture was mixed gently one more time by pipetting and \SI{10}{\micro\litre} droplets were generated in each well of a TC-treated 96 well plate (Corning, 3603).
+When the matrix started to gell the plate was turned up-side down for 10 seconds, after which it was turned right-side up for 10 seconds.
+This was repeated for a total of 5 times, after which the plates were kept at room temperature.
+After 20 minutes of gelling \SI{200}{\micro\litre} of $GM_0$ medium was added to each well and the organoids were incubated at \SI{37}{\degreeCelsius} \SI{5}{\percent} $CO_2$ for a total of 11 days, or until the timepoint of single-cell multiome.
+At day 3 of differentiation medium was changed with basal GM supplemented with retinoic acid (Sigma-Aldrich, R2625-100MG) \SI{0.25}{\micro\molar} and Smoothened agonist (Sigma-Aldrich, 566661-500UG) \SI{1}{\micro\molar}.
+At day 5, 7 and 9 the medium was changed with basal GM.
+
+\subsubsection*{Chicken embryos}
+Fertilized chicken eggs, obtained from ... %TODO
+were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4.
+Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.
+
 \subsection*{Method details}
@@
 %%%  subjects).
+
+\subsubsection*{Human neural tube organoid culture}
+
+\paragraph*{Factor 13 activation}
+\SI{1}{\units} of reconstituted Thrombin (Sigma-Aldrich, 10602400001) in Ehrbars buffer (TRIS \SI{10}{\milli\molar}; NaCl
+\SI{150}{\milli\molar}; $CaCl_2$ \SI{25}{\milli\molar}; pH 7.4) per \SI{90}{\units} of reconstituted Fibrogammin (UZ Leuven pharmacy)
+was mixed and put in water bath at \SI{37}{\degreeCelsius} for 30 minutes while swirling every 5 minutes. Activated
+factor 13 was filter sterilized, aliquoted and stored at \SI{-80}{\degreeCelsius}.
+
+\paragraph*{PEG-A hydrogel precursor synthesis}
+PEG-A hydrogel precursors, as described in Ranga \textit{et al.} nature
+communications\cite{ranga3DNicheMicroarrays2014}, were kindly gifted by professor Adrian Ranga (Department of Mechanical
+Engineering at KU Leuven).
+
+\paragraph*{Human pluripotent stem cell culture}
+TC-treated well plates were coated using Geltrex Matrix (ThermoFisher, A1413301) as follows. Geltrex Matrix was thawed
+in ice over night at \SI{4}{\degreeCelsius} and diluted in DMEM/F12 (ThermoFisher, 11320033) \SI{1}{\percent} V/V.
+Sufficient amount of diluted Geltrex Matrix was added to each well to coat the surface and the culture plates were
+incubated at \SI{37}{\degreeCelsius} for a minimum of 60 minutes. Remaining Geltrex mixture was removed prior to
+culture. Human Pluripotent stem cells were cultured on the Geltrex Matrix coated well plates at \SI{37}{\degreeCelsius}
+\SI{5}{\percent} $CO_2$ in Essential 8 Flex Medium (ThermoFisher, A2858501) supplemented with Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}.
+Medium was changed every other day.
+Cells were passaged when they reached \SI{70}{\percent} by washing them twice using DPBS without Calcium and Magnesium (ThermoFisher, 14190) and adding EDTA (ThermoFisher, 15575020) \SI{0.5}{\milli\molar} followed by incubation at \SI{37}{\degreeCelsius}.
+After 3-4 minutes the EDTA solution was aspirated and replaced by Essential 8 Flex Medium.
+Cells were scraped using a cell scraper (Corning, 734-0385) and split in a 1:6 ratio.
+
+\paragraph*{single cell suspension}
+PEG hydrogel droplets were removed from the 96 well plate by gently scraping with a \SI{100}{\micro\litre} pipette tip
+and transferred to a tube on ice using a p1000 pipette by putting the pipette tip on the droplet and creating a gentle
+negative pressure, making sure that the droplet does not enter the pipette tip.
+Organoids were dissociated by adding room temperature TrypLE Express Enzyme (ThermoFisher, 12605010) and incubation at \SI{37}{\degreeCelsius} while pipetting up and down until no large pieces of hydrogel were visible anymore or a maximum of 7.5 minutes.
+Immediately after the TrypLE reaction was inhibited by adding twice the volume of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)).
+The single-cell suspension was pelleted at 500 RCF for 4 minutes, resuspended in DMEM/F12 FBS and counted on LUNA-FL™ Dual Fluorescence Cell Counter.
+Immediately after the cells were processed for 10x single-cell multiome.
+
+\subsubsection*{Human neural tube organoid multiome}
+
+\paragraph*{10x multiome library preparation}
+%TODO Valerie
+
+\paragraph*{10x multiome library sequencing}
+%TODO Valerie
+
+\subsubsection*{Human embryo sample preparation}
+
+\paragraph*{10x multiome library preparation}
+%TODO Camiel
+
+\paragraph*{10x multiome library sequencing}
+%TODO Camiel
+
+\subsubsection*{Human neural tube data analysis}
+
+\paragraph*{Samples}
+Eight individual single-cell multiome libraries were generated:
+\begin{itemize}
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
+  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 6 and 7 days of differentiation cultured in \SI{2}{\percent}
+    PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 4, 5, 8 and 9 days of differentiation cultured in
+    \SI{2}{\percent} PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 8 and 9 days of differentiation cultured in \SI{2}{\percent}
+    PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 10 and 11 days of differentiation cultured in \SI{2}{\percent}
+    PEG hydrogel matrix.
+\end{itemize}
+
+For each sample a mix of H9-ESC, Sigma iPSC and BJ1 iPSC cells were used.
+
+\paragraph*{RNA count matrices and ATAC fragment files}
+single-cell RNA count matrices and single-cell ATAC fragment files were generated for each sample using the Cell Ranger
+ARC (v2.0.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_023897/resolver}{RRID:SCR\_023897}) command
+cellranger-arc count command using as the GRCh38 reference genome supplied with Cell Ranger ARC (v2020-A-2.0.0).
+
+\paragraph*{Cell line demultiplexing}
+Variants were called from in house Omni-ATAC-seq\cite{corcesImprovedATACseqProtocol2017} data that was performed on each
+cell line individually (unpublished data). 
+First, bam files per cell line were merged using samtools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002105/resolver}{RRID:SCR\_002105}) merge, sorted using samtools sort and indexed using samtools index.
+Next, for each cell line merged bam file common variants were called using the bcftools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_005227/resolver}{RRID:SCR\_005227}) mpileup command using common short genetic variants from 1000 Genomes project as targets parameter (-T) and 8,000 as max depth (-D) parameter followed by the call command from bcftools.
+
+next ... demuxlet etc
+
+\paragraph*{single-cell RNA quality control}
+
+\paragraph*{single-cell ATAC quality control}
 

### 2025-04-04 manuscript.tex
continue M&M.

@@
 Pluripotent stem cells were cultured in flat bottom, TC-treated 6 well plates (Avantor; 734-2323). For culture details see
-Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven (%TODO Number%)  
+Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven  Number
+%TODO: add ECD
 . Human neural tube organoids were generated based on Fattah \textit{et al.} 2021 Nature communications\cite{
@@
 \subsubsection*{Chicken embryos}
-Fertilized chicken eggs, obtained from ... %TODO
+Fertilized chicken eggs, obtained from ... 
+%TODO: add vendor
 were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4.
@@
 \paragraph*{10x multiome library preparation}
-%TODO Valerie
+%TODO: Valerie
 
 \paragraph*{10x multiome library sequencing}
-%TODO Valerie
+%TODO: Valerie
 
@@
 \paragraph*{10x multiome library preparation}
-%TODO Camiel
+%TODO :Camiel
 
 \paragraph*{10x multiome library sequencing}
-%TODO Camiel
+%TODO: Camiel
 
@@
 \begin{itemize}
-  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix.
-  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
-  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix.
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix (IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG).
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG).
+  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (6-7\_1-2-3\_1 and 6-7\_1-2-3\_2).
   \item A single library containing a mix of organoids at 6 and 7 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix.
+    PEG hydrogel matrix (6-7).
   \item A single library containing a mix of organoids at 4, 5, 8 and 9 days of differentiation cultured in
-    \SI{2}{\percent} PEG hydrogel matrix.
+    \SI{2}{\percent} PEG hydrogel matrix (8-9\_4-5).
   \item A single library containing a mix of organoids at 8 and 9 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix.
+    PEG hydrogel matrix (8-9).
   \item A single library containing a mix of organoids at 10 and 11 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix.
+    PEG hydrogel matrix (10-11).
 \end{itemize}
@@
 cell line individually (unpublished data). 
-First, bam files per cell line were merged using samtools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002105/resolver}{RRID:SCR\_002105}) merge, sorted using samtools sort and indexed using samtools index.
-Next, for each cell line merged bam file common variants were called using the bcftools (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_005227/resolver}{RRID:SCR\_005227}) mpileup command using common short genetic variants from 1000 Genomes project as targets parameter (-T) and 8,000 as max depth (-D) parameter followed by the call command from bcftools.
-
-next ... demuxlet etc
+First, bam files per cell line were merged using samtools\cite{liSequenceAlignmentMap2009} (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002105/resolver}{RRID:SCR\_002105}) merge, sorted using samtools sort and indexed using samtools index.
+Next, for each cell line merged bam file common variants were called using the bcftools\cite{liStatisticalFrameworkSNP2011} (v1.11; \href{https://rrid.site/data/record/nlx_144509-1/SCR_005227/resolver}{RRID:SCR\_005227}) mpileup command using common short genetic variants from 1000 Genomes project as targets parameter (-T) and 8,000 as max depth (-D) parameter followed by the call command from bcftools.
+Next, single-cell multiome samples were demultixpled per cell line using popscle (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026707/resolver}{RRID:SCR\_026707}) demuxlet\cite{kangMultiplexedDropletSinglecell2018} after first sorting the variant VCF file as the single-cell multiome sample bam file, using a custom script sort\_VCF\_same\_as\_bam, filtering the sample bam file to only retain reads that overlap with SNPs in the VCF file and have a valid barcode passing the Cell Ranger ARC quality control, using a custom script filter\_bam\_file\_for\_popscle\_dsc\_pileup, (custom scripts can be found in the popscle\_helper\_tools GitHub repository: \href{https://github.com/aertslab/popscle_helper_tools}{https://github.com/aertslab/popscle\_helper\_tools}) and creating a pileup of reads for each SNP in the VCF file and each cell barcode using the single-cell multiome sample bam file and finally popscle demuxlet was run on the sample pileup, variant VCF file using GT as field (--field) parameter.
 
 \paragraph*{single-cell RNA quality control}
+Cells with less than 3 genes expressed and genes expressed in less than 200 cells were removed using the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions filter\_cells and filter\_genes.
+Next, quality control metrics were calculated using the Scanpy function calculate\_qc\_metrics using qc\_vars=["mt"],
+percent\_top = None and log1p = False as parameters.
+Cells with less than \SI{25}{\percent} mitochondrial counts, more than a total of 2,000 counts and less than a total of
+9,000 genes with counts were kept resulting in a total of 27,552 cells passing quality control.
+
+\paragraph*{single-cell RNA batch effect correction}
+Batch effect correction was performed using harmonypy\cite{korsunskyFastSensitiveAccurate2019} (v0.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_022798/resolver}{RRID:SCR\_022798}) using as batch key the combination of experimental batch and cell line on 50 PCA coordinates calculated using the Scanpy \cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) function pca.
+
+\paragraph*{single-cell RNA cell type annotation}
+Cells were first annotated as Facial mesenchyme, Neural crest, Neuron, Peripheral neuron, Pluripotent stem cell or
+Neuronal progenitor by 
+first normalizing, log transforming and scaling the gene expression matrix using the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions normalize\_per\_cell, log1p and scale
+and clustering the cells using the Leiden algorithm\cite{traagLouvainLeidenGuaranteeing2019} at a resolution of 5.1 (making use of the leiden function from Scanpy), 
+calculating the median expression per cluster, 
+binarizing this matrix using a gene expression threshold of 0, 
+and calculating the jaccard distance using Scipy\cite{virtanenSciPy10Fundamental2020} (v1.9.3; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) between the binarized median expression matrix and a knowledge matrix where containing marker genes per cell type and where the value is 1 for markers of that cell type and 0 otherwise constructed from the marker gene table (\textbf{Supplemental table S1}).
+For each cluster, the annotation with minimal distance was selected breaking ties by selecting a random annotation with equal distance.
+Neuronal progenitors were further annotated as Neuronal progenitor, Neuronal progenitor early neuron, KRT19+ neuronal
+progenitors, FOXG1+ neuronal progenitors, NFE2L3+ neuronal progenitors and KRT19+/NFE2L3+ neuronal progenitors using the
+same procedure at a Leiden resolution of 1.9 after subsetting for neuronal progenitors and re-normalizing, scaling and
+integrating (see batch effect correction) the data.
+non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors were further annotated as FP, p3, p0-2, dp1-6 or RP by subsetting non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors, re-normalizing, scaling and integrating (see batch effect correction) the data.
+Next, a force-directed graph embedding\cite{islamCharacterizationSinglecellTranscriptional2011, jacomyForceAtlas2ContinuousGraph2014} was calculated using the Scanpy function draw\_graph followed by diffusion pseudotime inference\cite{haghverdiDiffusionPseudotimeRobustly2016} using the Scanpy function dpt and binned in 15 and 50 bins using the Pandas\cite{thepandasdevelopmentPandasdevPandasPandas2024} (v1.5.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018214/resolver}{RRID:SCR\_018214}) cut function.
+Using the same procedure as described above cell type annotations were determined for binned pseudotime in 50 bins and
+for each bin of the binned pseudotime in 15 bins the most common annotation was selected.
+Neural crest cells were further annotated as pre-migratory neural crest, bonafide neural crest, EMT neural crest,
+mesenchyme and peripheral neuron manually based on the expression of marker genes.
+Neuron cells and Neuronal progenitor early neuron cells were further annotated as ASCL1+ Neurons, SST+ Neurons, SOX2+ Neurons,
+ASCL1+ Neurons, UNCX+ Neurons, GAD1+ Neurons and GATA2+ Neurons based on the expression of marker genes.
+
+\paragraph*{single-cell ATAC consensus peaks}
+A set of consensus peaks was called by first generating pseudobulk ATAC-seq profiles for each of the highest resolution
+cell types (see single-cell RNA cell type annotation) using the export\_pseudobulk function, with default parameters, from pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023}
+(v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618}) followed by peak calling using MACS\cite{zhangModelbasedAnalysisChIPSeq2008} (v2.2.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_013291/resolver}{RRID:SCR\_013291} ) using the pycisTopic function peak\_calling, using default parameters, and consensus peak generation using the pycisTopic function get\_consensus\_peaks, using a peak width of 500 bp and default parameters.
+This resulted in 488,211 consensus peaks.
 
 \paragraph*{single-cell ATAC quality control}
+Sample and barcode level quality control data was calculated using the compute\_qc\_stats function from
+pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618})
+cells were filtered based on the number of unique fragments per cell, Fraction of Reads in Peaks (FRIP) and
+transcription start site (TSS) enrichment. Following thresholds were used for each sample:
+\begin{itemize}
+  \item sample IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG
+    \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.5
+      \item FRIP: \textgreater 0.5
+      \item TSS enrichment: \textgreater 5
+    \end{itemize}
+  \item sample IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.5
+      \item FRIP: \textgreater 0.5
+      \item TSS enrichment: \textgreater 5
+    \end{itemize}
+  \item sample 10-11
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.25
+      \item FRIP: \textgreater 0.45
+      \item TSS enrichment: \textgreater 6
+    \end{itemize}
+  \item sample 6-7
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.5
+      \item FRIP: \textgreater 0.45
+      \item TSS enrichment: \textgreater 6
+    \end{itemize}
+  \item sample 6-7\_1-2-3\_1
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.75
+      \item FRIP: \textgreater 0.5
+      \item TSS enrichment: \textgreater 7
+    \end{itemize}
+  \item sample 6-7\_1-2-3\_2
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.75
+      \item FRIP: \textgreater 0.5
+      \item TSS enrichment: \textgreater 6
+    \end{itemize}
+  \item sample 8-9
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.25
+      \item FRIP: \textgreater 0.4
+      \item TSS enrichment: \textgreater 7
+    \end{itemize}
+  \item sample 8-9\_4-5
+  \begin{itemize}
+      \item log unique number of fragments: \textgreater 3.75
+      \item FRIP: \textgreater 0.5
+      \item TSS enrichment: \textgreater 6
+    \end{itemize}
+\end{itemize}
+
+Cells that were predicted as doublets by demuxlet (see: Cell line demultiplexing) were also filtered out resulting in a
+total of 28,705 cells passing quality control.
+
+\paragraph*{single-cell ATAC topic modeling}
+Four topic models were generated: one on all cells, one on the subset of cells annotated as neuronal progenitors, one on
+the subset of cells annotated as neuron (Neuron cells and Neuronal progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
+Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
+Topic modeling was performed separately for all cells and for each subset of cells.
+Model evaluation was performed using the pycisTopic function evaluate\_model that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.
+A model of 50 topics was selected for the model containing all cells,
+a model of 30 topics was selected for the model containing neuronal progenitors,
+a model of 25 topics was selected for the model containing neurons and
+a model of 20 topics for the model containing neural crest cells.
+The region topic probability matrix was thresholded using the binarize\_topics function from pycisTopic using the otsu
+method\cite{otsuThresholdSelectionMethod1979} and the cell topic probability matris using the same function using the li
+method\cite{liMinimumCrossEntropy1993}.
+
+\paragraph*{Training $DeepNeuralTube_O$}
+A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
+CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
+1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
+layer of dimension 75 was trained to classify consensus peaks in topics based on their genomic sequence. GELU was used as activaiton function, 17 as
+kernel size, 4 as max pooling size, a dropout of 0.15 was used for the convolutional layers, a dropout of 0.5 was used
+for the dense layers, the convolutional layers were l2 normalized using an epsilon value of 1e-6 and the dense layers were
+l2 normalized using an epsilon value of 1e-3.
+The Adam optimizer was used with a learning rate of 0.001 that was dynamically scaled and binary cross entropy as a loss function and a batch size of 1,024.
+Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
+As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
+neurons and neural crest together with 10 topics form the topic model containing all cells representing the pluripotent
+stem cell state were used for a total of 75 classes/topics. 
+Data was augmented by training on both forward and reverse complement sequence and \SI{10}{\percent} of the data was
+used for validation and \SI{10}{\percent} for testing for a total of 5,192,175 training samples, 650,280 test samples
+and 647,805 validation samples. 
+Based on the validation area under the precision recall curve and validation area under the receiver operator curve the
+model after 23 epochs was selected.
+The classes of the model correspond as follows to the different data subsets: 
+\begin{itemize}
+  \item $0 \leq class_{index} < 25 $: neuron
+  \item $25 \leq class_{index} < 55$: Neuronal progenitor
+  \item $55 \leq class_{index} < 65$: Neural crest
+  \item $65 \leq class_{index} < 75$: pluripotent stem cell
+\end{itemize}
+
+\subsubsection*{Human embryo data analysis}
+
+\paragraph*{RNA count matrices and ATAC fragment files}
+
+% TODO: Camiel
+
+\paragraph*{single-cell RNA quality control}
+
+% TODO: Camiel
+
+\paragraph*{single-cell RNA cell type annotation}
+
+%TODO: Camiel
+
+\paragraph*{single-cell ATAC consensus peaks}
+
+%TODO: Camiel
+
+\paragraph*{single-cell ATAC quality control}
+Sample and barcode level quality control data was calculated using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) command pycistopic qc, using a TSS annotation bed file that was generated using the pycisTopic command pycistopic tss get\_tss using hsapiens\_gene\_ensembl as biomart name.
+High quality cells were selected based on the number of unique fragments per cell and TSS enrichment.
+The minimal number of unique fragments per cell was automatically determined using otsu thresholding \cite{otsuThresholdSelectionMethod1979} and a TSS enrichment threshold of 7 was used.
+This resulted in 38,998 high quality cells passing quality control.
+
+\paragraph*{single-cell ATAC topic modeling}
+Three topic models were generated: one on the subset of cells annotated as neuronal progenitors, on on the subset of
+cells annotated as neuron and one on the subset of cells annotated as neural crest. First, a fragment count matrix was
+generated for each sample using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function create\_fragments\_matrix\_from\_fragments, this matrix was binarized (\textgreater 0), merged in one count matrix using the Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) function sparse.hstack and subsequently split in a matrix for each of neuronal progenitors, neurons and neural crest cells and saved as matrix market format.
+A mallet corpus file was generated for each binary matrix using the pycisTopic command pycistopic topic\_modeling mallet create\_corpus and finally topic modeling was run using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic command: pycistopic topic\_modeling mallet run using 300 iterations and default parameters otherwise.
+Model evaluation was performed using the pycisTopic command pycistopic topic\_modeling mallet stats that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.
+A model of 60 topics was selected for the model containing neuronal progenitors,
+a model of 30 topics was selected for the model containing neurons and
+a model of 30 topics for the model containing neural crest cells.
+The region topic probability matrix was thresholded using the binarize\_topics function from pycisTopic using the otsu
+method\cite{otsuThresholdSelectionMethod1979} and the cell topic probability matris using the same function using the li
+method\cite{liMinimumCrossEntropy1993}.
+
+\paragraph*{Training $DeepNeuralTube_E$}
+A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
+CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
+1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
+layer of dimension 75 was trained to classify consensus peaks in topics based on their genomic sequence. GELU was used as activaiton function, 17 as
+kernel size, 4 as max pooling size, a dropout of 0.15 was used for the convolutional layers, a dropout of 0.5 was used
+for the dense layers, the convolutional layers were l2 normalized using an epsilon value of 1e-6 and the dense layers were
+l2 normalized using an epsilon value of 1e-3.
+The Adam optimizer was used with a learning rate of 0.001 that was dynamically scaled and binary cross entropy as a loss function and a batch size of 1,024.
+Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
+As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
+neurons and neural crest together were used for a total of 120 classes/topics. 
+Data was augmented by training on both forward and reverse complement sequence and \SI{10}{\percent} of the data was
+used for validation and \SI{10}{\percent} for testing for a total of 11,935,075 training samples, 1,494,835 test samples
+and 1,497,350 validation samples. 
+Based on the validation area under the precision recall curve and validation area under the receiver operator curve the
+model after 36 epochs was selected.
+The classes of the model correspond as follows to the different data subsets: 
+\begin{itemize}
+  \item $0 \leq class_{index} < 30 $: neuron
+  \item $30 \leq class_{index} < 90$: Neuronal progenitor
+  \item $90 \leq class_{index} < 120$: Neural crest
+\end{itemize}
+
+\subsubsection*{Cell cycle annotation}
+Marker genes of the S-phase and G2 and M phase of the cell cycle\cite{tiroshDissectingMulticellularEcosystem2016} were
+used in the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.11.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) function score\_genes\_cell\_cycle.
+
+\subsubsection*{Motif enrichment analysis}
+
+A cistarget motif ranking database was created by first getting the genomic sequence of each organoid consensus peak with 1,000 bp background padding using the create\_fasta\_with\_padded\_bg\_from\_bed script from the create\_cisTarget\_databases repository (\href{https://github.com/aertslab/create_cisTarget_databases}{https://github.com/aertslab/create\_cisTarget\_databases}) and scoring each DNA sequence using Cluster-Buster\cite{frithClusterBusterFindingDense2003}(\href{https://rrid.site/data/record/nlx_144509-1/SCR_024810/resolver?}{RRID:SCR\_024810}) using the create\_cisTarget\_databases command create\_cistarget\_motif\_databases followed by the generation of the ranking databases using the command convert\_motifs\_or\_tracks\_vs\_regions\_or\_genes\_scores\_to\_rankings\_cistarget\_dbs.
+Next, the SCENIC+\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0a2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026702/resolver?}{RRID:SCR\_026702}) command scenicplus grn\_inference motif\_enrichment\_cistarget command was used to perform cisTarget\cite{potierUsingCisTargetXPredict2012} motif enrichment analysis for each organoid and embryo topic.
+
+\subsubsection*{Annotating embryo cells based on organoid topics}
+A matrix multiplication between the region-topic probabilities of organoid topics and the binarized embryo fragment
+count matrix was performed to obtain an imputed embryo cell topics probability matrix (where the topics are organoid
+topics). This matrix was binarized by taking the top 1,000 cells per topic.
+
+\subsubsection*{\textit{de novo} motif discovery}
+For all organoid and embryo topics using respectively $DeepNeuralTube_O$ and $DeepNeuralTube_E$ attribution scores
+were calculated using the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function integrated\_grad, on 1,000 regions with the highest prediction score for that topic according to the model.
+Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709).
+First seqlets were called using the tfmodisco-lite function extract\_seqlets using the attribution scores as input and using parameters
+sliding\_window\_size = 15, 
+flank = 5, 
+suppress = 12, 
+target\_fdr = 0.2, 
+min\_passing\_windows\_frac = 0.03, 
+max\_passing\_windows\_frac = 0.3 and 
+weak\_threshold\_for\_counting\_sig = 0.8. 
+Next, the attribution score of each seqlet was extracted based on the seqlet coordinates and negative and positive seqlets were defined based on the sum of the attribution score without the flank filtering out seqlets for which the abolute value of the sum of the attribution score was lower than or equal to X.
+Finally, separately for positive and negative seqlets 1,000 seqlets were sampled and used in the tfmodisco-lite function seqlet\_to\_patterns using parameters 
+min\_overlap\_while\_sliding = 0.7,
+nearest\_neighbors\_to\_compute = 500,
+affmat\_correlation\_threshold = 0.15,
+tsne\_perplexity = 10,
+n\_leiden\_iterations = -1,
+n\_leiden\_runs = 50,
+frac\_support\_to\_trim\_to = 0.2,
+min\_num\_to\_trim\_to = 30,
+trim\_to\_window\_size = 20,
+initial\_flank\_to\_add = 5,
+prob\_and\_pertrack\_sim\_merge\_thresholds = [(0.8,0.8), (0.5, 0.85), (0.2, 0.9)],
+prob\_and\_pertrack\_sim\_dealbreaker\_thresholds = [(0.4, 0.75), (0.2,0.8), (0.1, 0.85), (0.0,0.9)],
+subcluster\_perplexity = 50,
+merging\_max\_seqlets\_subsample = 300,
+final\_min\_cluster\_size = 20,
+min\_ic\_in\_window = 0.6,
+min\_ic\_windowsize = 6 and 
+ppm\_pseudocount= 0.001.
+Next, similar patterns were clustered together by first calculating a pairwise pattern-to-pattern similarity metric
+using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}).
+The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID:SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}).
+
+\subsubsection*{Validation of DeepNeuralTube organoid based on Vista enhancers}
+Data was downloaded from the VISTA Enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025} (\href{https://gitlab.com/egsb-mfgl/vista-data/-/raw/main/experiments.tsv.gz}{https://gitlab.com/egsb-mfgl/vista-data/-/raw/main/experiments.tsv.gz}) and the DNA sequence of each element was obtain either from the human genome (hg38) or the mouse genome (mm10) dependent on the origin of each sequence.
+In case a sequence was shorter than 500 bp (the input size of DeepNeuralTube) the sequence was extended to 500 bp based on the genomic coordinates.
+In case the sequence was larger than 500 bp a sliding window, with step size of 1, was produced over the sequence.
+A prediction score for each sequence and using both $DeepNeuralTube_O$ and $DeepNeuralTube_E$ was calculated and per experiment and output class of the model the maximum prediction score was retained.
+Ground truth neural tube enhancers were defined as enhancers that were positive in the tissue "nt", in at least 3 embryos and at
+least \SI{50}{\percent} of tested embryos. 
+Ground truth facial mesenchyme enhancers were defined as enhancers that were positive in the tissue "fm", in at least 3 embryos and at
+least \SI{50}{\percent} of tested embryos. 
+For both neural tube enhancers and facial mesenchyme enhancers precision-recall curves and receiver-operator curves were
+calculated using respectively the function metrics.precision\_recall\_curve and metrics.roc\_curve from Scikit-learn
+(v1.5.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002577/resolver}{RRID:SCR\_002577}) using the maximum
+prediction score across classes related to neural tube respectively facial mesenchyme in both organoid and embryo. 
+We consider classes related to neural tube in the organoid as classes of the following topics: 
+neural crest Topic 5,
+neural crest Topic 7,
+neuron Topic 1,
+neuron Topic 2,
+neuron Topic 3,
+neuron Topic 4,
+neuron Topic 6,
+neuron Topic 10,
+neuron Topic 11,
+neuron Topic 12,
+neuron Topic 13,
+neuron Topic 15,
+neuron Topic 16,
+neuron Topic 18,
+neuron Topic 19,
+neuron Topic 20,
+neuron Topic 21,
+neuron Topic 23,
+neuron Topic 24,
+neuron Topic 25,
+progenitor Topic 1,
+progenitor Topic 3,
+progenitor Topic 8,
+progenitor Topic 9,
+progenitor Topic 11,
+progenitor Topic 13,
+progenitor Topic 14,
+progenitor Topic 16,
+progenitor Topic 19,
+progenitor Topic 21,
+progenitor Topic 23,
+progenitor Topic 24,
+progenitor Topic 25,
+progenitor Topic 29 and
+progenitor Topic 30
+We consider classes related to neural tube in the embryo as classes of the following topics: 
+│   │   "neuron_Topic_1",  #
+     7   │   │   "neuron_Topic_3",  #
+     6   │   │   "neuron_Topic_5",  #
+     5   │   │   "neuron_Topic_6",  #
+     4   │   │   "neuron_Topic_7",
+     3   │   │   "neuron_Topic_8",  #
+     2   │   │   "neuron_Topic_9",  #
+     1   │   │   "neuron_Topic_10",
+  1016   │   │   "neuron_Topic_11",  #
+     1   │   │   "neuron_Topic_12",
+     2   │   │   "neuron_Topic_13",  #
+     3   │   │   "neuron_Topic_14",
+     4   │   │   "neuron_Topic_15",  #
+     5   │   │   "neuron_Topic_17",  #
+     6   │   │   "neuron_Topic_18",  #
+     7   │   │   "neuron_Topic_19",  #
+     8   │   │   "neuron_Topic_22",  #
+     9   │   │   "neuron_Topic_24",  #
+    10   │   │   "neuron_Topic_26",  #
+    11   │   │   "neuron_Topic_27",  #
+    12   │   │   "neuron_Topic_29",
+    13   │   │   "neuron_Topic_30",
+    14   │   │   "progenitor_Topic_1",  #
+    15   │   │   "progenitor_Topic_3",
+    16   │   │   "progenitor_Topic_4",  #
+    17   │   │   "progenitor_Topic_8",  #
+    18   │   │   "progenitor_Topic_10",
+    19   │   │   "progenitor_Topic_11",
+    20   │   │   "progenitor_Topic_16",
+    21   │   │   "progenitor_Topic_17",
+    22   │   │   "progenitor_Topic_21",  #
+    23   │   │   "progenitor_Topic_22",  #
+    24   │   │   "progenitor_Topic_28",  #
+    25   │   │   "progenitor_Topic_29",  #
+    26   │   │   "progenitor_Topic_31",  #
+    27   │   │   "progenitor_Topic_32",  #
+    28   │   │   "progenitor_Topic_36",
+    29   │   │   "progenitor_Topic_40",  #
+    30   │   │   "progenitor_Topic_41",  #
+    31   │   │   "progenitor_Topic_44",  #
+    32   │   │   "progenitor_Topic_45",
+    33   │   │   "progenitor_Topic_49",  #
+    34   │   │   "progenitor_Topic_51",
+    35   │   │   "progenitor_Topic_57",  #
+    36   │   │   "progenitor_Topic_58",  #
+    37   │   │   "progenitor_Topic_59",
+    38   │   │   "neural_crest_Topic_12",  #
+    39   │   │   "neural_crest_Topic_13",  #
+    40   │   │   "neural_crest_Topic_15",
+    41   
+We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
+neural crest Topic 3
+We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
+neural crest Topic 1, 
+neural crest Topic 19, 
+neural crest Topic 21 and 
+neural crest Topic 29
 

### 2025-04-04 manuscript.tex
continue M&M.

@@
 We consider classes related to neural tube in the embryo as classes of the following topics: 
-│   │   "neuron_Topic_1",  #
-     7   │   │   "neuron_Topic_3",  #
-     6   │   │   "neuron_Topic_5",  #
-     5   │   │   "neuron_Topic_6",  #
-     4   │   │   "neuron_Topic_7",
-     3   │   │   "neuron_Topic_8",  #
-     2   │   │   "neuron_Topic_9",  #
-     1   │   │   "neuron_Topic_10",
-  1016   │   │   "neuron_Topic_11",  #
-     1   │   │   "neuron_Topic_12",
-     2   │   │   "neuron_Topic_13",  #
-     3   │   │   "neuron_Topic_14",
-     4   │   │   "neuron_Topic_15",  #
-     5   │   │   "neuron_Topic_17",  #
-     6   │   │   "neuron_Topic_18",  #
-     7   │   │   "neuron_Topic_19",  #
-     8   │   │   "neuron_Topic_22",  #
-     9   │   │   "neuron_Topic_24",  #
-    10   │   │   "neuron_Topic_26",  #
-    11   │   │   "neuron_Topic_27",  #
-    12   │   │   "neuron_Topic_29",
-    13   │   │   "neuron_Topic_30",
-    14   │   │   "progenitor_Topic_1",  #
-    15   │   │   "progenitor_Topic_3",
-    16   │   │   "progenitor_Topic_4",  #
-    17   │   │   "progenitor_Topic_8",  #
-    18   │   │   "progenitor_Topic_10",
-    19   │   │   "progenitor_Topic_11",
-    20   │   │   "progenitor_Topic_16",
-    21   │   │   "progenitor_Topic_17",
-    22   │   │   "progenitor_Topic_21",  #
-    23   │   │   "progenitor_Topic_22",  #
-    24   │   │   "progenitor_Topic_28",  #
-    25   │   │   "progenitor_Topic_29",  #
-    26   │   │   "progenitor_Topic_31",  #
-    27   │   │   "progenitor_Topic_32",  #
-    28   │   │   "progenitor_Topic_36",
-    29   │   │   "progenitor_Topic_40",  #
-    30   │   │   "progenitor_Topic_41",  #
-    31   │   │   "progenitor_Topic_44",  #
-    32   │   │   "progenitor_Topic_45",
-    33   │   │   "progenitor_Topic_49",  #
-    34   │   │   "progenitor_Topic_51",
-    35   │   │   "progenitor_Topic_57",  #
-    36   │   │   "progenitor_Topic_58",  #
-    37   │   │   "progenitor_Topic_59",
-    38   │   │   "neural_crest_Topic_12",  #
-    39   │   │   "neural_crest_Topic_13",  #
-    40   │   │   "neural_crest_Topic_15",
-    41   
+neuron Topic 1, 
+neuron Topic 3, 
+neuron Topic 5, 
+neuron Topic 6, 
+neuron Topic 7,
+neuron Topic 8, 
+neuron Topic 9, 
+neuron Topic 10,
+neuron Topic 11, 
+neuron Topic 12,
+neuron Topic 13, 
+neuron Topic 14,
+neuron Topic 15, 
+neuron Topic 17, 
+neuron Topic 18, 
+neuron Topic 19, 
+neuron Topic 22,
+neuron Topic 24, 
+neuron Topic 26, 
+neuron Topic 27, 
+neuron Topic 29,
+neuron Topic 30,
+progenitor Topic 1, 
+progenitor Topic 3,
+progenitor Topic 4, 
+progenitor Topic 8, 
+progenitor Topic 10,
+progenitor Topic 11,
+progenitor Topic 16,
+progenitor Topic 17,
+progenitor Topic 21, 
+progenitor Topic 22, 
+progenitor Topic 28, 
+progenitor Topic 29, 
+progenitor Topic 31, 
+progenitor Topic 32, 
+progenitor Topic 36,
+progenitor Topic 40, 
+progenitor Topic 41, 
+progenitor Topic 44, 
+progenitor Topic 45,
+progenitor Topic 49, 
+progenitor Topic 51,
+progenitor Topic 57, 
+progenitor Topic 58, 
+progenitor Topic 59,
+neural crest Topic 12, 
+neural crest Topic 13, 
+neural crest Topic 15,
 We consider classes related to facial mesenchyme in the organod as classes of the following topics: 

### 2025-04-07 manuscript.tex
M&M Seppe done.

@@
 For both neural tube enhancers and facial mesenchyme enhancers precision-recall curves and receiver-operator curves were
-calculated using respectively the function metrics.precision\_recall\_curve and metrics.roc\_curve from Scikit-learn
+calculated using respectively the function metrics.precision\_recall\_curve and metrics.roc\_curve from
+Scikit-learn\cite{JMLR:v12:pedregosa11a}
 (v1.5.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002577/resolver}{RRID:SCR\_002577}) using the maximum
@@
 progenitor Topic 29 and
-progenitor Topic 30
+progenitor Topic 30. 
 We consider classes related to neural tube in the embryo as classes of the following topics: 
@@
 neural crest Topic 13, 
-neural crest Topic 15,
-We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
-neural crest Topic 3
+neural crest Topic 15. 
+We consider classes related to facial mesenchyme in the organod as classes of the following topic: 
+neural crest Topic 3.
 We consider classes related to facial mesenchyme in the organod as classes of the following topics: 
@@
 neural crest Topic 21 and 
-neural crest Topic 29
+neural crest Topic 29.
+
+\subsubsection*{Clustering of \textit{de novo} and enriched motifs}
+A pairwise similarity matrix between all enriched motifs (see Motif enrichment analysis) and \textit{de novo} motifs was
+calculated using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}).
+The Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.11.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions pca, neighbors, tsne and leiden were used to generate a tSNE-embedding of the motifs based on 50 principle components and to cluster the motifs at a leiden resolution of 2.
+
+\subsubsection*{Chicken embryo enhancer reporter assay}
+
+\paragraph*{Enhancer cloning}
+
+Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
+Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq.
+Using PCR.
+A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and
+%TODO: Add concentration
+PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
+\SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
+for 5 minutes.
+Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
+hour at \SI{37}{\degreeCelsius} and gel extraction of 4,5 kb band using QIAquick Gel Extraction Kit (Qiagen, 28704).
+Enhancer candiates were cloned into the target vector using an NEBuilder (New England Biolabs, E5520S) reaction with a linarized plasmid to insert ratio of 1:2 (45 minutes at \SI{50}{\degreeCelsius}).
+Next, \SI{2.5}{\micro\litre} of the rection mixture was transformed into \SI{25}{\micro\litre} Stellar Competent Cells (Takara Bio, 636763).
+Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice.
+Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
+over night.
+Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
+concentration of \SI{4}{\micro\gram\per\micro\litre}.
+
+\paragraph*{Injection control cloning}
+Plasmid pCI H2B-RFP (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_92398/resolver}{RRID:Addgene\_92398})
+was ordered from Addgene (\href{https://rrid.site/data/record/nlx_144509-1/SCR_002037/resolver}{RRID:SCR\_002037}),
+transformed in Stellar Competent Cells (Takara Bio, 636763).
+Cells were thawed 15 minutes on ice, plasmid + cell mixture was kept for an additional 30 minutes on ice followed by 45 second heat shot at \SI{42}{\degreeCelsius} and 5 minutes on ice.
+Cells were incubutated for 1 hour at \SI{37}{\degreeCelsius}, single colonies were then grown on Carbencillin plates
+over night.
+Finally, plasmids were purified using NucleoBond Xtra Maxi Plus Kit (Macherey-Nagel, 740416.50) and eluded at a target
+concentration of \SI{4}{\micro\gram\per\micro\litre}.
+
+\paragraph*{\textit{ex ovo} enhancer-reporter injection}
+Preparation of chicken embryos and injection was performed as described in Williams and Sauka-Spengler 2021\cite{
+williamsExOvoElectroporation2021}.
+Eggs were incubated as described above (see Chicken embryos). A hole was created in the egg shell using the back of
+foreceps, thick albumin was discarded and thin albumin was collected. When most albumin was removed, the egg shell was
+cut open and yolk turned so that embryo faces upwards. More albumin was removed from the yolk by stroking the yolk using
+serrated forceps, moving away from (and not touching) the embryo. A Whatman filter paper with perforated hole was placed
+on top of the embryo and yolk and whatman paper was cut using dissection scissors and embryo was removed and placed in
+Ringer solution 
+(
+  $NaCl_2$ \SI{62.5}{\micro\molar}; 
+  KCl \SI{5}{\micro\molar}; 
+  $CaCl_2\cdot2H_2O$ \SI{1.5}{\micro\molar};
+  $NaHPO_4\cdot7H_2O$ \SI{0.75}{\micro\molar};
+  $KH_2PO_4$ \SI{0.15}{\micro\molar}
+).
+Embryo was placed in custom elektroporation cuvette (see Williams and Sauka-Spengler 2021\cite{
+williamsExOvoElectroporation2021}), enhancer-reporter plasmid mixture (
+enhancer-reporter plasmid \SI{2}{\micro\gram\per\micro\litre};
+injection control plasmid \SI{1}{\micro\gram\per\micro\litre};
+food coloring
+) was injected using micro injector in the cavity between the blastoderm and underlying viteline membrane and
+embryo was elektroorated (5 pules of 5V: 100 ms on, 50 ms off). Embryo was grown for 24 hours at \SI{37}{\degreeCelsius}
+and imaged whole mount
+
+\subsubsection*{Code table}
+For each pattern of the same cluster (see \textit{de novo} motif discovery) the optimal pairwise alignment was calculated using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}).
+A seed pattern was selected, for each clusters, as the patterning with the highest information content.
+All patterns were aligned to the seed pattern and the average contribution score for each nucleotide and position was calculated for all pattern instances (seqlets) of the cluster and using all relevant classes of both models.
+
+\subsubsection*{Footprinting analysis}
+TN5 cut site BigWigs were calculated for all organoid topics using the bigwig command from
+scatac\_fragment\_tools ()
+%TODO: add RRID
+using the --cut-sites option.
+Next, for each pattern (converted to a PWM) of the FOX cluster of neuronal progenitors (cluster 6) motif hits were calculated on the top
+1,000 regions of all neuronal progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
+Overlapping hits were resolved by taking the hit with the highest $-log_{10}(pvalue)$. 
+For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
+region originated, was retained.
+FOX patterns were aligned using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) to a seed pattern defined as the pattern with the highest information content.
+Finally, regions were centered on the pattern hit position and TN5 cute site coverage was calculated using pyBigWig
+(v0.3.24; \href{https://rrid.site/data/record/nlx_144509-1/SCR_024807/resolver}{RRID:SCR\_024807}).
+
+\subsubsection*{General and specific FOXA2 target regions}
+FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793}).
+For each pattern (converted to a PWM) of all neuronal progenitor pattern cluster motif hits were calculated on the top
+1,000 regions of all neuronal progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
+For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
+region originated, was retained.
+Attribution scores were normalized by dividing over the sum of attribution scores over all patterns per sequence.
+FOX hits were defined as hits with a value greater than 0.0008 in organoid and 0.0004 in embryo.
+To define general hits, the FOX hits were intersected with FOXA2 ChIP-seq peaks in HepG2 and A549 using bedtools (v2.31;
+\href{https://rrid.site/data/record/nlx_144509-1/SCR_006646/resolver?q=bedtools&l=bedtools&i=rrid:scr_006646}{RRID:SCR\_006646}) intersect using option -F 0.4 and peaks intersecting with either of the two ChIP-seq peak sets were defined as general.
+
+\subsubsection*{Anterior-posterior HOX motif replacement}
+For all patterns (convered to a PWM and trimmed based on information content using a threshold of 0.1) belonging to cluster 4 or 8 of anterior-posterior \textit{de novo} patterns hits
+were calculated on the top, based on region-topic probabilities, 3,000 regions for topic 31 and topic 61 using  FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
+Hits for cluster 4 were replaced to 5\lq-AGGGATTAG-3\lq and hits for cluster 8 were replaced to 5\lq-CTAATTAG-3\lq.
+If the hit was reverse complement, the reverse complement of those sequences was implanted.
+
+\subsubsection*{Anterior-posterior synthetic enhancer design}
+5\lq-GGATTAG-3\lq and 5\lq-CTAATTAG-3\lq were used as motifs in the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function enhancer\_design\_motif\_insertion to design 200 synthetic enhancers of length 500 for both topic 31 and 61 using up to 5 motif instances.
+
+\subsubsection*{Anterior-posterior logistic regression model}
+Motif scores for all patterns (convered to a PWM) belonging to cluster 4 or 8 of anterior-posterior \textit{de novo} patterns hits
+were calculated on the top, based on region-topic probabilities, 3,000 regions for topic 31 and topic 61 using  FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using threshold parameter set to 0.5 to make sure each region gets a score for at least one pattern per cluster.
+For each sequence and cluster the maximum score, based on the $-log_{10}(p-value)$ of the hit, was retained.
+Regions were split in train and test set using the train\_test\_split function from
+Scikit-learn\cite{JMLR:v12:pedregosa11a}
+(v1.5.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002577/resolver}{RRID:SCR\_002577}) using test size of
+\SI{33}{\percent} and random state 42.
+A logistic regression model was fit (on training regions) to predict the difference between the log of topic 31 and topic 61 region topic
+probabilities making use of the cluster 4 and cluster 8 motif scores as features.
+
+\subsubsection*{Scoring facial GWAS variants}
+SNPs reaching at least suggestive significance and lead SNPs were downloaded from Figshare doi: \href{https://doi.org/10.6084/m9.figshare.c.4667261.v7}{https://doi.org/10.6084/m9.figshare.c.4667261.v7}\cite{whiteInsightsGeneticArchitecture2021}.
+For each SNP the hg38 genomic coordinate and alternative alleles was downloaded using the Entrez Utilities (\href{https://rrid.site/data/record/nlx_144509-1/SCR_013249/resolver}{RRID:SCR\_013249}) API, scored using both $DeepNeuralTube_O$ and $DeepNeuralTube_E$ and difference between the prediction of reference and alternative allele was calculated (delta prediction score).
+
+\subsubsection*{logistic regression models for each cell states}
+For each cell type, neuronal progenitors dorso-ventral patterning, neuronal progenitors anterior-posterior patterning, neurons,
+and neural crest, motif scores for all patterns (converted to a PWM) patterns belonging to clusters of the cell type\lq s
+code-table hits were calculated on the top, based on region-topic probabilities, 1,000 regions for all topics of that
+cell type\lq s code-table using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using threshold parameter set to 0.5 to make sure each region gets a score for at least one pattern per cluster.
+For each sequence and cluster the maximum score, based on the $-log_{10}(p-value)$ of the hit, was retained.
+Regions were split in train and test set using the train\_test\_split function from
+Scikit-learn\cite{JMLR:v12:pedregosa11a}
+(v1.5.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_002577/resolver}{RRID:SCR\_002577}) using test size of
+\SI{33}{\percent} and random state 42 for each topic.
+For each topic per cell type a logistic regression model was fit (on training regions) to predict the difference between
+the log of the region-topic probability and the maximum, across all other topics of that cell type, log of the
+region-topic probability. Making use of the motif scores as features.
 
@@
 %%%  assumptions of the statistical approach.
+
+Statistical details including the statistical test used, exact sample size, definition of center, dispersion and
+precision measures and the method used to determine significance can be found in the figure legends.
+%TODO: make sure it is there.
 

### 2025-04-08 manuscript.tex
C.M. comments

@@
 \author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
-\author[1,2,3,\orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
+\author[1,2,3, 4, \orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
 \author[1,2,3]{Valerie Christiaens}
+\author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
+\author[5, \orcidlink{0000-0003-2931-8015}]{Erik Sundström}
+\author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
 \author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}
@@
 \affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
+\affil[4]{Karolinska Insitute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
+\affil[5]{Karolinska Institutet, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}
 
@@
 Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
-genome. Especially organoids derived from pluripotent stem cells are interesting in this context, given that they
+genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
 demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
-takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore, give insights into how genomes
+takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
 interpret both cell intrinsic and extrinsic signals. For example, epigentic
@@
 model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
-example, early human development.\par
+example, early human development.
+One such system is early human development, the study of which is especially important for our understanding of human health and disease.
+\par
 
@@
 
-Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts, at a transcriptomic level.
-For example, for brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
+Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
 However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.
@@
 
-During vertebrate development both the neural tube and neural crest originate from the neural ectoderm. The neural tube
-consist of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo later giving rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}.
+During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
+consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo.
+These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}.
 Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}.
 The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied.
-Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of each individual cell types are still limited.\par
+Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of individual cell types are still limited.\par
 
@@
 organoid system.\par
-To evaluate the biological relevance of the cell types and states obtained from the organoid, we performed 10x single
+To evaluate the biological relevance of the cell types and states obtained from the organoid, we also performed 10x single
 cell multiome on the head of a single four weeks post conception human embryo, resulting in 25,785 high quality cells (
@@
 quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
-clear that both early differentiating peripheral neurons and to a greater extend early differentiating neurons of the
+clear that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the
 central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
@@
 across identical states (\textbf{Supplementary Fig. S1c-d}).\par
+
+%TODO:D) The black outline on green text (EMX2/EN1 etc) makes it difficult to read and see the color. consider making it thinner
 
@@
 enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par
+%TODO: C.M.: I think the original VISTA enhancers were annotated based on localization only, so it would make sense that premig.
+NC would be very hard to differentiate from normal neural tube ectoderm.
+Might be good to point out why this overlap exists?
 Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
@@
 S4).
+%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
 
 Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
-accessibility of those domains.
-Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}.
+accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides
+\textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. 
+%TODO: Do we know which TFs these correspond to? If you do, mention it in the text
 We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains.
@@
 topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
-$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
+$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances of
 these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par
@@
 After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig.
-5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig.
+5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig.
 5d} and Supplementary Fig.
@@
 instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
+%TODO: C.M.: Mesenchyme+NC or calculated over all regions?
 
@@
 For subsequent topics SOX instances are not important anymore (\textbf{Fig.
-6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig.
+6b-c}) instead we find topics where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig.
 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig.
@@
 S7).\par
+%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
+That you don't see it in the organoid might indicate that part of this patterning is not present.
+https://www.nature.com/articles/19315
+%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
+%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning?
+Are all the same types of neurons there or might it be related to a missing cell type?
 
@@
 
+%TODO: number panels, in A middle, move legend so that text/lines do not class with guiding lines.
+Color for progenitor and neural crest are very similar and difficult to distinguish.
+I sometimes use this website to pick colors that are also colorblind friendly.
+https://colorbrewer2.org/#type=qualitative&scheme=Accent&n=4
+
 \afterpage{\clearpage}
@@
 The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance.
-Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in human.\par
+Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in humans.\par
 
@@
 are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
-DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
+DV patterning, early neuronal fate acquisition and neural crest tend to consist of clusters of 4-5 different
 TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
@@
 
-Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancer
-consist out of clusters of 4-5 different TFBS.
+Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancers
+consist of clusters of 4-5 different TFBS.
 Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain.
@@
   \item Conceptualization: S.D.W. \& S.A.
-  \item Experiments and sample preparation: S.D.W., V.C. \& C.M.
+  \item Experiments and sample preparation: S.D.W., V.C., E.S., L.H. \& C.M.
   \item Computational analysis: S.D.W. \& C.M.

### 2025-04-08 manuscript.tex
S.A. comments

@@
 enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
-transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the precise arrangement, copy
+transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the arrangement, copy
 number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}.
 Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}.
-For example, during immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, during human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}.
+For example, immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}.
 Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par
@@
 genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
-demonstrate how a single genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
+can be used to study how a genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
 takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
@@
 
-Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their primary counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
+Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their \textit{in vivo} counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
 However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.
-In this study, we will address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par
+In this study, we address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par
 
 During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
-consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, have different identities along the dorso-ventral (DV) axis of the embryo.
+consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, achieve different identities along the dorso-ventral (DV) axis of the embryo.
 These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}.
@@
 The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied.
-Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but a detailed descriptions of the enhancer-code of individual cell types are still limited.\par
+Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but detailed descriptions of the enhancer-code of individual cell types are still limited.\par
 
@@
 \subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
-We made use of human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
+We generated human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
 end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem
@@
 until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
-neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. In order to characterize the organoid-derived cell types to a greater level of detail at both
+neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. To characterize the organoid-derived cell types to a greater level of detail at both
 transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
@@
 neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
-Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and in the
-embryo only we could identify an endothelial, perivascular macrophage (PVM) / microglia (MIC) and otic placode
-population (\textbf{Fig. 1c}).\par
+Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and cell
+types missing from the organoid include endothelial cells, perivascular macrophages (PVM) / microglia (MIC) and otic placode (\textbf{Fig. 1c}).\par
 Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). Both
@@
 assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
-(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, we observed a gradient of anterior-to-posterior identities in the
-embryo (REF): from anterior to posterior expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig. 1g}).\par
+(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, in the embryo we observed a gradient of anterior-to-posterior identities (REF).
+Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig.
+1g}).\par
 Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
@@
 and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
-"$DeepNeuralTube_{organoid}$" and "$DeepNeuralTube_{embryo}$". The models use DNA sequence, of genomic regions, as input and
+"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". The models use DNA sequence of genomic regions as input and
 classify the regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class
@@
 (incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both the organoid cells (to
-train $DeepNeuralTube_{organoid}$) and embryo cells (to train $DeepNeuralTube_{embryo}$). To validate the ability of the models to
+train $DeepNeuralTube^o$) and embryo cells (to train $DeepNeuralTube^e$). To validate the ability of the models to
 predict cell type-specific enhancer \textit{activity}, we tested their performance on enhancers from the VISTA enhancer
@@
 facial mesenchyme (n=143). Neural tube enhancers could be classified with an area under the receiver-operator curve
-(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ (\textbf{Fig. 2b}; and
+(auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
 area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})), facial mesenchyme
-enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$
+enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$
 (\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). Of note, a fraction of neural tube
@@
 Might be good to point out why this overlap exists?
-Next, we wanted to assess the explainability of the models. Concretely we wanted to test whether the model learned known
+Next, we assessed the explainability of the models. Concretely we tested whether the model learned known
 TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic, to find
@@
 calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
-TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs used by the models. For both known motifs and \textit{de novo}
+TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs. For both known motifs and \textit{de novo}
 motifs we calculated pairwise similarities using TOMTOM (\textbf{REF}) followed by clustering, resulting in 56 motif
 clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
-(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
+(\textbf{Fig. 2e}) regions with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
 26 which is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). Most clusters (\textbf{Fig. 2f-g} and
 \textbf{Supplementary Table S2}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
-Thus, $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ use known-motifs as features to classify regions based
+Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
 on cell type-specific chromatin accessibility.\par
@@
 Both
-$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ explain this region similarly (Pearson correlation of x) and
+$DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) and
 highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). A version of this enhancer where 5\% of the
@@
     \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
-    $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
+    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
     using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
@@
     \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
-    for $DeepNeuralTube_{organoid}$ (top two panels) and $DeepNeuralTube_{embryo}$ (bottom two panels) and enhancer-reporter
+    for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
     assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
-    $DeepNeuralTube_{organoid}$ (top) and $DeepNeuralTube_{embryo}$ (bottom) of a genomic enhancer human neural crest
+    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic enhancer human neural crest
     enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
@@
 \subsection*{
-  Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors
+  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
 }
@@
 To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
-contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$.
-We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities.
+contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube^o$ and $DeepNeuralTube^e$.
+We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities.
 Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig.
 3e}).
-Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
+Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns
 underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
@@
 3h}).
-FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig.
+FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig.
 3h}).\par
@@
 TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
-peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
-that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
+peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. The floorplate peaks
+that overlap with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
 did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
@@
 FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
-bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
-the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
+bound by FOXA2 is higher than floorplate specific regions. Indeed, FOXA2 target regions that are specific to the floorplate mostly
 have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
-Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
+Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance tends
 to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
@@
 \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic.
-Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
-Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+Contribution scores for $DeepNeuralTube^o$ are not outlined.
+Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
 \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics.
@@
 For this purpose, we calculated the pairwise correlation between region-topic
-probabilities of all AP and DV topics (\textbf{Fig. 4b}). From this it is clear that both AP and DV patterning as well
+probabilities of all AP and DV topics (\textbf{Fig. 4b}). Both AP and DV patterning as well
 as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
@@
 
-$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
+$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
 together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
-explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
+explaining the predictions of $DeepNeuralTube^e$ on the regions of those domains we found four distinct
 homeodomain TF binding motifs, one for each domain (\textbf{Fig.
@@
 S4).
-In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs.
-The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig.
+In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs, correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig.
 4d-e} and Supplementary Fig.
-S4).
+S4)
 %TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
@@
 4f}).
-Indeed, $DeepNeuralTube_{embryo}$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig.
+Indeed, $DeepNeuralTube^e$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig.
 4f}).
@@
 
-From these results it seems that a relatively simplistic code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains.
+These results suggest that an unsophisticated code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains.
 Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig.
@@
 \textbf{c}, Heatmap showing region-topic probabilities for the top 3,000 regions per topic (y-axis) and each anterior-posterior topic (x-axis).
-\textbf{d}, Heatmap showing prediction scores of $DeepNeuralTube_{embryo}$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis).\textbf{e}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topics (x-axis).\textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31.
+\textbf{d}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis).\textbf{e}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topics (x-axis).\textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31.
 \textbf{g}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 61 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 31 (bottom); n = 200.
@@
 \subsection*{
-  Deciphering the enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
+  The enhancer code underlying pre-migratory and migratory neural crest and facial mesenchyme.
 }
@@
 crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
-$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
+$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube^e$ to explain the top 1,000 regions for each organoid and embryo
 topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
-$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances of
+$DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
 these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par
@@
 Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig.
-5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig.
+5b-c}); and moreover all other states of both the neuronal progenitors (\textbf{Fig.
 3e}) and neurons (\textbf{Fig.
@@
 instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
-instances co-occur and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
+instances co-occur, and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
 %TODO: C.M.: Mesenchyme+NC or calculated over all regions?
@@
 $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
-suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score).
+suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ and calculated the difference in prediction score across both alleles (delta prediction score).
 At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models.
@@
 S6}).
-As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}.
+As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}.
 The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig.
@@
 \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic.
-Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
-Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+Contribution scores for $DeepNeuralTube^o$ are not outlined.
+Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
 \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics.
@@
 \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
-\textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube_{organoid}$ and crosses for $DeepNeuralTube_{embryo}$) across the human genome.
+\textbf{f}, Manhattan plot showing -log p value of the SNP being associated with variations in facial shape\cite{whiteInsightsGeneticArchitecture2021} (positive y-axis) and the maximum (across classes of the model) of the absolute value of the difference in prediction score between the reference and alternative allele (negative y-axis; circles for $DeepNeuralTube^o$ and crosses for $DeepNeuralTube^e$) across the human genome.
 Lead SNPs are shown with a bigger dot size.
 \textbf{g}, Zoom in of chromosome one with the locus of SNP rs1555067 shown as an inset, the SNP itself is shown in red.
-\textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube_{organoid}$ of the references (top) and alternative (bottom) allele of SNP rs1555067.
+\textbf{h}, Nucleotide contribution scores for Topic 58 of $DeepNeuralTube^o$ of the references (top) and alternative (bottom) allele of SNP rs1555067.
 The location of the SNP is indicated with a dotted line.
@@
 \subsection*{
-  Deciphering the enhancer code underlying early neuronal differentiation.
+  The enhancer code underlying early neuronal differentiation.
 }
@@
 Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
-acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube_{organoid}$ and
-$DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each topic and generated a code table based on the
+acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
+$DeepNeuralTube^e$ to explain the top 1,000 regions for each topic and generated a code table based on the
 identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
@@
 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}.
-Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig.
+Additionally, in the embryo topics, instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig.
 6d} and Supplementary Fig.
@@
 \textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic.
-Contribution scores for $DeepNeuralTube_{organoid}$ are not outlined.
-Contribution scores for $DeepNeuralTube_{embryo}$ are outlined and negated.
+Contribution scores for $DeepNeuralTube^o$ are not outlined.
+Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
 \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neuron related topics.
@@
   (Continued)  \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
-for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
+for using $DeepNeuralTube^o$ (left) and $DeepNeuralTube^e$ (right). \textbf{c}, scatter plot showing
 location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
@@
 
-In order to create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes.
+To create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes.
 Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo.
-The fact that cell states were defined in an unbiased manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance.
+The fact that cell states were defined in a data driven manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance.
 Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in humans.\par
@@
 are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
-DV patterning, early neuronal fate acquisition and neural crest tend to consist of clusters of 4-5 different
+DV patterning, early neuronal fate acquisition and neural crest tend to consist of heterotypic clusters of 4-5 different
 TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
-within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
+within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs underlies
 the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
@@
 TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest.
-For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances.
-Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig.
+For all these TFs we found dimer motifs with a fixed number of nucleotides in between the two TFBS instances.
+Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing, with instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig.
 8}).
@@
 The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par
-
+%TODO: S.A.: But retain strongest footprint ?!
 Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancers
 consist of clusters of 4-5 different TFBS.
-Homeotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain.
+Homotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain.
 Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}.
@@
 In conclusion, we provide an enhancer-code atlas of early human neuronal and neural crest development. We confirm the
-paradigm that enhancers consist out of clusters of TFBS with soft syntactic rules but also provide a counter example to
+paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules but also provide a counter example to
 this paradigm.
 Furthermore, we validate the biological relevance of neural tube organoids.
-Finally, we provide $DeepNeuralTube$, two sequence-to-chromatin accessibility models, freely to the research community, which can be used to predict, interpret and design enhancers underlying early human neuronal and neural crest development.
+Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, freely to the research community, which can be used to predict, interpret and design enhancers underlying early human neuronal and neural crest development.
 

### 2025-04-08 manuscript.tex
Introduction rewrite.

@@
 
-\title{Deciphering enhancer rules underlying early human neuronal and neural crest development.}
+\title{An enhancer code compendium of early human neuronal and neural crest development}
 \date{}
@@
 
-Genomes of complex multi-cellular organisms encode all cell types/states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} of that organism in specific non-coding elements named
-enhancers\cite{davidsonRegulatoryGenomeGene2010}. The code words used by enhancers are
-transcription factor binding sites (TFBS)\cite{davidsonRegulatoryGenomeGene2010} and the arrangement, copy
-number and strength of individual TFBS define the enhancer and its cell type-specific activity\cite{dewinterModellingDesignTranscriptional2025}.
-Due to the soft syntax rules underlying the enhancer-code\cite{zeitlingerSevenMythsHow2020}, DNA sequence based deep learning methods have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025}.
-For example, immune cell differentiation\cite{maslovaDeepLearningImmune2020}, fly brain cell types\cite{janssensDecodingGeneRegulation2022}, liver cell types\cite{bravogonzalez-blasSinglecellSpatialMultiomics2024}, melanoma states\cite{minnoyeCrossspeciesAnalysisEnhancer2020, atakInterpretationAllelespecificChromatin2021}, human brain development\cite{mannensChromatinAccessibilityHuman2024}, mammalian and chicken brain cell types\cite{heckerEnhancerdrivenCellType2025} and pluripotent cell stem cells\cite{avsecBaseresolutionModelsTranscriptionfactor2021}.
-Furthermore, it has been demonstrated recently that the code learned by these models is sufficient to design novel enhancers with cell type-specific activity\cite{taskiranCelltypedirectedDesignSynthetic2024, dealmeidaTargetedDesignSynthetic2024, gosaiMachineguidedDesignCelltypetargeting2024, yinIterativeDeepLearningdesign2024}.\par
-
-Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the
-genome. Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
-can be used to study how a genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,
-takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}. Organoids can furthermore give insights into how genomes
-interpret both cell intrinsic and extrinsic signals. For example, epigentic
-modifications\cite{zenkSinglecellEpigenomicReconstruction2024}, transcription factor (TF)
-perturbations\cite{fleckInferringPerturbingCell2023} or morphogen
-signaling\cite{sanchis-callejaDecodingMorphogenPatterning2024}. Another benefit of organoids is that they can be used to
-model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. For
-example, early human development.
-One such system is early human development, the study of which is especially important for our understanding of human health and disease.
+Genomes of complex multi-cellular organisms specify all cell types and states\cite{trapnellDefiningCellTypes2015, arendtOriginEvolutionCell2016, fleckWhatCellType2023} through distinct non-coding elements named
+enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
+Central to the study of the genomic code of enhancers is the paradigm that 
+enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020}
+and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par
+
+DNA sequence based deep learning methods\cite{
+  alipanahiPredictingSequenceSpecificities2015,
+  zhouPredictingEffectsNoncoding2015,
+  kelleyBassetLearningRegulatory2016,
+  quangDanQHybridConvolutional2016,
+  movvaDecipheringRegulatoryDNA2019,
+  kelleySequentialRegulatoryActivity2018,
+  yuanScBassetSequencebasedModeling2022,
+  chenSequencebasedGlobalMap2022,
+  maslovaDeepLearningImmune2020,
+  minnoyeCrossspeciesAnalysisEnhancer2020,
+  avsecBaseresolutionModelsTranscriptionfactor2021,
+  avsecEffectiveGeneExpression2021,
+  dealmeidaDeepSTARRPredictsEnhancer2022,
+  linderPredictingRNAseqCoverage2025a,
+  kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
+and largely support the enhancer-code paradigm. 
+However, these studies are predominantly limited to a small number of cell type or cell states\cite{
+  maslovaDeepLearningImmune2020,
+  minnoyeCrossspeciesAnalysisEnhancer2020,
+  atakInterpretationAllelespecificChromatin2021,
+  avsecEffectiveGeneExpression2021,
+  bravogonzalez-blasSinglecellSpatialMultiomics2024}, and often focus on model organisms\cite{
+  janssensDecodingGeneRegulation2022,
+  heckerEnhancerdrivenCellType2025}, with only few studies profiling human development\cite{mannensChromatinAccessibilityHuman2024}.\par
+
+Organoids are a reductionistic biological model system that can be used to interrogate how cell types are encoded in the human genome. 
+They can be used to model biological systems that are otherwise - due to ethical and/or practical reasons - inaccessible for study. 
+Organoids derived from pluripotent stem cells are especially interesting in this context, given that they
+can be used to study how the human genome can generate multiple distinct cell types\cite{mccauleyPluripotentStemCellderived2017,takebeOrganoidsDesign2019, hoferEngineeringOrganoids2021}.\par
+
+%%% What is missing and bridge
+
+Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols 
+and how they compare to their \textit{in vivo} counterparts at a transcriptomic level\cite{
+  heIntegratedTranscriptomicCell2024,
+  xuIntegratedTranscriptomicCell2023}. 
+However, such comparisons at the level of the enhancer-code are still limited\cite{
+  zenkSinglecellEpigenomicReconstruction2024,
+  fleckInferringPerturbingCell2023}.
+
+In this study, we address these limitations by determining the enhancer-codes of cell types and cell states generated by human neural tube organoids, 
+a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}
+and compare these codes to those of of the head of a four weeks post conception human embryo.
+Through these comparisons we confirm the enhancer-code paradigm, 
+but also provide a counter examples specifically related to the enhancer-code underlying anterior-posterior patterning.
 \par
-
-%%% What is missing and bridge
-
-Current efforts are ongoing to determine the breadth of human cell types that can be generated using organoid protocols and how they compare to their \textit{in vivo} counterparts at a transcriptomic level in for example, brain organoids\cite{heIntegratedTranscriptomicCell2024} and endoderm-derived organoids\cite{xuIntegratedTranscriptomicCell2023}.
-However, such comparisons at the level of the enhancer-code are still limited\cite{zenkSinglecellEpigenomicReconstruction2024, fleckInferringPerturbingCell2023}.
-In this study, we address this gap by determining the enhancer-codes of cell types/states generated by human neural tube organoids, a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021} and compare these codes to those of cell types/states of the head of a four weeks post conception human embryo.\par
-
-During vertebrate development, both the neural tube and neural crest originate from the neural ectoderm. The neural tube
-consists of proliferating neuronal progenitors that, through opposing gradients of SHH and BMP/Wnt signaling, achieve different identities along the dorso-ventral (DV) axis of the embryo.
-These early identities later give rise to distinct neuronal subtypes of the central nervous system\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020}.
-Neural crest cells originate from the neural plate border, through intermediate signals of BMP and Wnt\cite{simoes-costaEstablishingNeuralCrest2015}, and migrate throughout the body giving rise to a multitude of cell types including the peripheral nervous system, craniofacial skeleton and melanocytes\cite{simoes-costaEstablishingNeuralCrest2015}.
-The gene regulatory networks underlying both DV patterning of neuronal progenitors\cite{sagnerEstablishingNeuronalDiversity2019, delasRepressiveInteractionsGene2020} and neural crest induction, specification, migration and differentiation\cite{simoes-costaEstablishingNeuralCrest2015, candido-ferreiraMultilayeredTranscriptionalControl2023} are already well studied.
-Furthermore, studies using chromatin accessibility profiling methods at the bulk and single cell level, profiling enhancers encoding cell types and states of neuronal progenitors and neural crest cells are starting to emerge\cite{delasDevelopmentalCellFate2023, zhangCisregulatoryLogicIntegrating2024, huEndothelialRegulatoryCircuits2025} but detailed descriptions of the enhancer-code of individual cell types are still limited.\par
 

### 2025-04-08 manuscript.tex
Intro: compendium

@@
 %%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
+
+%Context
+Enhancers encode cell types and cell states using 
+%need
+%task
+%results
+%conclusion
+%implication
 
@@
 
-In this study, we address these limitations by determining the enhancer-codes of cell types and cell states generated by human neural tube organoids, 
-a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}
-and compare these codes to those of of the head of a four weeks post conception human embryo.
+In this study, we address these limitations by building an enhancer-codes compendium of cell types and cell states generated by human neural tube organoids, 
+a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
+We compare this compendium to that of the head of a four weeks post conception human embryo.
 Through these comparisons we confirm the enhancer-code paradigm, 

### 2025-04-08 manuscript.tex
Summary and keywords.

@@
 %Context
-Enhancers encode cell types and cell states using 
+Enhancers encode cell types as clusters of transcription factor binding sites. 
 %need
-%task
-%results
+Studies of enhancer-codes are predominantly limited to a small number of cell types at a time and often focus
+on model organisms.
+%task and results
+In this study we generate a single-cell multiome atlas of differentiating human neural tube organoids 
+and of the head of a four weeks post conception human embryo. 
+Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
+are in common between the organoid model and the human embryo and we provide a compendium of enhancer-codes 
+for each individual cell type. 
+We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a single nucleosome depleted region. 
+Enhancers encoding anterior-posterior patterning, however, form a exception to this rule where they are encoded using
+homotypic clusters of homeodomain transcription factor binding sites.
 %conclusion
@@
 %%%  for some journals, keywords are chosen by the editors.
+
+Enhancer-code, human neural tube, human development, deep learning, sequence-to-function
 

### 2025-04-08 manuscript.tex
Go through results again.

@@
 \subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
-We generated human neural tube organoids (hNTOs) to model early human neuronal and neural crest development. To this
-end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem
-cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices. These were
-incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3
-until day 5 (\textbf{Fig. 1a}). These conditions were previously shown to generate multiple neuronal progenitor and
-neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. To characterize the organoid-derived cell types to a greater level of detail at both
-transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
-scRNA-seq) at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}),
-resulting in 22,862 high quality cells (\textbf{Fig. 1b}). Note, that similar transcriptomic and chromatin accessibility
-states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the
-organoid system.\par
-To evaluate the biological relevance of the cell types and states obtained from the organoid, we also performed 10x single
-cell multiome on the head of a single four weeks post conception human embryo, resulting in 25,785 high quality cells (
-\textbf{Fig. 1c}). We annotated both the organoid and embryo cells based on the expression of marker genes
-(\textbf{Supplementary Table S1}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,
-neural crest cells, facial mesenchyme cells and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
-Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) and cell
-types missing from the organoid include endothelial cells, perivascular macrophages (PVM) / microglia (MIC) and otic placode (\textbf{Fig. 1c}).\par
-Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). Both
-organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning as
-is evident from the expression of dorso-ventral marker TFs (REF): from ventral to dorsal \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig.
-1d}).
+
+We generated human neural tube organoids to model early human neuronal and neural crest development. 
+To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices.
+The organoids were incubated for 11 days and we supplemented the medium with retinoic acid (RA) and Smoothened agonist (SAG) from day 3 until day 5 (\textbf{Fig.
+1a}).
+These conditions were previously shown to generate multiple neuronal progenitor and
+neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
+To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
+scRNA-seq) 
+at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
+resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
+Note, that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par
+
+To evaluate the biological relevance of the cell types and states obtained from the organoid, 
+we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
+resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
+We annotated both the organoid and embryo cells based on the expression of marker genes
+(\textbf{Supplementary Table S1}). 
+In both systems we could identify neuronal progenitos, early differentiating neurons,
+neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
+Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
+while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par
+
+Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). 
+Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
+\textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
 In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
-A cluster that still expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction
-marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig. 1e}). Finally,
-focusing on the neural crest cells, in both the organoid and the embryo we obtained a population of pre-migratory neural
-crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial
-  mesenchyme cells (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression). In the embryo,
-  but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2}
-reflecting the otic placode (\textbf{Fig. 1f}).\par
-Given that we performed single-cell multiome on the entire head of the four weeks post conception embryo we also
-assessed the expression of anterior-posterior marker TFs. While the organoid cells had either an anterior forebrain
-(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, in the embryo we observed a gradient of anterior-to-posterior identities (REF).
+This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig.
+1e}).
+Finally, focusing on the neural crest, in both the organoid and the embryo we identified a population of pre-migratory neural crest (\textit{GRHL2} expression and absence of \textit{ZEB2} expression), migratory neural crest (\textit{SOX10} and \textit{ZEB2} expression), facial mesenchyme (\textit{TWIST1} expression) and early differentiation peripheral neurons (\textit{NEUROD1} expression).
+In the embryo, but not in the organoid, we observed an additional population of cells marked by the expression of \textit{SOX10} and \textit{PAX2} reflecting the otic placode (\textbf{Fig. 1f}).\par
+
+Next, we assessed the expression of anterior-posterior marker TFs. 
+While the organoid cells had either an anterior forebrain(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
+in the embryo we observed a gradient of anterior-to-posterior identities.
 Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig.
-1g}).\par
-Cell cycle arrest is an important hallmark of the start of neuronal differentiation (REF), for this reason we scored
-both organoid and embryo cells for the expression of marker genes of the G1, G2M and S phase of the cell cycle and
-quantified the fraction of cells in each phase stratified by the main cell types (\textbf{Fig. 1h}). From this it is
-clear that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the
-central nervous system are in cell cycle arrest, while the other cell types are still cycling (\textbf{Fig. 1h}).\par
-Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo-derived cell types
-and states on a chromatin accessibility level we identified sets of co-accessible regions in the organoid model using
-topic modeling. We focused on topics (i.e., sets of co-accessible regions) that were enriched for marker TF motifs and
-calculated the average expression of marker genes for each cell group where these regions were co-accessible, for both
-organoid and embryo cells. We observed similar patterns of marker gene expression for the same cell state (defined by
-the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.83
-across identical states (\textbf{Supplementary Fig. S1c-d}).\par
+1g})\cite{braunComprehensiveCellAtlas2023}.\par
+
+Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
+we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig.
+1h}), an important hallmark of neuronal differentiating\cite{
+braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par
+
+Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
+we identified sets of co-accessible regions in the organoid model using topic modeling. 
+Focusing on topics that were enriched for marker TF motifs we calculated the average expression of marker genes for each cell group where these regions were co-accessible for both organoid and embryo cells.
+We observed similar patterns of marker gene expression for the same cell state (defined by
+the topic) across the organoid and embryo (\textbf{Fig. 1i}) 
+with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par
 
@@
     For organoid: 
-    ARX [0, 1.5],
-    NKX2-2 [0, 1.5],
-    PAX6 [0, 1.5],
-    IRX3 [0, 1.0],
-    ZIC1 [0, 1.5],
-    GDF7 [0, 1.5],
-    SIX3 [0, 1.5],
-    EMX2 [0, 1.5],
-    FOXG1 [0, 0.8],
-    BARHL1 [0, 1.0],
-    EN1 [0, 1.5],
-    HOXB3 [0, 1.5],
-    SOX2 [0, 1.5],
-    ASCL1 [0, 1.5],
-    ISL1 [0, 1.5],
-    GAD1 [0, 0.8],
-    UNXC [0, 0.8],
-    GATA2 [0, 0.8],
-    GRHL2 [0, 1.5],
+    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
+    IRX3, BAHRL1: [0, 1.0], 
+    FOXG1, GAD1, UNXC, GATA2, SOX10: [0, 0.8] and 
     ZEB2 [0, 3.0], 
-    SOX10 [0, 0.8],
-    NEUROD1 [0, 1.5],
-    PAX2 [0, 1.5],
-    TWIST1 [0, 1.5]
     For embryo: 
-    ARX [0, 1.5],
-    NKX2-2 [0, 1.5],
-    PAX6 [0, 1.5],
-    IRX3 [0, 1.5],
-    ZIC1 [0, 1.5],
-    GDF7 [0, 1.5]
-    SIX3 [0, 1.5],
-    EMX2 [0, 1.5],
-    FOXG1 [0, 0.8],
-    BARHL1 [0, 1.5],
-    EN1 [0, 1.5],
-    HOXB3 [0, 1.5],
-    SOX2 [0, 1.5],
-    ASCL1 [0, 1.5],
-    ISL1 [0, 1.0]
-    GAD1 [0, 0.8],
-    UNXC [0, 0.8],
-    GATA2 [0, 0.8]
-    GRHL2 [0, 1.5],
-    ZEB2 [0, 3.0],
-    SOX10 [0, 0.8]
-    NEUROD1 [0, 1.5],
-    PAX2 [0, 1.5],
-    TWIST1 [0, 2.0]
+    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
+    FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
+    ISL1 [0, 1.0], 
+    ZEB2 [0, 3.0] and
+    TWIST1 [0, 2.0].
   }
@@
 
-To gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, neural crest-
-and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
-"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". The models use DNA sequence of genomic regions as input and
-classify the regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class
-fashion; \textbf(Fig. 2a)). The classes of the models are defined based on topic modeling. To get a fine-grained
-definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest
-(incl. facial mesenchyme and peripheral neurons) and early differentiating neurons for both the organoid cells (to
-train $DeepNeuralTube^o$) and embryo cells (to train $DeepNeuralTube^e$). To validate the ability of the models to
-predict cell type-specific enhancer \textit{activity}, we tested their performance on enhancers from the VISTA enhancer
-browser (\textbf{REF}). Relevant for this study, this database contains enhancers active in the neural tube (n=349) and
-facial mesenchyme (n=143). Neural tube enhancers could be classified with an area under the receiver-operator curve
+In order to gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, 
+neural crest- and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
+"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
+The models use DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf(Fig.
+2a)).
+The classes of the models are defined based on topic modeling. 
+To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl.
+facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par
+
+We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
+using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
+Enhancers active in the neural tube (n=349) could be classified with an area under the receiver-operator curve
 (auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
-area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})), facial mesenchyme
-enhancers with an auROC of 0.78 and 0.79 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$
-(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). Of note, a fraction of neural tube
-enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by both models.\par
-%TODO: C.M.: I think the original VISTA enhancers were annotated based on localization only, so it would make sense that premig.
-NC would be very hard to differentiate from normal neural tube ectoderm.
-Might be good to point out why this overlap exists?
-Next, we assessed the explainability of the models. Concretely we tested whether the model learned known
-TF binding motifs as features. For this, we performed motif enrichment analysis on all regions for each topic, to find
-known motifs underlying the accessibility of each region. We
-also
-calculated gradients, for each model, on the top 1,000 regions (as predicted by the model) for each topic followed by
-TF-MoDISco (\textbf{REF}) to find \textit{de novo} motifs. For both known motifs and \textit{de novo}
-motifs we calculated pairwise similarities using TOMTOM (\textbf{REF}) followed by clustering, resulting in 56 motif
-clusters (\textbf{Fig. 2d}). Of note, most clusters contain known-motifs enriched in both organoid regions and embryo
-(\textbf{Fig. 2e}) regions with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
-26 which is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). Most clusters (\textbf{Fig. 2f-g} and
-\textbf{Supplementary Table S2}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.
+area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
+Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
+using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
+and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
+Of note, a fraction of neural tube enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by
+both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
+that is is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
+is not surprising.\par
+
+Next, we assessed the explainability of the models. Concretely we tested if the model learned known
+TF binding motifs as features. To this end, we calculated pairwise similarities\cite{
+  guptaQuantifyingSimilarityMotifs2007} between \textit{de novo} motifs learned by the model, 
+and identified using TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018}, and known motifs that are enriched in cell type specific regions.
+Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
+Most clusters contain known motifs enriched in regions specific to both organoid and embryo cell types (\textbf{Fig.
+2e}), with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2})
+and cluster 26 that is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). 
+More importantly, Most clusters contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
+%TODO: add percentages
 Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
 on cell type-specific chromatin accessibility.\par
-As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig.
-S2c}).
-This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig.
-2h}).
-Both
-$DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) and
-highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). A version of this enhancer where 5\% of the
-nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both
-models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig. 2h}).\par
-Finally, to also investigate the performance of the models on migratory neural crest, a cell type for which there are no
-validated enhancers in the Vista enhancer browser, we selected a genomic region that is specifically accessible in the
-migratory neural crest cell type of the organoid (\textbf{Supplementary Fig. S2d}). Both models predict this region
-to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e}) and nucleotide importance scores
-reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). We tested the enhancer activity of this region
-using a reporter assay in chicken embryos. Indeed the enhancer was specifically active in migratory neural crest 24
-hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par
+
+As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
+This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
+Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) 
+%TODO: Add correlation
+and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
+A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig.
+2h}).\par
+
+Finally, to also investigate the performance of the models on migratory neural crest, 
+a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
+we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Supplementary Fig. S2d}). 
+Both models predict this region to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e})
+and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
+Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
+Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par
 
@@
 }
+
 Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
-identities in both the organoid and the embryo data (\textbf{Fig. 3a}). From ventral to dorsal this is topic 36, 38, 33,
-54 and 48 in the organoid data and topic 34, 38, 79, 88 and 58 in the embryo data. As an example we first investigated
-two enhancers of \textit{SHH} that are strongly conserved and specific to the floorplate [\textbf{REF}]; SFPE1 and SFPE2.
-\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. in the organoids and the embryo; \textbf{Fig. 3b}).
-SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells
-belonging to the ventral topic, in both the organoid and embryo data (\textbf{Fig. 3c}). Explaining the accessibility of both
-enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1
-and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig. 3d}).\par
-
-To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
-contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube^o$ and $DeepNeuralTube^e$.
-We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities.
-Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig.
-3e}).
-Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns
-underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
-models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
-and RFX instances is important (\textbf{Fig.
-3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig.
+identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
+From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
+topic 34, 38, 79, 88 and 58 for the embryo. 
+As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
+floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
+\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
+SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
+Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig.
+3d}).\par
+
+To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we clustered 
+and aligned \textit{de novo} motifs, identified in doros-ventral regions, based on pairwise motif similarities. 
+Next, we caclulated the average contribution for all identified motif instances per cluster and for each topic 
+resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e})
+Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
+importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3e}). 
+For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
+are important (\textbf{Fig.
+3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig.
 3g} and Supplementary Fig.
 S3).
-In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig.
-3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig.
-3e-g} and Supplementary Fig.
-S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig.
-3e-g} and Supplementary Fig.
-S3).
-For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig.
-3e-g} and Supplementary Fig.
-S3).
-Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig.
-3e} and Supplementary Fig.
-S3).
+In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
+PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and 
+ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
+For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
+SOX instances are important that correlates with the expression of\textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
+Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
 Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
 3h}).
-FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig.
-3h}).\par
-
-It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
-embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
-asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
-TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
-peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. The floorplate peaks
-that overlap with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
-did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
-affinity of each individual instance (making use of the FOX motif score calculated using
-FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
-bound by FOXA2 is higher than floorplate specific regions. Indeed, FOXA2 target regions that are specific to the floorplate mostly
-have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
-Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance tends
-to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
-have a single instance, we observe that FOX instances in regions that are general tend to be of the form
-\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
+FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par
+
+It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
+Indeed, when centering on FOX instances we observe a strong Tn5 cute-site footprint (\textbf{Fig. 3i}). 
+We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
+For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
+We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did no as floorplate specific.
+Quantifying both the number of FOX instances per region and
+the affinity of each individual instance (making use of the FOX motif score calculated using
+FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
+specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. 3j}). 
+Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance 
+tends to be higher (\textbf{Fig. 3k}). These results suggest that the overall affinity of general FOXA2 target regions 
+is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
+Generating a \textit{de novo} motif from general and specific FOXA2 target regions that have a single FOX instance 
+reveals that FOX instances in regions that are generally bound tend to be of the form \textit{TGTTT\underline{AC}} 
+while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
 
@@
 \subsection*{
-  Anterior-posterior patterning in neuronal progenitors is regulated by different homeodomain transcription factors that
-  each recognize a distinct binding site.
-}
-
-Given that we observed cell states from different AP domains in the neuronal progenitors from the embryo (but not
-organoid; see \textbf{Fig. 1}) we decided to study the gene regulatory code underlying this patterning further. We
-identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59,
-Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). We first asked if the AP-program is distinct
-from the DV program.
+  Anterior-posterior patterning in neuronal progenitors is regulated by different homeodomain transcription factors that each recognize a distinct binding site.
+}
+
+In the embryo we observed neuronal progenitors states of different anterior-posterior (AP) identities (\textbf{Fig. 1g}).
+We identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). 
+We first asked if the AP-program is distinct from the DV program. 
 In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? 
-For this purpose, we calculated the pairwise correlation between region-topic
-probabilities of all AP and DV topics (\textbf{Fig. 4b}). Both AP and DV patterning as well
-as each of their sub-domains are mostly encoded in distinct genomic regions, with the exception of AP Topic 59 and 31
-that correlate with the region-topic probability of DV Topic 38 (which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
+For this purpose, we calculated the pairwise correlation between region-topic probabilities of all AP and DV topics (\textbf{Fig. 4b}). 
+This revealed that both AP and DV patterning are mostly encoded in distinct genomic regions, 
+with the exception of AP Topic 59 and 31 that correlate with the region-topic probability of DV Topic 38 
+(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
 
 $DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
-together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
-explaining the predictions of $DeepNeuralTube^e$ on the regions of those domains we found four distinct
-homeodomain TF binding motifs, one for each domain (\textbf{Fig.
-4e}).
+together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). 
+By explaining the predictions of $DeepNeuralTube^e$ on the regions of those domains we found four distinct
+homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
 From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
 Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. 
-Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}.
-Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig.
-4d-e} and Supplementary Fig.
-S4).
-In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs, correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig.
-4d-e} and Supplementary Fig.
-S4)
+Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
+cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
+Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
+and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). 
+In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs, 
+correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
+with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
+topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4)\par
 %TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
@@
 Focusing on the two most anterior domains, two similar homeodomain TF binding motifs seem to underlie the differential
-accessibility of those domains. Both have the core homeobox motif: \textit{TAAT} but one is flanked by the nucleotides
-\textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. 
+accessibility of those domains. Both have the core homeobox motif "\textit{TAAT}" 
+but one is flanked by the nucleotides \textit{TAG} while the other is flanked by the nucleotides \textit{CCCT}. 
 %TODO: Do we know which TFs these correspond to? If you do, mention it in the text
 We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains.
-To test this hypothesis, we first did a substitution experiment where we replaced all instances of \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice versa (\textbf{Fig.
-4f}).
-Indeed, $DeepNeuralTube^e$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} to  \textit{CTAAT\underline{TAG}} substitution (\textbf{Fig.
-4f}).
-Next, we designed enhancers from scratch by embedding either the  \textit{CTAAT\underline{TAG}} or \textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig.
-4g}).
+To test this hypothesis, we first did a substitution experiment where we replaced all instances of
+\textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice
+versa (\textbf{Fig. 4f}). 
+Indeed, $DeepNeuralTube^e$ predicts an anterior-to-posterior shift of accessibility of those regions after the \textit{CTAAT\underline{TAG}} to \textit{CTAAT\underline{CCCT}} substitution and 
+vice versa a posterior-to-anterior shift of accessibility after the \textit{CTAAT\underline{CCCT}} to
+\textit{CTAAT\underline{TAG}} substitution (\textbf{Fig. 4f}). 
+Next, we designed enhancers from scratch by embedding either the \textit{CTAAT\underline{TAG}} or 
+\textit{CTAAT\underline{CCCT}} motif\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025} 
+and using respectively Topic 61 and Topic 31 as target class (\textbf{Fig. 4g}). 
 Embedding of two instances of either motif was sufficient to reach predicted chromatin accessibility levels of equivalent ranges to the predicted chromatin accessibility levels of genomic regions (\textbf{Fig.
@@
 Next, to eliminate potential model bias we repeated the same experiment but optimized for the opposite class. 
-This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61 by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig.
-4h}).
-In this case, the synthetic design failed to converge for the target class but instead regions with high prediction score for the AP domain that corresponds to the motif used for embedding were generated.
-This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig.
-4h}).\par
+This is, we optimized for Topic 31 by embedding the \textit{CTAAT\underline{TAG}} motif and we optimized for Topic 61
+by embedding the \textit{CTAAT\underline{CCCT}} motif (\textbf{Fig. 4h}). 
+In this case, the synthetic design failed to converge for the target class but instead regions with high prediction 
+score for the AP domain that corresponds to the motif used for embedding were generated. 
+This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 
+and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig.
+4h}), even though the prediction score was optimized towards the opposite class\par
 
 These results suggest that an unsophisticated code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains. 
-Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig.
-4i}).
-Surprisingly, this very simple classifier reached high performance with area under the ROC of 0.82 and area under the precision recall curve of 0.80 (\textbf{Fig.
-4i}).\par
+Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains 
+in either Topic 61 or Topic 31 using as features only the motif scores of the two motifs identified above (\textbf{Fig. 4i}). 
+Surprisingly, this very simple classifier reached high performance with area under the ROC of 0.82 and area under the precision recall curve of 0.80 (\textbf{Fig. 4i}).\par
 
@@
 crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
-$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube^e$ to explain the top 1,000 regions for each organoid and embryo
-topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
-$DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
+$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
+Again, both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
 these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par
 
-Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig.
-5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig.
-5d} and Supplementary Fig.
-S5).
-Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig.
-5b-c}); and moreover all other states of both the neuronal progenitors (\textbf{Fig.
-3e}) and neurons (\textbf{Fig.
-6b}).
+Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
+instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) 
+the importance of TEAD instances correlates with \textit{TEAD3/4} expression, 
+GRHL with \textit{GRHL1/1} 
+and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
+Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
+negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
+and moreover a negative contribution to all other cell states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
 The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
@@
 The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). 
-Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig.
-5b-c}).
+Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). 
 We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. 
-The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig.
-5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig.
-5d} and Supplementary Fig.
-S5).
+The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), 
+ZIC with \textit{ZIC1} to \textit{ZIC5} and 
+the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
 The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
-After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig.
-5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig.
-5d} and Supplementary Fig.
-S5).\par
-
-Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
-pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
-importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
-expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). The latter is consistent with a recent study where it has been
-shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
+After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) 
+for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par
+
+Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. 
+Instances of a FOX patterns and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). 
+The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
+that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
+The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
+bind this specific dimer motif regulating the expression of genes involved in
 facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par
@@
 
-Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
-that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
-enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
-$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
-suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ and calculated the difference in prediction score across both alleles (delta prediction score).
-At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models.
-Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig.
-5f}).
-Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig.
-S6}).
+Recently, a genome wide association study (GWAS) identified 203 single-nucleotide polymorphisms (SNPs)
+that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
+Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
+we asked if, using $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity.
+Therefore for all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$
+and calculated the difference in prediction score across both alleles (delta prediction score). 
+At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
+Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
+Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S6}). 
 As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
-The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig.
-5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state.
-This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig.
-5h}).\par
+The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
+(\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme. 
+This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
+homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par
 
@@
 acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
-$DeepNeuralTube^e$ to explain the top 1,000 regions for each topic and generated a code table based on the
-identified pattern instances (\textbf{Fig. 6b}). Both models found similar patterns and the importance of the instances
-of these patterns is highly correlated across both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par
+$DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
+Both models found similar patterns and the importance of the instances of these patterns is highly correlated across
+both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par
 
@@
 instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
-These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
-\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig.
-6d} and Supplementary Fig.
-S7).
-Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig.
-6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig.
-6d} and Supplementary Fig.
-S7).
-Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig.
-6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}.
-For subsequent topics SOX instances are not important anymore (\textbf{Fig.
-6b-c}) instead we find topics where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig.
-6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig.
-6d} and Supplementary Fig.
-S7) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig.
-6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}.
-Additionally, in the embryo topics, instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig.
-6d} and Supplementary Fig.
-S7).
-Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig.
-6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig.
-6d} and Supplementary Fig.
-S7).\par
+These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
+correlating with the expression of respectively \textit{TEAD1},
+\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
+correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor 
+patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) 
+the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
+For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) 
+instead we find topics where a combination of EBOX and EBF instances are important 
+(organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
+(\textbf{Fig. 6d} and Supplementary Fig. S7) 
+and topics where a combination of EBF and ONECUT instances are important 
+(organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), 
+the latter correlating with the expression of \textit{ONECUT1-3}. 
+Additionally, in the embryo topics, instances of NKX are important in these latter two topics, 
+the importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
+and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+Finally, in the organoids we could identify an additional topic for which instances of GATA are important
+(\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
 %TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
@@
 
-Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
-tube organoids and the head of a four weeks post conception human embryo with almost one-to-one correspondence based on TFBS importances and TF expression.
-Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo.
-To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig.
-3e}, \textbf{Fig.
-4e}, \textbf{Fig.
-5b} and \textbf{Fig.
-6b}).
-The overall performance of these models was acceptably high (\textbf{Fig.
-7a}) with an average area under the ROC of 0.77 and area under the precision-recall curve of 0.39.
-Notably, all models for neural crest related states performed best (\textbf{Fig.
-7a}).
-Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig.
-7a}).
+Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in 
+respectively the neural tube organoids and the head of a four weeks post conception human embryo with 
+almost one-to-one correspondence based on TFBS importances and TF expression. 
+Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors 
+in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
+logistic regression models to classify genomic regions as belonging to the topic representing each state or not,
+making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). 
+The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average auROC of 0.77 
+and average auPR of 0.39. 
+Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
+Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models 
+resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). 
 Finally, we used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. 
@@
 In case less than three distinct TFBS had a positive coefficient for a state, we implanted multiple copies of each TFBS to come to a total of at least four TFBS. 
-The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig.
-7a}) - and were cell type specific (\textbf{Fig.
-7b}) - with the prediction score for neural crest related states being again the highest.
-Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig.
-7a}) while other are more specific to certain states.
-\par
+The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - 
+%TODO: add average prediction score
+and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. 
+Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par
 
 To conclude, our results indicate that enhancers specific to DV-patterning, neural crest states and early neuronal differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
-7a}) in regions of on average 134 - 154 bp in size (\textbf{Fig.
-7c}), corresponding to the size of one nucleosome depleted region (NDR).
-AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig.
-7a}).
+7a})
+in regions of on average 134 - 154 bp in size (\textbf{Fig. 7c}), 
+corresponding to the size of one nucleosome depleted region (NDR). 
+AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig. 7a}).
 
@@
 
-We built an enhancer-code atlas of early human neuronal and neural crest development. Providing enhancer-codes of
-neuronal progenitor DV- and AP-patterning, early neuronal fate acquisition and neural crest induction, migration and
-facial mesenchyme specification.\par
-
-To create this atlas, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the enhancer-codes.
-Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between the organoid and embryo.
-The fact that cell states were defined in a data driven manner by making use of topic modeling and that each of these states were independently identified in two separate datasets increases the confidence of their biological relevance.
-Furthermore, this high similarity between the organoid and the embyro derived cell states validates neural tube organoids as a biological model for early human neural tube and neural crest development, biological cell states that are otherwise not easily accessible in humans.\par
+We built a compendium of enhancer-codes underlying early human neuronal and neural crest development. P
+roviding enhancer-codes of neuronal progenitor DV- and AP-patterning, 
+early neuronal fate acquisition and 
+neural crest induction, migration and facial mesenchyme specification.\par
+
+To create this compendium, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) 
+dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can 
+effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the 
+enhancer-codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between 
+the organoid and embryo. The fact that cell states were defined in a data driven manner by making use of topic 
+modeling and that each of these states were independently identified in two separate datasets increases the confidence 
+of their biological relevance. Furthermore, this high similarity between the organoid and the embyro derived cell 
+states validates neural tube organoids as a biological model for early human neural tube and neural crest development, 
+biological cell states that are otherwise not easily accessible in humans.\par
 
 The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
-compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
-are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
-DV patterning, early neuronal fate acquisition and neural crest tend to consist of heterotypic clusters of 4-5 different
-TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
-within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs underlies
-the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
-model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
-model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
-of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
-TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest.
+compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. 
+Nonetheless, there are recurrent patterns that are shared among enhancers of different cell types. 
+In particular, enhancers underlying DV patterning, early neuronal fate acquisition and neural crest 
+tend to consist of heterotypic clusters of 4-5 different TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located within one NDR (+/-147bp). 
+That is to say, cooperative binding of multiple TFs underlies the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. 
+Here, we see most evidence for a model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer model\cite{arnostiTranscriptionalEnhancersIntelligent2005}).
+However, for multiple cell types we also observed evidence of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. 
+This was the case for TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} 
+and GRHL1/2 dimers, 
+TFAP2A and TFAP2B/C heterodimers\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} 
+and SOX10 homodimers in neural crest. 
 For all these TFs we found dimer motifs with a fixed number of nucleotides in between the two TFBS instances. 
-Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing, with instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig.
-8}).
-Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}.
+Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing,
+with instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig. 8}). 
+Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
+subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. 
 This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
-Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}.
-Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity.
+Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather 
+dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. 
+Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites 
+and/or binding sites of lower affinity. 
 The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par
 %TODO: S.A.: But retain strongest footprint ?!
-Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule that enhancers
-consist of clusters of 4-5 different TFBS.
+Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule 
+that enhancers consist of clusters of 4-5 different TFBS. 
 Homotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. 
 Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. 
-It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}.
+It has been a long standing question how homeodomain TFs achieve specific binding\cite{mannChapter3Hox2009, hubertHoxGenesDevelopment2023} 
+given that all homeodomain TFs seem to bind the same \textit{TAAT} motif\cite{bergerVariationHomeodomainDNA2008, noyesAnalysisHomeodomainSpecificities2008}. 
 It has been hypothesized that specificity is mainly achieved through cooperative binding with other TFs\cite{mannChapter3Hox2009}. 
 Indeed, we observe this for EN1/2 and HOXB3 that are predicted to co-bind with PAX5/8 and MEIS1/2 or PBX1/3 respectively forming specific dimer motifs.
-In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for respectively LHX2/9, RAX or SIX3/6 and OTX1/2, EMX2, DMBX1 or LMX1A that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites.
+In contrast, we suggest a novel mode where specificity is achieved through the homeodomain itself for respectively LHX2/9, RAX or SIX3/6 and OTX1/2, EMX2, DMBX1 or LMX1A 
+that preferentially bind \textit{TAATTA} and \textit{TAATCC} sites. 
 Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. 
-In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} that could be important for cell type specific binding.\par
-
-In conclusion, we provide an enhancer-code atlas of early human neuronal and neural crest development. We confirm the
-paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules but also provide a counter example to
-this paradigm.
+In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} 
+and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} 
+there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} 
+that could be important for cell type specific binding.\par
+
+In conclusion, we provide an enhancer-code compendium of early human neuronal and neural crest development. 
+We confirm the paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules 
+but also provide a counter example to this paradigm. 
 Furthermore, we validate the biological relevance of neural tube organoids. 
-Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, freely to the research community, which can be used to predict, interpret and design enhancers underlying early human neuronal and neural crest development.
+Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, 
+freely available to the research community, 
+which can be used to predict, 
+interpret and design enhancers underlying early human neuronal and neural crest development.
 
@@
 
-In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. The effect a
-TF has on target gene expression depends on its effector domain and interactions with
+In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. 
+The effect a TF has on target gene expression depends on its effector domain and interactions with
 co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
-the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. Large scales studies, in a variety of cell
-states, are now being performed to map TFs to co-factors and downstream effects on gene
-expression\cite{tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
+the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. 
+Large scales studies, in a variety of cell states, are now being performed to map TFs to co-factors and downstream effects on gene expression\cite{
+  tyckoHighThroughputDiscoveryCharacterization2020, alerasoolIdentificationFunctionalCharacterization2022,
 sotoCompendiumHumanTranscription2022, delrossoLargescaleMappingMutagenesis2023,
 mukundHighthroughputFunctionalCharacterization2023, bellComparativeCofactorScreens2024,
-tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. These studies may especially be
-important to infer repressive roles of TFs on gene expression, an effect that is currently difficult to study with state
-of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
-dewinterModellingDesignTranscriptional2025}. In particular for the neural tube, it is known that repressive interactions
-in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. In this study we could only predict a strong
-repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
+tyckoDevelopmentCompactTranscriptional2024, doughtySinglemoleculeStatesLink2024}. 
+These studies may especially be important to infer repressive roles of TFs on gene expression, 
+an effect that is currently difficult to study with state of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
+dewinterModellingDesignTranscriptional2025}. 
+In particular for the neural tube, it is known that repressive interactions in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. 
+In this study we could only predict a strong repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
 systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
@@
 
-Furthermore, the expression of a gene is in general regulated by multiple enhancers in the gene's locus\cite{perryShadowEnhancersFoster2010,
+Furthermore, the expression of a gene is in general regulated by multiple enhancers in the gene's locus\cite{
+  perryShadowEnhancersFoster2010,
 martinez-araSystematicAnalysisIntrinsic2022, bergmanCompatibilityRulesHuman2022, zuinNonlinearControlTranscription2022,
 gschwindEncyclopediaEnhancergeneRegulatory2023, martinez-araLargescaleAnalysisIntegration2024,
-loubiereDevelopmentalHousekeepingTranscriptional2024, thomasEnhancerCooperativityCan2025}. In this study we focused on
-the enhancer code of individual enhancers but not how the signal of multiple enhancers is integrated to control target
-gene expression. Currently efforts are ongoing to build large scale models for modeling gene expression making use of the
+loubiereDevelopmentalHousekeepingTranscriptional2024, thomasEnhancerCooperativityCan2025}.
+In this study we focused on the enhancer code of individual enhancers 
+but not how the signal of multiple enhancers is integrated to control target gene expression. 
+Currently efforts are ongoing to build large scale models for modeling gene expression making use of the
 code of all enhancers in the gene's locus\cite{avsecEffectiveGeneExpression2021,linderPredictingRNAseqCoverage2025a,
@@
 
-I%f you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
+%f you are publishing in a STAR Methods journal, please refer to the appropriate guide for authors:
 %\begin{itemize}
@@
 the subset of cells annotated as neuron (Neuron cells and Neuronal progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
-Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
+Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/datmccallumMALLETMachineLearning2002618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
 Topic modeling was performed separately for all cells and for each subset of cells.
@@
 generated for each sample using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function create\_fragments\_matrix\_from\_fragments, this matrix was binarized (\textgreater 0), merged in one count matrix using the Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) function sparse.hstack and subsequently split in a matrix for each of neuronal progenitors, neurons and neural crest cells and saved as matrix market format.
-A mallet corpus file was generated for each binary matrix using the pycisTopic command pycistopic topic\_modeling mallet create\_corpus and finally topic modeling was run using Mallet\cite{mccallumMALLETMachineLearning} (v202108; RRID:SCR\_026708) using the pycisTopic command: pycistopic topic\_modeling mallet run using 300 iterations and default parameters otherwise.
+A mallet corpus file was generated for each binary matrix using the pycisTopic command pycistopic topic\_modeling mallet create\_corpus and finally topic modeling was run using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; RRID:SCR\_026708) using the pycisTopic command: pycistopic topic\_modeling mallet run using 300 iterations and default parameters otherwise.
 Model evaluation was performed using the pycisTopic command pycistopic topic\_modeling mallet stats that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.

### 2025-04-08 manuscript.tex
M&M V.C.

@@
 Pluripotent stem cells were cultured in flat bottom, TC-treated 6 well plates (Avantor; 734-2323). For culture details see
-Methods details below. All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven  Number
-%TODO: add ECD
+Methods details below. 
+All human iPSC and ESC experiments were approved by the Ethical Committee of the KU Leuven (ECD S63316)
 . Human neural tube organoids were generated based on Fattah \textit{et al.} 2021 Nature communications\cite{
@@
 Immediately after the TrypLE reaction was inhibited by adding \SI{2}{\milli\liter} of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)).
-Cells were pelleted at 300 RCF for \SI{3}{\minute} and after counting on a LUNA-FL™ Dual Fluorescence Cell Counter resuspended to \SI{3.5e6}{cells\per\milli\litre} in $GM_0$ ( basal GM (N-2 Supplement (ThermoFisher, 17502048) 1x, B-27 Supplement (ThermoFisher, 17504044) 1x, GlutaMax Supplement (ThermoFisher, 35050061) 1x, Sodium Pyruvate (ThermoFisher, 11360070) \SI{1}{\milli\molar}, MEM Non-Essential Amino Acids Solution (ThermoFisher, 11140050) 1x, Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}, Neurobasal Medium (ThermoFisher, 21103049) \SI{50}{\percent} V/V and DMEM/F12 (ThermoFisher, 11320033) \SI{50}{\percent} V/V) with RHO/ROCK inhibitor (Y-27632 (Dihydrochloride), Stemcell technologies, 100-1044) \SI{1}{\micro\molar}).
+Cells were pelleted at 300 RCF for \SI{3}{\minute} and after counting on a LUNA-FL Dual Fluorescence Cell Counter resuspended to \SI{3.5e6}{cells\per\milli\litre} in $GM_0$ ( basal GM (N-2 Supplement (ThermoFisher, 17502048) 1x, B-27 Supplement (ThermoFisher, 17504044) 1x, GlutaMax Supplement (ThermoFisher, 35050061) 1x, Sodium Pyruvate (ThermoFisher, 11360070) \SI{1}{\milli\molar}, MEM Non-Essential Amino Acids Solution (ThermoFisher, 11140050) 1x, Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram\per\milli\litre}, Neurobasal Medium (ThermoFisher, 21103049) \SI{50}{\percent} V/V and DMEM/F12 (ThermoFisher, 11320033) \SI{50}{\percent} V/V) with RHO/ROCK inhibitor (Y-27632 (Dihydrochloride), Stemcell technologies, 100-1044) \SI{1}{\micro\molar}).
 Next, the polyethylene glycol A (PEG-A) hydrogel matrix (see Methods detail below) was prepared by first adding Corning Mouse Laminin (Avantor, 734-1098) \SI{0.1}{\milli\gram\per\milli\liter} followed by activated factor 13 (see Methods details below) in that order to a PEG mixture (see Methods details below; PEG \SI{1.5}{\percent} or \SI{2}{\percent} V/V, buffer (TRIS \SI{0.05}{\milli\molar}, $CaCl_2$ \SI{0.05}{\milli\molar}; pH 7.6; filter sterilized) in $GM_0$ medium).
@@
 \subsubsection*{Chicken embryos}
-Fertilized chicken eggs, obtained from ... 
-%TODO: add vendor
-were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4.
+Fertilized chicken eggs, obtained from "Kwekerij van het Hallerbos" were incubated at \SI{37}{\degreeCelsius} at \SI{70}{\percent} humidity for 30 hours to reach Humburger Hamilton (HH) stage 4.
 Afterwards, embryos were removed from the egg and kept in culture for an additional 24 hours.
@@
 Immediately after the TrypLE reaction was inhibited by adding twice the volume of DMEM/F12 FBS (Fetal Bovine Serum (FBS; ThermoFisher, A5256701) \SI{20}{\percent} V/V; Penicillin-Streptomycin (ThermoFisher, 15140122) \SI{50}{\micro\gram} in DMEM/F12 (ThermoFisher, 11320033)).
-The single-cell suspension was pelleted at 500 RCF for 4 minutes, resuspended in DMEM/F12 FBS and counted on LUNA-FL™ Dual Fluorescence Cell Counter.
+The single-cell suspension was pelleted at 500 RCF for 4 minutes, resuspended in DMEM/F12 FBS and counted on LUNA-FL Dual Fluorescence Cell Counter.
 Immediately after the cells were processed for 10x single-cell multiome.
@@
 
+\paragraph*{10x multiome nuclei isolation}
+To isolate nuclei from the single cells of the dissociated neural tube organoids, protocol CG000365 (10x Genomics) 
+was followed. 
+For each developmental time point, 500k cells were resuspended in 100 µl nuclei lysis buffer 
+(\SI{10}{\milli\molar} Tris-HCl pH 7.4; 
+\SI{10}{\milli\molar} NaCl; 
+\SI{3}{\milli\molar} $MgCl_2$; 
+\SI{0.1}{\percent} Tween-20; 
+\SI{0.1}{\percent} NP40; 
+\SI{0.01}{\percent} Digitonin;
+\SI{1}{\percent} BSA; 
+\SI{1}{\milli\molar} DTT and 
+\SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor 
+in nuclease-free water
+). 
+After 5 min incubation on ice, \SI{1}{\milli\litre} chilled wash buffer was added to the lysed cells 
+(
+\SI{10}{\milli\molar} Tris-HCl pH 7.4; 
+\SI{10}{\milli\molar} NaCl; 
+\SI{3}{\milli\molar} $MgCl_2$; 
+\SI{0.1}{\percent} Tween-20; 
+\SI{1}{\percent} BSA; 
+\SI{1}{\milli\molar} DTT and 
+\SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor in 
+nuclease-free water
+). 
+Nuclei were pelleted by centrifugation at 500 g for 5 min at \SI{4}{\degreeCelsius} and 
+resuspended in a 1x nuclei buffer (10x Genomics) supplemented with 
+\SI{1}{\milli\molar} dithiothreitol and \SI{1}{\units\per\micro\litre} RNasin ribonuclease inhibitor. 
+Nuclei suspensions were passed through a \SI{40}{\micro\meter} Flowmi filter (VWR Bel-Art SP Scienceware). 
+Nuclei concentration was assessed using the LUNA-FL Dual Fluorescence Cell Counter.
+
 \paragraph*{10x multiome library preparation}
-%TODO: Valerie
-
-\paragraph*{10x multiome library sequencing}
-%TODO: Valerie
+Single-cell libraries were generated using the 10X Chromium Single-Cell Instrument and NextGEM Single Cell Multiome ATAC+Gene Expression kit (10X Genomics) according to the manufacturer’s protocol. 
+In brief, single nuclei of neural tube organoids were incubated for 60 min at \SI{37}{\degreeCelsius} 
+with a transposase that fragments the DNA in open regions of the chromatin and adds adapter sequences to the 
+ends of the DNA fragments. 
+After generation of nanoliter-scale gel bead-in-emulsions (GEMs), 
+GEMs were incubated in a C1000 Touch Thermal Cycler (Bio-Rad) under the following program: 
+\SI{37}{\degreeCelsius} for 45 min, \SI{25}{\degree} for 30 min, and hold at \SI{4}{\degreeCelsius}. 
+Incubation of the GEMs produced 10x barcoded DNA from the transposed DNA (for ATAC) and 10x barcoded, 
+full-length cDNA from poly-adenylated mRNA (for GEX). 
+This was followed by a quenching step that stopped the reaction. 
+After quenching, single-cell droplets were broken and the transposed DNA and full-length cDNA were isolated using
+Cleanup Mix containing Silane Dynabeads. 
+To fill gaps and generate sufficient mass for library construction, the transposed DNA and cDNA were amplified via PCR: 
+\SI{72}{\degreeCelsius} for 5 min; 
+\SI{98}{\degreeCelsius} for 3 min; 
+seven cycles of 
+\SI{98}{\degreeCelsius} for 20 s, 
+\SI{63}{\degreeCelsius} for 30 s, 
+\SI{72}{\degreeCelsius} for 1 min; 
+\SI{72}{\degreeCelsius} for 1 min; 
+and hold at \SI{4}{\degreeCelsius}. 
+The pre-amplified product was used as input for both ATAC library construction and cDNA amplification for gene
+expression library construction. 
+Illumina P7 sequence and a sample index were added to the single-strand DNA during ATAC library construction via PCR:
+\SI{98}{\degreeCelsius} for 45 s;
+7–9 cycles of \SI{98}{\degreeCelsius} for 20 s, 
+\SI{67}{\degreeCelsius} for 30 s, 
+\SI{72}{\degreeCelsius} for 20 s;
+\SI{72}{\degreeCelsius} for 1 min; 
+and hold at \SI{4}{\degreeCelsius}. 
+The sequencing-ready ATAC library was cleaned up with SPRIselect beads (Beckman Coulter). 
+Barcoded, full-length pre-amplified cDNA was further amplified via PCR: 
+\SI{98}{\degreeCelsius} for 3 min; 
+6–9 cycles of 
+\SI{98}{\degreeCelsius} for 15 s, 
+\SI{63}{\degreeCelsius} for 20 s, 
+\SI{72}{\degreeCelsius} for 1 min; 
+\SI{72}{\degreeCelsius} for 1 min; 
+and hold at \SI{4}{\degreeCelsius}. 
+Subsequently, the amplified cDNA was fragmented, end-repaired, A-tailed, and index adaptor ligated, 
+with SPRIselect cleanup in between steps. 
+The final gene expression library was amplified by PCR: 
+\SI{98}{\degreeCelsius} for 45 s; 
+5–16 cycles of 
+\SI{98}{\degreeCelsius} for 20 s,
+\SI{54}{\degreeCelsius} for 30 s, 
+\SI{72}{\degreeCelsius} for 20 s, 
+\SI{72}{\degreeCelsius} for 1 min; 
+and hold at \SI{4}{\degreeCelsius}. 
+The sequencing-ready GEX library was cleaned up with SPRIselect beads.
+
+\paragraph*{10x multiome sequencing}
+Before sequencing, the fragment size of every library was analyzed using the Bioanalyzer high-sensitivity chip. 
+All 10x single cell Multiome ATAC libraries were sequenced on NextSeq2000 instrument (Illumina) or on NovaSeq6000
+instrument (Illumina) 
+with the following sequencing parameters: 
+51 bp read 1, 
+8 bp index 1, 
+24 bp index 2, 
+51 bp read 2. 
+All 10x single cell Multiome Gene Expression libraries were sequenced on NextSeq2000 instrument (Illumina) or on NovaSeq6000 instrument (Illumina) 
+with the following sequencing parameters: 
+28 bp read 1, 
+10 bp index 1, 
+10 bp index 2, 
+90 bp read 2.
 

### 2025-04-08 manuscript.tex
Typo

@@
 
-We built a compendium of enhancer-codes underlying early human neuronal and neural crest development. P
-roviding enhancer-codes of neuronal progenitor DV- and AP-patterning, 
+We built a compendium of enhancer-codes underlying early human neuronal and neural crest development. 
+Providing enhancer-codes of neuronal progenitor DV- and AP-patterning, 
 early neuronal fate acquisition and 

### 2025-04-09 manuscript.tex
Changes throughout text.

@@
 \affil[3]{KU Leuven Department of Human Genetics, Leuven, Belgium}
-\affil[4]{Karolinska Insitute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
-\affil[5]{Karolinska Institutet, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}
+\affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
+\affil[5]{Karolinska Institute, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}
 
@@
 %need
-Studies of enhancer-codes are predominantly limited to a small number of cell types at a time and often focus
+Enhancer code studies are predominantly limited to a small number of cell types often focusing
 on model organisms.
 %task and results
-In this study we generate a single-cell multiome atlas of differentiating human neural tube organoids 
+In this study we generate a single-cell multiome atlas of human neural tube organoids 
 and of the head of a four weeks post conception human embryo. 
 Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
-are in common between the organoid model and the human embryo and we provide a compendium of enhancer-codes 
+are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
 for each individual cell type. 
-We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a single nucleosome depleted region. 
-Enhancers encoding anterior-posterior patterning, however, form a exception to this rule where they are encoded using
-homotypic clusters of homeodomain transcription factor binding sites.
+We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
+Enhancers encoding anterior-posterior patterning form an exception to this rule where they are encoded using
+homotypic clusters of homeodomain binding sites.
 %conclusion
+This compendium of enhancer codes thus provides gene regulatory insight into human development.
 %implication
@@
 enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
-Central to the study of the genomic code of enhancers is the paradigm that 
+Central to the study of the genomic code is the paradigm that 
 enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020}
@@
   kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
-and largely support the enhancer-code paradigm. 
+and largely support the enhancer code paradigm. 
 However, these studies are predominantly limited to a small number of cell type or cell states\cite{
@@
   xuIntegratedTranscriptomicCell2023}. 
-However, such comparisons at the level of the enhancer-code are still limited\cite{
+However, such comparisons at the level of the enhancer code are still limited\cite{
   zenkSinglecellEpigenomicReconstruction2024,
@@
 
-In this study, we address these limitations by building an enhancer-codes compendium of cell types and cell states generated by human neural tube organoids, 
+In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
 a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
 We compare this compendium to that of the head of a four weeks post conception human embryo.
-Through these comparisons we confirm the enhancer-code paradigm, 
-but also provide a counter examples specifically related to the enhancer-code underlying anterior-posterior patterning.
+Through these comparisons we confirm the enhancer code paradigm, 
+but also provide a counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
 \par
@@
 (\textbf{Supplementary Table S1}). 
-In both systems we could identify neuronal progenitos, early differentiating neurons,
+In both systems we could identify neuronal progenitors, early differentiating neurons,
 neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
@@
 
-Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cell types (\textbf{Fig. 1d-f}). 
+Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
 Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
@@
 
-Next, we assessed the expression of anterior-posterior marker TFs. 
-While the organoid cells had either an anterior forebrain(\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
-in the embryo we observed a gradient of anterior-to-posterior identities.
-Marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig.
+Next, we assessed the expression of anterior-posterior (AP) marker TFs. 
+While the organoid cells had either an anterior forebrain (\textit{SIX3}) or anterior spinal cord (\textit{HOXB3}) identity, 
+in the embryo we observed a gradient of anterior-to-posterior identities marked by the expression of \textit{SIX3}, \textit{EMX2}, \textit{FOXG1}, \textit{BARHL1}, \textit{EN1 }and \textit{HOXB3} (\textbf{Fig.
 1g})\cite{braunComprehensiveCellAtlas2023}.\par
@@
 Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
-we identified sets of co-accessible regions in the organoid model using topic modeling. 
-Focusing on topics that were enriched for marker TF motifs we calculated the average expression of marker genes for each cell group where these regions were co-accessible for both organoid and embryo cells.
-We observed similar patterns of marker gene expression for the same cell state (defined by
-the topic) across the organoid and embryo (\textbf{Fig. 1i}) 
+we identified sets of co-accessible regions in the organoid model using topic modeling 
+and identified equivalent cells in the embryo where these same regions were also co-accessible.
+We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
 with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par
@@
     \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
-    \textbf{a}, Schematic. \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
-    (\textbf{c}) cells colored by cell type identify. \textbf{d-f}, UMAP based on gene expression profiles of organoid (top)
-    and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons
-    (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
+    \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
+    \textbf{b-c}, UMAP based on gene expression profiles of organoid (\textbf{b}) and embryo
+    (\textbf{c}) cells colored by cell type identify. 
     (Continues next page)
@@
   \caption{
-    (Continued) \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
+    (Continued)
+    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
+    \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
     axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
-    color scale. \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
-    cell cycle phase identity and quantification of percentage of each cell for each cell cycle phase stratified by cell
-    type in the organoid (bottom left) and embryo (bottom right). \textbf{i}, Average gene expression level in
-    $log_{2}(CPM)$
-    (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
+    color scale. 
+    \textbf{h}, UMAP based on gene expression profiles in organoid (top left) and embryo (top right) colored by
+    cell cycle phase identity and quantification of percentage of each cell cycle phase stratified by cell
+    type in the organoid (bottom left) and embryo (bottom right). 
+    \textbf{i}, Average gene expression level in $log_{2}(CPM)$ (dot color) and fraction of cells expressing the gene (dot size) per cell state, defined by topic modeling, for organoid
     cells (top) and embryo cells (middle) and motif enrichment level as normalized enrichment score (NES; dot color) and
@@
     For organoid: 
-    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
+    Minimum is 0 and maximum is 1.5 for all genes except:
     IRX3, BAHRL1: [0, 1.0], 
@@
     For embryo: 
-    Minimum is 0 and maximum is 1.5 Log(CPM) for all genes except:
+    Minimum is 0 and maximum is 1.5 for all genes except:
     FOXG1, GAD1, UNCX, GATA2, SOX10 [0, 0.8], 
@@
 
-In order to gain insight into the sequence characteristics underlying cell type-specific activity of neural tube-, 
-neural crest- and derived cell type enhancers, we trained two sequence-to-function model convolutional neural networks:
+In order to gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
+neuronal progenitor, neural crest and neuronal cell type-specific enhancers, 
+we trained two sequence-to-function model convolutional neural networks:
 "$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
-The models use DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf(Fig.
-2a)).
+The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig.
+2a}).
 The classes of the models are defined based on topic modeling. 
@@
 and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
-Of note, a fraction of neural tube enhancers (24 \%) are predicted to be specific to pre-migratory neural crest by
+Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
 both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
-that is is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
+that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
 is not surprising.\par
@@
 Clustering this pairwise similarity matrix resulted in 56 clusters (\textbf{Fig. 2d}). 
-Most clusters contain known motifs enriched in regions specific to both organoid and embryo cell types (\textbf{Fig.
-2e}), with the exception of cluster 13 that is organoid specific (AP1 motif; \textbf{Supplementary Table S2})
-and cluster 26 that is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). 
-More importantly, Most clusters contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
-%TODO: add percentages
+All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
+organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
+More importantly, Most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
 Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
@@
 This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
-Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of x) 
-%TODO: Add correlation
+Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ explain this region similarly (Pearson correlation of 0.79) 
 and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
@@
     \textbf{DeepNeuralTube: an accurate sequence-to-function model revealing neuro-developmental enhancer code.} \\
-    \textbf{a}, Schematic. \textbf{b} receiver-operator curves for identifying neural tube (left) and facial mesenchyme the
-    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. \textbf{c-e}, UMAP of TomTom [REF] motif-to-motif similarities
-    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
-    colored by: cluster identity (\textbf{c}), model of origin (Continues next page)
+    \textbf{a}, Precision recall curves for $DeepNeuralTube^o$ and $DeepNeuralTube^e$ colored by main class and 
+    average area under (AUC) the precision recall curves per class (left) and schematic representation of the model
+    (right)
+    (Continues next page)
   }
@@
   \caption{
-    (Continued) (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
+    (Continued) 
+    \textbf{b}, receiver-operator curves for classifying neural tube (left) and facial mesenchyme enhancers using
+    $DeepNeuralTube^o$ and $DeepNeuralTube^e$. 
+    \textbf{c-e}, UMAP of TomTom\cite{guptaQuantifyingSimilarityMotifs2007} motif-to-motif similarities
+    using both \textit{de novo} motifs from DeepNeuralTube and known enriched motifs. Each dot is a motif. Motifs are
+    colored by: cluster identity (\textbf{c}), model of origin 
+    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
     \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
-    \textit{de novo} (right). \textbf{g}, contribution score for reference and mutant version of Vista enhancer hs111
+    \textit{de novo} (right). 
+    \textbf{g}, contribution score for reference and mutant version of VISTA enhancer hs111
     for $DeepNeuralTube^o$ (top two panels) and $DeepNeuralTube^e$ (bottom two panels) and enhancer-reporter
-    assay for reference and mutant version of hs111 (right two panels) [REF]. \textbf{h}, contribution score for
-    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic enhancer human neural crest
+    assay for reference and mutant version of hs111 (right two panels)\cite{
+    viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}. 
+    \textbf{h}, contribution score for
+    $DeepNeuralTube^o$ (top) and $DeepNeuralTube^e$ (bottom) of a genomic human neural crest
     enhancer (chr1:161200798-161201298; hg38) and enhancer-reporter assay in chicken embryo (HH4; right). Green is enhancer signal, pink is
@@
 To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we clustered 
-and aligned \textit{de novo} motifs, identified in doros-ventral regions, based on pairwise motif similarities. 
+and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
 Next, we caclulated the average contribution for all identified motif instances per cluster and for each topic 
-resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e})
+resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}).
 Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
@@
 For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
-SOX instances are important that correlates with the expression of\textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
+SOX instances are important correlating with the expression of\textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
 Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
-Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
+Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
 3h}).
@@
 For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
-We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did no as floorplate specific.
+We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did not as floorplate specific.
 Quantifying both the number of FOX instances per region and
@@
 while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
+%TODO: Revisit this analysis but intersect with floor plate topic instead of FOX hits!
 
@@
     \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. 
-\textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
-    and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells). (Continues next page).
+    (Continues next page).
   }
@@
   \caption{
-    (Continued) \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38.
-SFPE2 and SFPE1 is highlighted.
+    (Continued) 
+    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
+      and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
+    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
     \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
@@
 Signal is centered on the FOXA2 instance start position.
-\textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq signal (general; bottom) and those regions that don't overlap (specific; top).
-\textbf{k}, Boxplot (ADD NUMEBRS) showing $-log_10(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
+    \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq peaks (general; bottom) and those regions that do not overlap (specific; top).
+    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 236) and specific (n = 3114) regions, for organoid.
+    \textbf{k}, Boxplot showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
+    Two-sided p-values calculated using Mann-Whitney U test between distribution of motif scores for 
+    general regions with one FOX instance ($n_{organoid} = 47$ and $n_{embryo} = 42$) 
+    and specific regions with one FOX instance ($n_{organoid} = 1,508$ and $n_{embryo} = 1,467$).
     \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
@@
 \subsection*{
-  Anterior-posterior patterning in neuronal progenitors is regulated by different homeodomain transcription factors that each recognize a distinct binding site.
-}
-
-In the embryo we observed neuronal progenitors states of different anterior-posterior (AP) identities (\textbf{Fig. 1g}).
-We identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). 
+  Homotypic clusters of distinct homeodomain transcription factor binding sites encodes anterior-posterior patterning
+}
+
+In the embryo we identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig.
+4a}).
 We first asked if the AP-program is distinct from the DV program. 
@@
 $DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
-together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). 
-By explaining the predictions of $DeepNeuralTube^e$ on the regions of those domains we found four distinct
-homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
+together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
+for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
 From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
@@
   \caption{
-    \textbf{A simple enhancer code underlies anterior-posterior patterning in neuronal progenitors.} \\
+    \textbf{An unsophisticated enhancer code underlies anterior-posterior patterning in neuronal progenitors.} \\
   \textbf{a}, UMAP of embryo neuronal progenitors colored by cell-topic probabilities in RGB color scales for
 anterior-posterior topics. 
-\textbf{b}, Heatmap showing pairwise pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics.
-\textbf{c}, Heatmap showing region-topic probabilities for the top 3,000 regions per topic (y-axis) and each anterior-posterior topic (x-axis).
-\textbf{d}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis).\textbf{e}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topics (x-axis).\textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31.
+  \textbf{b}, Heatmap showing pairwise Pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics. 
+  \textbf{c}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis). 
+  \textbf{d}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topic (x-axis). 
+  \textbf{e}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in embryo
+  anterior-posterior topics. 
+  \textbf{f}, Boxplot showing prediction score before (wt) and after (alt) substituting all instances of \textit{CTAATTAG} with \textit{CTAATCCCT} (top) or substituting all instances of \textit{CTAATCCCT} with \textit{CTAATTAG} (bottom) in the topic 3,000 regions of either Topic 61 or Topic 31.
   \textbf{g}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 61 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 31 (bottom); n = 200.
   \textbf{h}, Boxplot showing prediction scores for all anterior-posterior topics after embedding 0 to 5 instances of \textit{CTAATTAG} while optimizing their location for Topic 31 (top) and after embedding 0 to 5 instances of \textit{CTAATCCCT} while optimzing their location for Topic 61 (bottom); n = 200.
-.\textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
+  \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
+  U.: UMAP
   }
@@
 crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
-$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
+$DeepNeuralTube^o$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
 Again, both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns and the importance of the instances of
@@
 The location of the SNP is indicated with a dotted line.
+  U: UMAP.
   }
@@
 $DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
-Both models found similar patterns and the importance of the instances of these patterns is highly correlated across
+Both models found similar patterns and the importance of the instances of these patterns is correlated across
 both models (0.47 average Pearson correlation coefficient; \textbf{Fig. 6b}).\par
@@
 For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
-instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
+instances of patterns from neuronal progenitors are important (\textbf{Fig. 6b-c}).
 These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
@@
 the latter correlating with the expression of \textit{ONECUT1-3}. 
-Additionally, in the embryo topics, instances of NKX are important in these latter two topics, 
-the importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
+Additionally, in the embryo we found topics where instances of NKX are important. 
+The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
 and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
@@
   \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neuron related topics. 
-\textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neuon related topics.
+  \textbf{d}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) for neuron related topics. 
   \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
@@
 \subsection* {
-  Four to five distinct transcription factor binding sites in a region of around 140 base pairs is sufficient to encode
-  all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
+  Four to five distinct transcription factor binding sites in a single nucleosome depleted region are sufficient to encode
+  all neuronal progenitor, neuronal and neural crest cell states.
 }
@@
 and average auPR of 0.39. 
-Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
+Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}).\par
+
 Next, for each cell state, we selected TFBS that had a positive coefficient for at least one of the logistic regression models 
 resulting in a total of twenty one TFBS for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). 
-Finally, we used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. 
-We used at least three TFBS in the design process for each state. 
-In case less than three distinct TFBS had a positive coefficient for a state, we implanted multiple copies of each TFBS to come to a total of at least four TFBS. 
-The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - 
-%TODO: add average prediction score
+We used the identified TFBS to design synthetic enhancers for each state using motif embedding\cite{taskiranCelltypedirectedDesignSynthetic2024, kempynckCREstedModelingGenomic2025}. 
+The resulting synthetic DNA sequences had an overall high prediction score (average 0.75 and 0.23 standard deviation) (\textbf{Fig. 7a}) - 
 and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. 
@@
 corresponding to the size of one nucleosome depleted region (NDR). 
-AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig. 7a}).
+AP-patterning enhancers seem to be an exception to this rule, where they are encoded using multiple instances of only two and sometimes even a single TFBS (\textbf{Fig. 7a}).
 
@@
 represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
-each topic (bottom). (Continues next page)}
+each topic (bottom). 
+(Continues next page)}
 \end{figure}
@@
   \caption{
-  (Continued)  \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
-for using $DeepNeuralTube^o$ (left) and $DeepNeuralTube^e$ (right). \textbf{c}, scatter plot showing
+  (Continued) 
+  \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
+ using $DeepNeuralTube^o$ (left) and $DeepNeuralTube^e$ (right). 
+  \textbf{c}, scatter plot showing
 location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
@@
 
-We built a compendium of enhancer-codes underlying early human neuronal and neural crest development. 
-Providing enhancer-codes of neuronal progenitor DV- and AP-patterning, 
+We built a compendium of enhancer codes underlying early human neuronal and neural crest development. 
+Providing enhancer codes of neuronal progenitor DV- and AP-patterning, 
 early neuronal fate acquisition and 
@@
 effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the 
-enhancer-codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between 
+enhancer codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between 
 the organoid and embryo. The fact that cell states were defined in a data driven manner by making use of topic 
@@
 For all these TFs we found dimer motifs with a fixed number of nucleotides in between the two TFBS instances. 
-Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing,
-with instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig. 8}). 
+Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bp spacing,
+with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 8}). 
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
@@
 and/or binding sites of lower affinity. 
-The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par
+The use of suboptimal TFBS is a reoccurring feature of cell type-specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par
 %TODO: S.A.: But retain strongest footprint ?!
@@
 
-In conclusion, we provide an enhancer-code compendium of early human neuronal and neural crest development. 
+In conclusion, we provide an enhancer code compendium of early human neuronal and neural crest development. 
 We confirm the paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules 
@@
 precision measures and the method used to determine significance can be found in the figure legends.
+For all boxplots: the top/lower hinge represents the upper/lower quartile and whiskers extend from the hinge to the 
+larges/smallest value no further than 1.5 times the interquartile range from the hinge, respectively. 
+The median is used as the center.
 %TODO: make sure it is there.
@@
     \textbf{Linking motif to TF in neurons}.\\
-    \textbf{a-k} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
-    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
-    that is annotated to (Continues next page):
+    (Continues next page):
  }
@@
     (Continued)
+  \textbf{a-k} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
+    topic (x-axis) and maximum expression (Max exp; y-axis) across topics and tau index using dot size for organoid (left) and embryo (right) for TFs with DNA binding domain
+    that is annotated to    
     TEA (\textbf{a}), 

### 2025-04-14 manuscript.tex
changes S.A.; changes C.M.

@@
 
-\title{An enhancer code compendium of early human neuronal and neural crest development}
+\title{An enhancer code compendium of early human neural and neural crest development}
 \date{}
@@
 \author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
-\author[5, \orcidlink{0000-0003-2931-8015}]{Erik Sundström}
+\author[5, 6, 7, \orcidlink{0000-0002-2628-7834}]{Ruth M. Williams}
+\author[5, 6, \orcidlink{0000-0001-9289-0263}]{Tatjana Sauka-Spengler}
 \author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
@@
 \affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
-\affil[5]{Karolinska Institute, Division of Neurodegeneration, Department of Neurobiology, Care Sciences and Society,Stockholm, Sweden}
+\affil[5]{Stowers Institute for Medical Research, Kansas City, Missouri, United States of America}
+\affil[6]{University of Oxford, MRC Weatherall Institute of Molecular Medicine, Radcliffe Department of Medicine, Oxford, United Kingdom}
+\affil[7]{University of Manchester, Faculty of Biology, Medicine and Health, Manchester, United Kingdom}
 
@@
 %need
-Enhancer code studies are predominantly limited to a small number of cell types often focusing
-on model organisms.
+Enhancer code studies have been limited to specific cell types in model organisms.
 %task and results
@@
 and of the head of a four weeks post conception human embryo. 
-Using DNA sequence based deep learning - DeepNeuralTube - we identify a total of 15 cell types that 
+Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
 are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
@@
 We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
-Enhancers encoding anterior-posterior patterning form an exception to this rule where they are encoded using
+Enhancers encoding anterior-posterior patterning form an exception to this rule, being encoded as
 homotypic clusters of homeodomain binding sites.
+% TODO: S.A. [some more results? Validation examples? Genetic variants?]
 %conclusion
-This compendium of enhancer codes thus provides gene regulatory insight into human development.
+This compendium of enhancer codes provides gene regulatory insight into human development.
 %implication
@@
 enhancers\cite{davidsonRegulatoryGenomeGene2010}. 
-Central to the study of the genomic code is the paradigm that 
-enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020}
+Central to the study of the genomic regulatory code is the paradigm that 
+enhancers consist of heterotypic clusters of transcription factor binding sites (TFBS)\cite{longEverChangingLandscapesTranscriptional2016,kimDecipheringMultiscaleQuantitative2023} with soft syntactic rules\cite{zeitlingerSevenMythsHow2020},
 and that cooperative binding of multiple transcription factors (TFs)\cite{lambertHumanTranscriptionFactors2018} underlies cell type-specific activity of the enhancer.\par
@@
   kempynckCREstedModelingGenomic2025} have shown major advances in the ability to model and explain enhancers\cite{zeitlingerSevenMythsHow2020,dewinterModellingDesignTranscriptional2025} 
-and largely support the enhancer code paradigm. 
-However, these studies are predominantly limited to a small number of cell type or cell states\cite{
+and largely support this enhancer code paradigm. 
+However, these studies have been predominantly limited to a small number of cell types or cell states\cite{
   maslovaDeepLearningImmune2020,
@@
 In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
-a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
+a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
 We compare this compendium to that of the head of a four weeks post conception human embryo.
-Through these comparisons we confirm the enhancer code paradigm, 
-but also provide a counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
+Through these comparisons we assess the enhancer code paradigm, 
+and also provide counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
 \par
@@
 
-\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
-
-We generated human neural tube organoids to model early human neuronal and neural crest development. 
+\subsection*{Human neural tube organoids recapitulate early neural and neural crest development}
+
+We generated human neural tube organoids to model early human neural and neural crest development. 
 To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices.
@@
 1a}).
-These conditions were previously shown to generate multiple neuronal progenitor and
+These conditions were previously shown to generate multiple neural progenitor and
 neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
@@
 resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
-Note, that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par
+Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par
 
 To evaluate the biological relevance of the cell types and states obtained from the organoid, 
-we also performed 10x single cell multiome on the head of a single four weeks post conception human embryo, 
+we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
 resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
@@
 (\textbf{Supplementary Table S1}). 
-In both systems we could identify neuronal progenitors, early differentiating neurons,
+In both systems we could identify neural progenitors, early differentiating neurons,
 neural crest, facial mesenchyme and early differentiating peripheral neurons (\textbf{Fig. 1b-c}).
 Additionally, in the organoid only we could identify a pluripotent stem cell population (\textbf{Fig. 1b}) 
-while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing (\textbf{Fig. 1c}).\par
-
-Next, we sub-clustered the neuronal progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
-Both organoid and embryo neuronal progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
+while perivascular macrophages (PVM) / microglia (MIC) and otic placode cells were missing in the organoid system (\textbf{Fig. 1c}).\par
+
+Next, we sub-clustered the neural progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
+Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
 \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
-In both the organoid and embryo system we obtained multiple early neuronal differentiation states.
-This includes a cluster that expresses the neuronal progenitor marker \textit{SOX2}, cells that express the early neuronal induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig.
+In both the organoid and embryo system we obtained multiple early neural differentiation states.
+This includes a cluster that expresses the neural progenitor marker \textit{SOX2}, cells that express the early neural induction marker \textit{ASCL1} and distinct clusters marked by the expression of \textit{GAD1}, \textit{UNCX} and \textit{GATA2} (\textbf{Fig.
 1e}).
@@
 we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig.
-1h}), an important hallmark of neuronal differentiating\cite{
+1h}), an important hallmark of neural differentiating\cite{
 braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par
 
-Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo on a chromatin accessibility level 
+Finally, to assess whether the organoid-derived cell types and states are also similar to the embryo at the level of chromatin accessibility
 we identified sets of co-accessible regions in the organoid model using topic modeling 
-and identified equivalent cells in the embryo where these same regions were also co-accessible.
+and identified equivalent cells in the embryo where the same regions were also co-accessible.
 We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
-with an average Pearson correlation coefficient of 0.83 across identical states (\textbf{Supplementary Fig. S1c-d}).\par
+with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Supplementary Fig. S1c-d}).\par
 
@@
   \caption{
-    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
+    \textbf{Human neural tube organoids recapitulate early neural and neural crest development.} \\
     \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
@@
     (Continued)
-    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neuronal progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
+    \textbf{d-f}, UMAP based on gene expression profiles of organoid (top) and embryo (bottom) cells types showing sub-clusters of neural progenitors (\textbf{d}), early differentiating neurons (\textbf{e}) and neural crest (\textbf{f}) along with expression of indicated genes in red-green-blue color scale.
     \textbf{g}, UMAP based on gene expression profiles of organoid (top; UMAP axis 1 and 2 shown) and embryo (bottom; UMAP
-    axis 2 and 3 shown) for neuronal progenitor subcluster along with the expression of indicated genes in red-green-blue
+    axis 2 and 3 shown) for neural progenitor subcluster along with the expression of indicated genes in red-green-blue
     color scale. 
@@
 \subsection*{
-  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
+  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neural and neural crest 
   development
@@
 
-In order to gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
-neuronal progenitor, neural crest and neuronal cell type-specific enhancers, 
+To gain insight into the sequence characteristics underlying cell type-specific chromatin accessibility of 
+neural progenitor, neural crest and neural cell type-specific enhancers, 
 we trained two sequence-to-function model convolutional neural networks:
-"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$". 
+"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$", on the organoid and embryo atlases respectively. 
 The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig.
 2a}).
-The classes of the models are defined based on topic modeling. 
-To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neuronal progenitors, neural crest (incl.
-facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) embryo (to train $DeepNeuralTube^e$).\par
-
-We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific activity 
+The classes of the models are defined as the topics from topic modeling. 
+To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl.
+facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par
+
+We tested the performance of $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to predict cell type-specific enhancer activity 
 using data from the VISTA enhancer browser\cite{viselVISTAEnhancerBrowsera2007, kosickiVISTAEnhancerBrowser2025}.
@@
 organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
-More importantly, Most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
+More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
 Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
@@
 and highlight the importance of multiple SOX TFBS (\textbf{Fig. 2h}). 
-A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a sinlge SOX TFBS (\textbf{Fig.
+A version of this enhancer where 5\% of the nucleotides are changed (hs111.1) does not have neural tube specific enhancer activity and both models predict that this loss of activity is due to the loss of a single SOX TFBS (\textbf{Fig.
 2h}).\par
@@
 \subsection*{
-  The enhancer code underlying dorso-ventral patterning of neuronal progenitors
-}
-
-Focusing on neuronal progenitors, we identified topics representing cells and regions of different dorso-ventral
+  The enhancer code underlying dorso-ventral patterning of neural progenitors
+}
+
+Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
 identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
@@
 
-To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we clustered 
+To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
 and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
-Next, we caclulated the average contribution for all identified motif instances per cluster and for each topic 
+Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
 resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}).
@@
 S3).
-In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
+In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
 PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and 
@@
 For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
-SOX instances are important correlating with the expression of\textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
+SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
 Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
@@
 It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
-Indeed, when centering on FOX instances we observe a strong Tn5 cute-site footprint (\textbf{Fig. 3i}). 
+Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. 3i}). 
 We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
@@
 specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. 3j}). 
-Furthermore, in case there is a single FOX instance in a general FOXA2 target region the affinity of this instance 
-tends to be higher (\textbf{Fig. 3k}). These results suggest that the overall affinity of general FOXA2 target regions 
+Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
+specific regions (\textbf{Fig. 3k}). These results suggest that the overall affinity of general FOXA2 target regions 
 is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
@@
   \caption{
-    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neuronal progenitors.} \\
-    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neuronal progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. 
+    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neural progenitors.} \\
+    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neural progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. 
     (Continues next page).
@@
 Signal is centered on the FOXA2 instance start position.
-    \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIp-seq peaks (general; bottom) and those regions that do not overlap (specific; top).
-    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 236) and specific (n = 3114) regions, for organoid.
-    \textbf{k}, Boxplot showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions (bottom) and specific regions (top) stratified by the number of FOXA2 instance per region.
-    Two-sided p-values calculated using Mann-Whitney U test between distribution of motif scores for 
-    general regions with one FOX instance ($n_{organoid} = 47$ and $n_{embryo} = 42$) 
-    and specific regions with one FOX instance ($n_{organoid} = 1,508$ and $n_{embryo} = 1,467$).
-    \textbf{l}, \textit{de novo} motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
+    \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIP-seq peaks (general; bottom) and those regions that do not overlap (specific; top).
+    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 2,130) and specific (n = 4,529) regions, for organoid.
+    \textbf{k}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions.
+* indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
+    \textbf{l}, de novo motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
  }
@@
 \subsection*{
-  Homotypic clusters of distinct homeodomain transcription factor binding sites encodes anterior-posterior patterning
+  Homotypic clusters of distinct homeodomain transcription factor binding sites encode anterior-posterior patterning
 }
@@
 $DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
+% TODO: S.A.: How does the organoid model score all these AP topics?
 together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
@@
 From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
-Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. 
-Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
+Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
+Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
 cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
@@
 and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). 
-In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs, 
+In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
 correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
@@
 We were surprised that this slight difference in the two motifs would be enough for encoding two separate AP domains.
-To test this hypothesis, we first did a substitution experiment where we replaced all instances of
+To test this hypothesis, we first performed a substitution experiment where we replaced all instances of
 \textit{CTAAT\underline{TAG}} with \textit{CTAAT\underline{CCCT}} in regions of the most anterior domain and vice
@@
 score for the AP domain that corresponds to the motif used for embedding were generated. 
-This is embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 
+In other words, embedding of the \textit{CTAAT\underline{TAG}} motif still generated regions with high prediction score for Topic 61 
 and embedding of the \textit{CTAAT\underline{CCCT}} motif generated regions with high prediction score for Topic 31 (\textbf{Fig.
-4h}), even though the prediction score was optimized towards the opposite class\par
-
-These results suggest that an unsophisticated code using two distinct motifs is sufficient to distinguish between the two utmost neuronal progenitor AP domains. 
+4h}), even though the prediction score was optimized towards the opposite class.\par
+
+These results suggest that an unsophisticated, homotypic code using two distinct motifs is sufficient to distinguish between the two utmost neural progenitor AP domains. 
 Encouraged by this finding, we trained a logistic regression model to classify genomic regions of those two domains 
@@
   \caption{
-    \textbf{An unsophisticated enhancer code underlies anterior-posterior patterning in neuronal progenitors.} \\
-  \textbf{a}, UMAP of embryo neuronal progenitors colored by cell-topic probabilities in RGB color scales for
+    \textbf{An unsophisticated enhancer code underlies anterior-posterior patterning in neural progenitors.} \\
+  \textbf{a}, UMAP of embryo neural progenitors colored by cell-topic probabilities in RGB color scales for
 anterior-posterior topics. 
@@
 Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
-crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
+crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neural progenitors, we used
 $DeepNeuralTube^o$ and $DeepNeuralTube^e$ to generate a neural crest code-table (\textbf{Fig. 5b}). 
@@
 negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
-and moreover a negative contribution to all other cell states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
+and moreover a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
 The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
@@
 index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
-method we found three sets of regions that are predicted to contain instances of different set of TFs. One set where
-instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear receptor, AP2 and ZIC
+method we found three sets of regions that are predicted to contain instances of different sets of TFs. One set where
+instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear Receptor, AP2 and ZIC
 instances co-occur, and one set where instances of FOX and TWIST1/homeodomain co-occur.\par
@@
 we asked if, using $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity.
-Therefore for all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$
-and calculated the difference in prediction score across both alleles (delta prediction score). 
+For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$
+and calculated the difference in prediction score between both alleles (delta prediction score). 
 At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
@@
 This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
-homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par
+homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimer (\textbf{Fig. 5h}).\par
 
@@
 \subsection*{
-  The enhancer code underlying early neuronal differentiation.
-}
-
-Focusing on early neuronal differentiation, we identified several topics for different stages of neuronal fate
+  The enhancer code underlying early neural differentiation.
+}
+
+During early neural differentiation, we identified several topics for different stages of neural fate
 acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
@@
 For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
-instances of patterns from neuronal progenitors are important (\textbf{Fig. 6b-c}).
+instances of patterns from neural progenitors are important (\textbf{Fig. 6b-c}).
 These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
@@
 correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
-Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor 
-patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) 
-the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
-For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) 
+Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
+patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}). 
+The importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
+For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}); 
 instead we find topics where a combination of EBOX and EBF instances are important 
@@
 and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
-Finally, in the organoids we could identify an additional topic for which instances of GATA are important
-(\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
+Finally, in the organoids we could identify an additional topic with instances of GATA
+(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
 %TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
@@
 Next, we analyzed the co-occurence of the identified patterns by calculating the Jaccard index of regions based on the presence of instances of each pattern.
-We find that instances of neuronal progenitor patterns tend to co-occur the most (\textbf{Fig.
+We find that instances of neural progenitor patterns tend to co-occur the most (\textbf{Fig.
 6e}), in line with our results in figure 3.\par
@@
   \captionsetup{justification=raggedright}
-  \caption{ \textbf{Deciphering the enhancer code of early neuronal differentiation.} \\
+  \caption{ \textbf{Deciphering the enhancer code of early neural differentiation.} \\
   \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics. 
@@
 \subsection* {
-  Four to five distinct transcription factor binding sites in a single nucleosome depleted region are sufficient to encode
-  all neuronal progenitor, neuronal and neural crest cell states.
+  Extracting general rules from the enhancer compendium
 }
@@
 almost one-to-one correspondence based on TFBS importances and TF expression. 
-Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors 
+Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neural progenitors 
 in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
 logistic regression models to classify genomic regions as belonging to the topic representing each state or not,
-making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). 
+making use of the motif scores of the identified patterns from each code table (\textbf{Fig.
+3e}, \textbf{Fig.
+4e}, \textbf{Fig.
+5b} and \textbf{Fig.
+6b}).
+We used regions of the topic of interest as foreground and regions of the other topics of the same cell type-class as background.
 The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average auROC of 0.77 
 and average auPR of 0.39. 
-Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}).\par
+Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
+However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S8). \par
 
@@
 
-To conclude, our results indicate that enhancers specific to DV-patterning, neural crest states and early neuronal differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
+These results indicate that enhancers specific to DV-patterning, neural crest states and early neural differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
 7a})
@@
 corresponding to the size of one nucleosome depleted region (NDR). 
-AP-patterning enhancers seem to be an exception to this rule, where they are encoded using multiple instances of only two and sometimes even a single TFBS (\textbf{Fig. 7a}).
+AP-patterning enhancers seem to be an exception to this rule, as they are encoded by multiple instances of only two and sometimes even a single TFBS (\textbf{Fig. 7a}).
 
@@
 represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
-each topic (bottom). 
+each topic (bottom). Cell type-classes are indicated by colors. 
 (Continues next page)}
@@
 
-We built a compendium of enhancer codes underlying early human neuronal and neural crest development. 
-Providing enhancer codes of neuronal progenitor DV- and AP-patterning, 
-early neuronal fate acquisition and 
+We built a compendium of enhancer codes underlying early human neural and neural crest development. 
+Providing enhancer codes of neural progenitor DV- and AP-patterning, 
+early neural fate acquisition and 
 neural crest induction, migration and facial mesenchyme specification.\par
@@
 Nonetheless, there are recurrent patterns that are shared among enhancers of different cell types. 
-In particular, enhancers underlying DV patterning, early neuronal fate acquisition and neural crest 
+In particular, enhancers underlying DV patterning, early neural fate acquisition and neural crest 
 tend to consist of heterotypic clusters of 4-5 different TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located within one NDR (+/-147bp). 
@@
 Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bp spacing,
-with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 8}). 
+with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 9}). 
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
@@
 %TODO: S.A.: But retain strongest footprint ?!
-Our results indicate that enhancers underlying neuronal progenitor AP-patterning form an exception to the rule 
+Our results indicate that enhancers underlying neural progenitor AP-patterning form an exception to the rule 
 that enhancers consist of clusters of 4-5 different TFBS. 
@@
 
-In conclusion, we provide an enhancer code compendium of early human neuronal and neural crest development. 
+In conclusion, we provide an enhancer code compendium of early human neural and neural crest development. 
 We confirm the paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules 
@@
 which can be used to predict, 
-interpret and design enhancers underlying early human neuronal and neural crest development.
+interpret and design enhancers underlying early human neural and neural crest development.
 
@@
 systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
-(\textbf{Supplementary Fig. S9}).\par
+(\textbf{Supplementary Fig. S10}).\par
 
@@
 \paragraph*{10x multiome library preparation}
-%TODO :Camiel
+Tissue was gently minced using a razor blade and cells were dissociated using a papain dissociation kit (Worthington)
+with \SI{200}{\units\per\milli\litre} DNAse at \SI{37}{\degreeCelsius} for 10 min following the manufacturer’s provided protocol.
+The suspension was triturated using a glass pipette to dissolve remaining clumps before being filtered 
+through a \SI{30}{\micro\meter} filter (CellTrics). 
+After dissociation, the cells were washed with 1X EBSS and concentrated (200g, 5 min). 
+1 x $10^6$ cells were pelleted (500g, 5 min) in a 2 ml LoBind Eppendorf tube. 
+Nuclei were then isolated from the pellet by incubating them with \SI{100}{\micro\litre}
+dissociation mix 
+(
+  \SI{0.001}{\percent} digitonin, 
+  \SI{0.01}{\percent} NP40, 
+  \SI{1}{\milli\molar} dithiothreitol, 
+  \SI{1}{\units\per\micro\litre} RNAse inhibitor, 
+  \SI{0.1}{\percent} Tween-20, 
+  \SI{1}{\percent} BSA, 
+  \SI{10}{\milli\molar} Tris-HCl, 
+  \SI{10}{\milli\molar} NaCl, 
+  \SI{3}{\milli\molar} $MgCl_2$
+) 
+for 5 min before adding 1 ml wash buffer 
+  (
+  \SI{1}{\milli\molar} dithiothreitol, 
+  \SI{1}{\units\per\micro\litre} RNAse inhibitor, 
+  \SI{0.1}{\percent} Tween-20, 
+  \SI{1}{\percent} BSA, 
+  \SI{10}{\milli\molar} Tris-HCl,
+  \SI{10}{\milli\molar} NaCl, 
+  \SI{3}{\milli\molar} $MgCl_2$
+), pelleting the isolated nuclei (500g, 5 min) and resuspending in 1X nuclei buffer (10X Genomics). 
+Single-cell libraries were generated using the 10X Chromium Single-Cell Instrument and 
+NextGEM Single Cell Multiome ATAC+Gene Expression kit (10X Genomics) according to the manufacturer’s protocol as 
+described in the human neural tube organoids section. 
+In total, 4 ATAC and 4 Gene Expression libraries were generated for the embryonal sample.
 
 \paragraph*{10x multiome library sequencing}
-%TODO: Camiel
+The fragment size of the libraries were analyzed using the Bioanalyzer high-sensitivity chip. 
+All embryonal libraries were first sequenced on the NextSeq2000 instrument (Illumina) before being sequenced deeper on 
+the NovaSeq6000 instrument (lllumina). 
+The ATAC libraries were sequenced with the following parameters: 50 bp read 1, 8 bp index 1, 24 bp index 2, 49 bp read 2. 
+The Gene Expression libraries were sequenced with the following parameters: 28 bp read 1, 10 bp index 1, 10 bp index 2, 90 bp read 2.
 
@@
 Cells were first annotated as Facial mesenchyme, Neural crest, Neuron, Peripheral neuron, Pluripotent stem cell or
-Neuronal progenitor by 
+neural progenitor by 
 first normalizing, log transforming and scaling the gene expression matrix using the Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions normalize\_per\_cell, log1p and scale
@@
 For each cluster, the annotation with minimal distance was selected breaking ties by selecting a random annotation with equal distance.
-Neuronal progenitors were further annotated as Neuronal progenitor, Neuronal progenitor early neuron, KRT19+ neuronal
-progenitors, FOXG1+ neuronal progenitors, NFE2L3+ neuronal progenitors and KRT19+/NFE2L3+ neuronal progenitors using the
-same procedure at a Leiden resolution of 1.9 after subsetting for neuronal progenitors and re-normalizing, scaling and
+neural progenitors were further annotated as neural progenitor, neural progenitor early neuron, KRT19+ neural
+progenitors, FOXG1+ neural progenitors, NFE2L3+ neural progenitors and KRT19+/NFE2L3+ neural progenitors using the
+same procedure at a Leiden resolution of 1.9 after subsetting for neural progenitors and re-normalizing, scaling and
 integrating (see batch effect correction) the data.
-non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors were further annotated as FP, p3, p0-2, dp1-6 or RP by subsetting non FOXG1+, NFE2L3+ or KRT19+ neuronal progenitors, re-normalizing, scaling and integrating (see batch effect correction) the data.
+non FOXG1+, NFE2L3+ or KRT19+ neural progenitors were further annotated as FP, p3, p0-2, dp1-6 or RP by subsetting non FOXG1+, NFE2L3+ or KRT19+ neural progenitors, re-normalizing, scaling and integrating (see batch effect correction) the data.
 Next, a force-directed graph embedding\cite{islamCharacterizationSinglecellTranscriptional2011, jacomyForceAtlas2ContinuousGraph2014} was calculated using the Scanpy function draw\_graph followed by diffusion pseudotime inference\cite{haghverdiDiffusionPseudotimeRobustly2016} using the Scanpy function dpt and binned in 15 and 50 bins using the Pandas\cite{thepandasdevelopmentPandasdevPandasPandas2024} (v1.5.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018214/resolver}{RRID:SCR\_018214}) cut function.
@@
 mesenchyme and peripheral neuron manually based on the expression of marker genes.
-Neuron cells and Neuronal progenitor early neuron cells were further annotated as ASCL1+ Neurons, SST+ Neurons, SOX2+ Neurons,
+Neuron cells and neural progenitor early neuron cells were further annotated as ASCL1+ Neurons, SST+ Neurons, SOX2+ Neurons,
 ASCL1+ Neurons, UNCX+ Neurons, GAD1+ Neurons and GATA2+ Neurons based on the expression of marker genes.
@@
 \paragraph*{single-cell ATAC topic modeling}
-Four topic models were generated: one on all cells, one on the subset of cells annotated as neuronal progenitors, one on
-the subset of cells annotated as neuron (Neuron cells and Neuronal progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
+Four topic models were generated: one on all cells, one on the subset of cells annotated as neural progenitors, one on
+the subset of cells annotated as neuron (Neuron cells and neural progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
 Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/datmccallumMALLETMachineLearning2002618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
@@
 A model of 50 topics was selected for the model containing all cells,
-a model of 30 topics was selected for the model containing neuronal progenitors,
+a model of 30 topics was selected for the model containing neural progenitors,
 a model of 25 topics was selected for the model containing neurons and
@@
 Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
-As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
+As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neural progenitors,
 neurons and neural crest together with 10 topics form the topic model containing all cells representing the pluripotent
@@
   \item $0 \leq class_{index} < 25 $: neuron
-  \item $25 \leq class_{index} < 55$: Neuronal progenitor
+  \item $25 \leq class_{index} < 55$: neural progenitor
   \item $55 \leq class_{index} < 65$: Neural crest
@@
 \paragraph*{RNA count matrices and ATAC fragment files}
-
-% TODO: Camiel
+single-cell RNA count matrices and single-cell ATAC fragment files were generated using the 
+Cell Ranger ARC (v2.0.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_023897/resolver}{RRID:SCR\_023897}) command cellranger-arc count using GRCh38.p13 Gencode V35 primary sequence assembly as a reference genome as described in Braun \textit{et al.}\cite{braunComprehensiveCellAtlas2023}.
 
 \paragraph*{single-cell RNA quality control}
-
-% TODO: Camiel
+Cells with less than 3 genes expressed and genes expressed in less than 200 cells were removed using the 
+Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.8.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions filter cells and filter genes. 
+Next, quality control metrics were calculated using the Scanpy function calculate qc metrics using qc vars=[”mt”], 
+percent top = [20] and log1p = True as parameters. 
+Cells were filtered if they were more than 5 median absolute distances (MADs) away from the median log1p transcript count,
+log1p number of genes or percentage counts in top 20 genes, 
+or more than 3 MADs away from the median mitochondrial fraction, resulting in 34,453 high quality cells. 
+Doublets were detected using Scrublet\cite{wolockScrubletComputationalIdentification2019} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_018098/resolver}{RRID:SCR\_018098}) with a threshold of 0.2 and expected doublet rate of 6\%.
+Counts were normalized to 10,000, log transformed and scaled. 
+The top 20,000 highly variable genes were identified using the highly\_variable\_genes function using flavor=’seurat’. 
+The top 50 PCA components were used to calculate KNN neighbors and leiden clustering at a resolution of 0.3.
 
 \paragraph*{single-cell RNA cell type annotation}
-
-%TODO: Camiel
+Clusters that were \textit{TUBB3}+, \textit{ELAVL3}+, \textit{INA}+, \textit{STMN2}+ and \textit{MAP2}+ were annotated as neurons, 
+with the additional expression of \textit{SIX}1 indicating peripheral neurons. 
+\textit{SOX2}, \textit{PAX6}, \textit{RFX4} and \textit{PTPRZ} were used to annotate neural progenitor cells. 
+Neural crest was identified based on the expression of \textit{SOX10}, \textit{TFAP2A}, \textit{TFAP2B}, 
+\textit{PAX3} and \textit{GRHL2}. 
+\textit{FOXC1}, \textit{FOXC2} and \textit{TWIST1} were used as mesenchymal markers 
+and \textit{FLT1}, \textit{PTPRM} and \textit{CLDN5} conversely were used to annotate endothelial cells. 
+\textit{OTOGL} and \textit{OC90} marked the Otic placode and microglia were based on \textit{RUNX1} and \textit{CSF1R}.
+Neurons and neural progenitors were subclustered and neuronal subtypes were annotated based on the expression of 
+\textit{ASCL1}, \textit{UNCX}, \textit{GAD1}, \textit{GATA2}, \textit{SST} and \textit{SOX2}.
+Conversely, the floor and roof plate in were identified among the neural progenitor cells based on the expression of 
+\textit{SHH} and \textit{MSX1} respectively. 
+Conversely, regionalization was assessed based on expression of
+\textit{FOXG1}/\textit{EMX2} (forebrain), \textit{OTX2}/\textit{EN1} (midbrain), \textit{PAX8} (hindbrain) and \textit{HOXB3} (spinal cord).
 
@@
 
-%TODO: Camiel
+Pseudobulk ATAC-seq profiles were generated for each of the leiden clusters by using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, RRID:SCR\_026618) export pseudobulk function, with default parameters.
+Peak calling was done using the pycisTopic peak calling function, which uses MACS\cite{zhangModelbasedAnalysisChIPSeq2008} (v2.2.9.1; RRID:SCR 013291 ). 
+Per cluster peaksets were merged using the pycisTopic function get consensus peaks, 
+using a peak width of 500 bp and default parameters, resulting in 610,282 peaks.
 
@@
 \paragraph*{single-cell ATAC topic modeling}
-Three topic models were generated: one on the subset of cells annotated as neuronal progenitors, on on the subset of
+Three topic models were generated: one on the subset of cells annotated as neural progenitors, on on the subset of
 cells annotated as neuron and one on the subset of cells annotated as neural crest. First, a fragment count matrix was
-generated for each sample using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function create\_fragments\_matrix\_from\_fragments, this matrix was binarized (\textgreater 0), merged in one count matrix using the Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) function sparse.hstack and subsequently split in a matrix for each of neuronal progenitors, neurons and neural crest cells and saved as matrix market format.
+generated for each sample using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v2.0a, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) function create\_fragments\_matrix\_from\_fragments, this matrix was binarized (\textgreater 0), merged in one count matrix using the Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}) function sparse.hstack and subsequently split in a matrix for each of neural progenitors, neurons and neural crest cells and saved as matrix market format.
 A mallet corpus file was generated for each binary matrix using the pycisTopic command pycistopic topic\_modeling mallet create\_corpus and finally topic modeling was run using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; RRID:SCR\_026708) using the pycisTopic command: pycistopic topic\_modeling mallet run using 300 iterations and default parameters otherwise.
 Model evaluation was performed using the pycisTopic command pycistopic topic\_modeling mallet stats that calculates model coherence\cite{mimno-etal-2011-optimizing}, log-likelihood\cite{griffithsFindingScientificTopics2004}, model density\cite{arunFindingNaturalNumber2010} and a divergence-based metric\cite{caoDensitybasedMethodAdaptive2009}.
-A model of 60 topics was selected for the model containing neuronal progenitors,
+A model of 60 topics was selected for the model containing neural progenitors,
 a model of 30 topics was selected for the model containing neurons and
@@
 Model was built and trained using Tensorflow (v2.41; \href{https://rrid.site/data/record/nlx_144509-1/SCR_016345/resolver}{RRID:SCR\_016345}). 
-As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neuronal progenitors,
+As training data all binarized topics (see single-cell ATAC topic modeling) from the subset of neural progenitors,
 neurons and neural crest together were used for a total of 120 classes/topics. 
@@
   \item $0 \leq class_{index} < 30 $: neuron
-  \item $30 \leq class_{index} < 90$: Neuronal progenitor
+  \item $30 \leq class_{index} < 90$: neural progenitor
   \item $90 \leq class_{index} < 120$: Neural crest
@@
 using the --cut-sites option.
-Next, for each pattern (converted to a PWM) of the FOX cluster of neuronal progenitors (cluster 6) motif hits were calculated on the top
-1,000 regions of all neuronal progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
+Next, for each pattern (converted to a PWM) of the FOX cluster of neural progenitors (cluster 6) motif hits were calculated on the top
+1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
 Overlapping hits were resolved by taking the hit with the highest $-log_{10}(pvalue)$. 
@@
 FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793}).
-For each pattern (converted to a PWM) of all neuronal progenitor pattern cluster motif hits were calculated on the top
-1,000 regions of all neuronal progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
+For each pattern (converted to a PWM) of all neural progenitor pattern cluster motif hits were calculated on the top
+1,000 regions of all neural progenitor topics using FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using default parameters.
 For each sequence the maximum hit, based on the average contribution score of that hit for the topic from which the
@@
 \subsubsection*{logistic regression models for each cell states}
-For each cell type, neuronal progenitors dorso-ventral patterning, neuronal progenitors anterior-posterior patterning, neurons,
+For each cell type, neural progenitors dorso-ventral patterning, neural progenitors anterior-posterior patterning, neurons,
 and neural crest, motif scores for all patterns (converted to a PWM) patterns belonging to clusters of the cell type\lq s
@@
   \caption{
-    \textbf{Linking motif to TF in dorso-ventral neuronal progenitors}.\\
+    \textbf{Linking motif to TF in dorso-ventral neural progenitors}.\\
     \textbf{a-h} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
@@
   \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_4.png}
-  \caption{\textbf{Linking motif to TF in anterior-posterior neuronal progenitors}.\\
+  \caption{\textbf{Linking motif to TF in anterior-posterior neural progenitors}.\\
     \textbf{a-e} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
@@
   \caption{
+    \textbf{Genome wide performance of logistic regression models.}\\
+    \textbf{a-b}, Scatter plot showing performance of the logistic regression models as  area under the (AUC) precision recall (PR) curve (\textbf{a}) and receiver operator curve (ROC) (\textbf{b}) for different cell types for organoid (O) and embryo (E) represented by topic numbers.
+Performance is shown to classify the contrast between topics of the same major class (i.e., neuron progenitor dorso-ventral, neuron progenitor anterior-posterior, neural crest and neuron; x-axis) and genome wide classification of consensus peaks as belonging to a certain topic or not.
+}
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_9.png}
+  \caption{
     \textbf{SOX dimer instances with 4bp spacing are most accessible in migratory neural crest}\\
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_9.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_10.png}
   \caption{

### 2025-04-18 manuscript.tex
Move parts of figure 3 to supplementary figure 4

@@
     \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
-    \textbf{i}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance.
-Signal is centered on the FOXA2 instance start position.
-    \textbf{j}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIP-seq peaks (general; bottom) and those regions that do not overlap (specific; top).
-    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 2,130) and specific (n = 4,529) regions, for organoid.
-    \textbf{k}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions.
-* indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
-    \textbf{l}, de novo motif for regions with a single FOXA2 instance that are specific (left, top) or general (right, top) and heatmap color coded by nucleotide identity for individual regions for both specific (left, bottom) and general (right, bottom) regions that have a single FOXA2 instance.
 }
@@
 Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
-and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). 
+and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S5). 
 In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
@@
 with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
-topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4)\par
+topic (\textbf{Fig. 4d-e} and Supplementary Fig. S5)\par
 %TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
@@
 GRHL with \textit{GRHL1/1} 
-and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
+and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
 Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
@@
 5d} and Supplementary Fig.
-S5).
+S6).
 This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
@@
 ZIC with \textit{ZIC1} to \textit{ZIC5} and 
-the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
+the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
 The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
 After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) 
-for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par
+for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S6).\par
 
@@
 The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
-that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). 
+that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
 The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
@@
 Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
-Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S6}). 
+Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S7}). 
 As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
@@
 correlating with the expression of respectively \textit{TEAD1},
-\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
 Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
-correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
 Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
@@
 (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
-(\textbf{Fig. 6d} and Supplementary Fig. S7) 
+(\textbf{Fig. 6d} and Supplementary Fig. S8) 
 and topics where a combination of EBF and ONECUT instances are important 
@@
 The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
-and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). 
+and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
 Finally, in the organoids we could identify an additional topic with instances of GATA
-(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par
+(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S8).\par
 %TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
@@
 Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
-However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S8). \par
+However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S9). \par
 
@@
 Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bp spacing,
-with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 9}). 
+with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 10}). 
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
@@
 systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
-(\textbf{Supplementary Fig. S10}).\par
+(\textbf{Supplementary Fig. S11}).\par
 
@@
   \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_4.png}
+  \caption{
+    \textbf{a}, Scaled average and normalized number of cut sites (black, top) and contribution score (red, top) across genomic regions with a FOXA2 instance and heatmap showing normalized number of cut sites (bottom) for individual regions with a FOXA2 instance.
+Signal is centered on the FOXA2 instance start position.
+    \textbf{b}, Bar chart showing the fraction of genomic regions that has a given number of FOXA2 instances for FOXA2 regions that overlap with HepG2 or A549 FOXA2 ChIP-seq peaks (general; bottom) and those regions that do not overlap (specific; top).
+    Two-sided p-value calculated using Mann-Whitney U test between distribution of number of FOX instances in general (n = 2,130) and specific (n = 4,529) regions, for organoid.
+    \textbf{c}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions.
+* indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
+  }
+\end{figure}
+
+\newpage
+\begin{figure}[p]
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_5.png}
   \caption{\textbf{Linking motif to TF in anterior-posterior neural progenitors}.\\
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_5.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_6.png}
   \caption{
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_6.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_7.png}
   \caption{
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_7.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_8.png}
   \caption{
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_8.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_9.png}
   \caption{
@@
 \begin{figure}[p]
-  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_10.png}
+  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_11.png}
   \caption{

### 2025-04-18 manuscript.tex
Add microscopy images to figure 3

@@
 \author[1,2,3,\orcidlink{0000-0001-7907-1247}]{Seppe De Winter}
-\author[1,2,3, 4, \orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
-\author[1,2,3]{Valerie Christiaens}
-\author[4, \orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
-\author[5, 6, 7, \orcidlink{0000-0002-2628-7834}]{Ruth M. Williams}
-\author[5, 6, \orcidlink{0000-0001-9289-0263}]{Tatjana Sauka-Spengler}
-\author[4, \orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
+\author[1,2,3,4,\orcidlink{0000-0002-1318-2603}]{Camiel Mannens}
+\author[1,2,3,\orcidlink{0000-0002-5024-8691}]{Valerie Christiaens}
+\author[1,2,3,\orcidlink{0000-0002-0876-2179}]{Roel Vandepoel}
+\author[4,\orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
+\author[5,6,7,\orcidlink{0000-0002-2628-7834}]{Ruth M. Williams}
+\author[5,6,\orcidlink{0000-0001-9289-0263}]{Tatjana Sauka-Spengler}
+\author[4,\orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
 \author[1,2,3,*,\orcidlink{0000-0002-8006-0315}]{Stein Aerts}
@@
 SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
+When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
+(\textbf{Fig. 3d}).
 Explaining the accessibility of both enhancers using DeepNeuralTube reveals two FOX (one partial) and two RFX (one partial) binding sites in SFPE1 and two FOX, a ZEB and a TEAD binding site in SFPE2 (\textbf{Fig.
-3d}).\par
+3e}).\par
 
@@
 Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
-resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}).
+resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
 Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
-importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3e}). 
+importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
 For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
 are important (\textbf{Fig.
-3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig.
-3g} and Supplementary Fig.
+3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig.
+3h} and Supplementary Fig.
 S3).
-In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); 
-PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and 
-ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
+In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
+PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3) and 
+ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
 For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
-SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3e-g} and Supplementary Fig. S3). 
-Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). 
+SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
+Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Supplementary Fig. S3). 
 Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
-3h}).
-FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3h}).\par
+3i}).
+FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par
+
+To assess the ability of DeepNeuralTube for prioritizing enhancer with cell type-specific activity we selected a genomic
+region that was predicited to be specific for ventral neural progenitors (\textbf{Fig. 3k}). Explaining this region,
+reveals multiple FOX instances and a SOX instance by both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 3j}). 
+Finally, we tested the activity of this region in chicken embryos, revealing that it indeed is specifically active in
+the floorplate of the neural tube (\textbf{Fig. 3l}).\par
 
 It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
-Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. 3i}). 
+Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Supplementary Fig. S4a}). 
 We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
@@
 FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
-specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. 3j}). 
+specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Supplementary Fig. S4b}). 
 Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
-specific regions (\textbf{Fig. 3k}). These results suggest that the overall affinity of general FOXA2 target regions 
+specific regions (\textbf{Supplementary Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
 is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
-Generating a \textit{de novo} motif from general and specific FOXA2 target regions that have a single FOX instance 
-reveals that FOX instances in regions that are generally bound tend to be of the form \textit{TGTTT\underline{AC}} 
-while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par
-%TODO: Revisit this analysis but intersect with floor plate topic instead of FOX hits!
 
@@
     \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
-    \textbf{d}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
-    \textbf{e}, Code table showing average contribution score of pattern instances for each dorso-ventral topic.
+    \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
+    green is GFP driven by SFPE1 enhancer.
+    \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
+    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic.
 Contribution scores for $DeepNeuralTube^o$ are not outlined.
 Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
-    \textbf{f}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
-    \textbf{g}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
-    \textbf{h}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
+    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
+    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
+    \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
+    \textbf{j}, Nucleotide contribution score for chr18:77367159-77367659; hg38 for organoid topic 33 (top) and embryo
+    topic 34 (bottom).
+    \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
+    embryo (botto).
+    \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
+    green is GFP driven by chr18:77367159-77367659; hg38 enhancer.
 }

### 2025-04-18 manuscript.tex
contributions
make text of fig 1 clearer

@@
 with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Supplementary Fig. S1c-d}).\par
-
-%TODO:D) The black outline on green text (EMX2/EN1 etc) makes it difficult to read and see the color. consider making it thinner
 
@@
   \item Conceptualization: S.D.W. \& S.A.
-  \item Experiments and sample preparation: S.D.W., V.C., E.S., L.H. \& C.M.
+  \item Experiments and sample preparation: S.D.W., V.C., L.H., C.M., R.M.W., \& R.V.P.
   \item Computational analysis: S.D.W. \& C.M.
-  \item Writing – original draft: S.D.W. \& S.A.
-  \item Writing – review \& editing:
+  \item Writing – original draft: S.D.W., C.M. \& S.A.
+  \item Writing – review \& editing: S.D.W., C.M. \& S.A.
 \end{itemize}
@@
 Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
-Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq.
-Using PCR.
+Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR.
 A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and
-%TODO: Add concentration
 PCR program: \SI{95}{\degreeCelsius} for 3 minutes, 15 cycles of: \SI{98}{\degreeCelsius} for 20 seconds,
 \SI{58}{\degreeCelsius} for 15 seconds and \SI{72}{\degreeCelsius} for 2.5 minutes, followed by \SI{72}{\degreeCelsius}
-for 5 minutes.
+for 5 minutes was used.
 Linarized plasmids were purified using QIAquick PCR purification kit (Qiagen, 28104), followed by DpnI digestion for 1
@@
 TN5 cut site BigWigs were calculated for all organoid topics using the bigwig command from
-scatac\_fragment\_tools ()
-%TODO: add RRID
+scatac\_fragment\_tools (v0.1.3; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026713/resolver?}{RRID:SCR\_026713})
 using the --cut-sites option.

### 2025-04-22 manuscript.tex
acknowledgments

@@
     \textbf{k}, Prediction scores for chr18:77367159-77367659; hg38 for doros-ventral topics in organoid (top) and
-    embryo (botto).
+    embryo (bottom).
     \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
-    green is GFP driven by chr18:77367159-77367659; hg38 enhancer.
+    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}
 }
@@
 
+We thank Peter L\"onnerberg for preprocessing the human embryo data and making it available through EGA.
+
 %%%  Use this section to acknowledge contributions 
@@
 %\end{description}
+
+% TODO: see https://www.cell.com/information-for-authors/star-supplemental-information
 

### 2025-04-25 manuscript.tex
M&M: cryosectioning

@@
 embryo was elektroorated (5 pules of 5V: 100 ms on, 50 ms off). Embryo was grown for 24 hours at \SI{37}{\degreeCelsius}
-and imaged whole mount
+and imaged whole mount or sectioned and imaged.
+
+\paragraph*{Chicken embryo cryosectioning}
+Embryos were fixed in \SI{4}{\percent} paraformaldehyde for 1 hour at room temperature followed by three 
+10 minutes washes with PBS. 
+Embryos were transferred to \SI{15}{\percent} V/V sucrose in PBS at room temperature until they sank to the bottom. 
+Next, embryos were transferred to a mixture consiting of \SI{7.5}{\percent} V/V gelatin and \SI{15}{\percent} V/V 
+sucrose in PBS and were incubated overnight at \SI{37}{\degreeCelsius}. 
+In the meantime \SI{20}{\percent} V/V gelatin was dissolved in PBS at \SI{60}{\degreeCelsius} and this was kept at 
+\SI{37}{\degreeCelsius} once the gelatin was dissolved. 
+After the overnight incubating, embryos were transferred to the gelatin solution and kept for a minimum of 4 hours at 
+\SI{37}{\degreeCelsius}. 
+Finally, embryos in gelatin solution were transferred to a cryomold and quickly frozen on an aluminum block cooled by 
+dry ice.
+Cryosections of \SI{20}{\micro\meter} were made at \SI{-14}{\degreeCelsius} and collected on SuperFrost Plus Adhesion 
+slides (Fisher Scientific, 10149870). 
+Sections were stained with DAPI (\SI{0.1}{\milli\gram\per\milli\litre} in PBS; Sigma-Aldrich, 28718-90-3) 
+for 15 minutes, followed by a PBS wash and air dried. Cover slip was mounted using Mowiol 4-88 mounting medium
+(Sigma-Aldrich, 81381-50G).
 

### 2025-04-25 manuscript.tex
Abstract

@@
 %task and results
-In this study we generate a single-cell multiome atlas of human neural tube organoids 
+We generate a single-cell multiome atlas of human neural tube organoids 
 and of the head of a four weeks post conception human embryo. 
@@
 for each individual cell type. 
-We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region. 
-Enhancers encoding anterior-posterior patterning form an exception to this rule, being encoded as
-homotypic clusters of homeodomain binding sites.
-% TODO: S.A. [some more results? Validation examples? Genetic variants?]
+We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region, using soft syntactic rules. 
+We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
+Finally, we validate DeepNeuralTube using enhancer reporter assays and on human genetic variants.
 %conclusion
-This compendium of enhancer codes provides gene regulatory insight into human development.
+The enhancer compendium is available as a resource at: something.aertslab.org
 %implication

### 2025-04-25 manuscript.tex
Data and code

@@
 %%%  components. Each component must be present.
+
+Raw sequencing data, scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks, and cell- and 
+region topic probabilities for the human neural tube organoids are available on Gene Expression Omnibus (GEO) using 
+accession number \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE295337}{GSE295337}
+(reviewer token: ). %TODO: add before submission.
+Raw sequencing data of the human embryo are available through the European Genome Phenome Archive 
+(). %TODO: Add link to EGA
+scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks and cell-and region topic
+probabilities of the human embryos are available through Zenodo (). %TODO: upload embryo data to Zenodo and provide link.
+$DeepNeuralTube_O$ and $DeepNeuralTube_E$ are available via Zenodo (). %TODO: upload models to Zenodo and provide link.
+Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the \href{}{UCSC genome browser} %TODO: build genome browser track hub.
+The enhancer compendium generated in this study is available at something.aertslab.org. %TODO: generate landing page
+
+Code to reproduce all figures in this manuscript as well as a full history of the manuscript itself is available 
+through GitHub: \href{https://github.com/SeppeDeWinter/DeWinter_NeuralTube}{SeppeDeWinter/DeWinter\_NeuralTube}.
 

### 2025-04-25 manuscript.tex
Abstract

@@
 
-%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
-
-%Context
-Enhancers encode cell types as clusters of transcription factor binding sites. 
-%need
+By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
+recruitment. 
 Enhancer code studies have been limited to specific cell types in model organisms. 
-%task and results
-We generate a single-cell multiome atlas of human neural tube organoids 
-and of the head of a four weeks post conception human embryo. 
-Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that 
-are shared between the organoid model and the human embryo and we provide a compendium of enhancer codes 
-for each individual cell type. 
-We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 transcription factor binding sites in a nucleosome depleted region, using soft syntactic rules. 
+Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
+To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
+Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
+model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
+We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
 We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
-Finally, we validate DeepNeuralTube using enhancer reporter assays and on human genetic variants.
-%conclusion
-The enhancer compendium is available as a resource at: something.aertslab.org
-%implication
+Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
+facial mesenchyme enhancers. 
+Our neural tube enhancer compendium is available as a resource at: something.aertslab.org
 

### 2025-05-12 manuscript.tex
Supplement

@@
 resulting in 22,862 high quality cells (\textbf{Fig. 1b}). 
-Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the organoid system.\par
+Note that similar transcriptomic states were obtained from all three cell lines (\textbf{Fig. S1a-b}) demonstrating the robustness of the organoid system.\par
 
@@
 We annotated both the organoid and embryo cells based on the expression of marker genes
-(\textbf{Supplementary Table S1}). 
+(\textbf{Table S1}). 
 In both systems we could identify neural progenitors, early differentiating neurons,
@@
 We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
-with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Supplementary Fig. S1c-d}).\par
+with an average Pearson correlation coefficient of 0.83 across matched states (\textbf{Fig. S1c-d}).\par
 
@@
     TWIST1 [0, 2.0].
+    Also see Figure S1
   }
@@
 (auROC) of 0.75 and 0.80 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2b}; and
-area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})). 
+area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Fig. S2a-b})). 
 Enhancers active in the facial mesenchyme (n=143) could be classified with an auROC of 0.78 and 0.79 
 using resp. $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 2c}; 
-and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). 
+and auPR of 0.17 for both models (\textbf{Fig. S2a-b})). 
 Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
@@
 All clusters, except cluster 47 (organoid specific CG motif), contain known motifs enriched in regions specific to both 
-organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Supplementary Table S2}).
-More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Supplementary Table S2}). 
+organoid and embryo cell types (\textbf{Fig. 2e} and \textbf{Table S2}).
+More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Table S2}). 
 Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
@@
 
-As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). 
+As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Fig. S2c}). 
 This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). 
@@
 a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
-we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Supplementary Fig. S2d}). 
-Both models predict this region to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e})
+we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
+Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
 and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
@@
     constitutive enhancer that serves as electroporation control.
+    Also see Figure S2
   }
@@
 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig.
-3h} and Supplementary Fig.
+3h} and Fig.
 S3).
 In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
-PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3) and 
-ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
+PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
+ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
 For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
-SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Supplementary Fig. S3). 
-Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Supplementary Fig. S3). 
+SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
+Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
 Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
@@
 It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
-Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Supplementary Fig. S4a}). 
+Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
 We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
@@
 FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
-specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Supplementary Fig. S4b}). 
+specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. S4b}). 
 Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
-specific regions (\textbf{Supplementary Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
+specific regions (\textbf{Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
 is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
@@
     \textbf{l}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
-    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}
+    green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
+    Also see Figure S3 and S4.
 }
@@
 Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
-and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S5). 
+and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
 In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
@@
 with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
-topic (\textbf{Fig. 4d-e} and Supplementary Fig. S5)\par
+topic (\textbf{Fig. 4d-e} and Fig. S5)\par
 %TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
@@
   \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
-  U.: UMAP
+  U.: UMAP. 
+  Also see Figure S5.
 }
@@
 GRHL with \textit{GRHL1/1} 
-and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
+and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Fig. S6). 
 Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
@@
 and moreover a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
-The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig.
-5d} and Supplementary Fig.
-S6).
+The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Fig. S6). 
 This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
@@
 ZIC with \textit{ZIC1} to \textit{ZIC5} and 
-the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
+the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Fig. S6). 
 The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}.
 After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) 
-for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S6).\par
+for which instances of RFX are important (\textbf{Fig. 5b-c}) correlating with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Fig. S6).\par
 
@@
 The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
-that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S6). 
+that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Fig. S6). 
 The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
@@
 Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
-Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S7}). 
+Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Fig. S7}). 
 As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
@@
   U: UMAP.
+  Also see Figure S6 and S7.
   }
@@
 correlating with the expression of respectively \textit{TEAD1},
-\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
+\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Fig. S8). 
 Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
-correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
+correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Fig. S8). 
 Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
@@
 (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
-(\textbf{Fig. 6d} and Supplementary Fig. S8) 
+(\textbf{Fig. 6d} and Fig. S8) 
 and topics where a combination of EBF and ONECUT instances are important 
@@
 The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
-and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S8). 
+and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Fig. S8). 
 Finally, in the organoids we could identify an additional topic with instances of GATA
-(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S8).\par
+(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Fig. S8).\par
 %TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
@@
   \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+  Also see Figure S8.
   }
@@
 Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). 
-However, the genome wide precision and recall of these models to identify topic regions is still low (Supplementary Fig. S9). \par
+However, the genome wide precision and recall of these models to identify topic regions is still low (Fig. S9). \par
 
@@
 deviations) for organoid (bottom left) and embryo (bottom right).
+Also see Figure S9.
 }
@@
 Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bp spacing,
-with instances with 4 bp spacing being the most accessible (\textbf{Supplementary Fig. 10}). 
+with instances with 4 bp spacing being the most accessible (\textbf{Fig. 10}). 
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
@@
 systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
-(\textbf{Supplementary Fig. S11}).\par
+(\textbf{Fig. S11}).\par
 
@@
 scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks and cell-and region topic
-probabilities of the human embryos are available through Zenodo (). %TODO: upload embryo data to Zenodo and provide link.
-$DeepNeuralTube_O$ and $DeepNeuralTube_E$ are available via Zenodo (). %TODO: upload models to Zenodo and provide link.
-Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the \href{}{UCSC genome browser} %TODO: build genome browser track hub.
+probabilities of the human embryos are available through Zenodo (\href{https://doi.org/10.5281/zenodo.15281826}{10.5281/zenodo.15281826}). 
+$DeepNeuralTube_O$ and $DeepNeuralTube_E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
+Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the UCSC genome browser: 
+\begin{itemize}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorso-ventral UCSC session}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neuron}{neuron UCSC session}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20anterior%2Dposterior}{anterior-posterior
+      UCSC session}
+\end{itemize}
 The enhancer compendium generated in this study is available at something.aertslab.org. %TODO: generate landing page
@@
 
+The authors declare no competing interests.
+
 \section*{DECLARATION OF GENERATIVE AI AND AI-ASSISTED TECHNOLOGIES}
@@
 
-% TODO: see https://www.cell.com/information-for-authors/star-supplemental-information
+% see https://www.cell.com/information-for-authors/star-supplemental-information
+
+\begin{description}
+  \item Document S1. Table S1
+  \item Table S2. Excel file containing motifs for clustering in Figure 2.
+\end{description}
 
@@
 %\textsuperscript{b}More footnotes
+%
+%NOTE: For the journal Cell, supplemental figures should instead be uploaded as separate files, the same process as the main figures; their titles and legends should be placed in the main text, directly after the main figure legends and any tables, under a “supplemental figures” heading
 
@@
 The median is used as the center.
-%TODO: make sure it is there.
 
@@
     the same biological state(matching) and other topics (non-matching) across organoid and embryo.
+    Related to Figure 1.
   }
@@
     $DeepNeuralTube_E$ (bottom) for each topic for neural crest enhancer chr1:161200798-161201298; hg38.
+    Related to Figure 2.
   }
@@
     (\textbf{h}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+    Related to Figure 3.
   }
@@
 * indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
+    Related to Figure 4.
   }
@@
     that is annotated to: Homeodomain or Homeodomain; Paired box or Paired box (\textbf{a-e}). Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+    Related to Figure 4.
   }
@@
     Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+    Related to Figure 5.
   }
@@
     $DeepNeuralTube_E$ (right) for SNPs with a minimal absolute delta prediction score of 0.2.
+    Related to Figure 5.
   }
@@
     Annotations according to Lambert et al.\cite{lambertHumanTranscriptionFactors2018}.
+    Related to Figure 6.
   }
@@
 Performance is shown to classify the contrast between topics of the same major class (i.e., neuron progenitor dorso-ventral, neuron progenitor anterior-posterior, neural crest and neuron; x-axis) and genome wide classification of consensus peaks as belonging to a certain topic or not.
+    Related to Figure 7.
 }

### 2025-05-12 manuscript.tex
KRT refs.

@@
 %%%  file.
+
+%TODO: add reference to https://www.sciencedirect.com/science/article/pii/S2213671114003580?via%3Dihub and put number in
+%KRT for Verfaillie et al. and pierre
 
@@
 were calculated using the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function integrated\_grad, on 1,000 regions with the highest prediction score for that topic according to the model.
-Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709).
+Next, patterns/motifs were discovered \textit{de novo} using a custom implementation of tfmodisco-lite (v2.0.0; RRID:SCR\_026709)\cite{schreiberTfmodiscolite2025}.
 First seqlets were called using the tfmodisco-lite function extract\_seqlets using the attribution scores as input and using parameters
@@
 Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
-Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522}) was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR.
+Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522})\cite{williamsReconstructionGlobalNeural2019} was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR.
 A PCR mixture of KAPA HiFi hot start ready mix (Roche, 07958935001 \SI{50}{\percent} V/V), \SI{0.8}{\micro\gram\per\micro\litre} plasmid and \SI{0.3}{\micro\molar} of forward and reverse primer and
@@
 \paragraph*{Injection control cloning}
-Plasmid pCI H2B-RFP (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_92398/resolver}{RRID:Addgene\_92398})
+Plasmid pCI H2B-RFP
+(\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_92398/resolver}{RRID:Addgene\_92398})\cite{williamsReconstructionGlobalNeural2019}
 was ordered from Addgene (\href{https://rrid.site/data/record/nlx_144509-1/SCR_002037/resolver}{RRID:SCR\_002037}),
@@
 \subsubsection*{General and specific FOXA2 target regions}
-FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793}).
+FOXA2 ChIP-seq peaks in HepG2 (\href{https://www.encodeproject.org/files/ENCFF466FCB/}{ENCFF466FCB}) and A549 \href{https://www.encodeproject.org/files/ENCFF686MSH/}{ENCFF686MSH} were downloaded from ENCODE (\href{https://rrid.site/data/record/nlx_144509-1/SCR_006793/resolver}{RRID:SCR\_006793})\cite{encodeprojectconsortiumIntegratedEncyclopediaDNA2012, luoNewDevelopmentsEncyclopedia2020, hitzENCODEUniformAnalysis2023}.
 For each pattern (converted to a PWM) of all neural progenitor pattern cluster motif hits were calculated on the top

### 2025-05-15 manuscript.tex
Update author & acknowledgments

@@
 
-\title{An enhancer code compendium of early human neural and neural crest development}
+\title{An enhancer code compendium of early human neuronal and neural crest development}
 \date{}
@@
 \author[4,\orcidlink{0000-0003-1869-0372}]{Lijuan Hu}
-\author[5,6,7,\orcidlink{0000-0002-2628-7834}]{Ruth M. Williams}
-\author[5,6,\orcidlink{0000-0001-9289-0263}]{Tatjana Sauka-Spengler}
 \author[4,\orcidlink{0000-0002-3491-3444}]{Sten Linnarsson}
@@
 \affil[4]{Karolinska Institute, Department of Biochemistry and Biophysics, Stockholm, Sweden}
-\affil[5]{Stowers Institute for Medical Research, Kansas City, Missouri, United States of America}
-\affil[6]{University of Oxford, MRC Weatherall Institute of Molecular Medicine, Radcliffe Department of Medicine, Oxford, United Kingdom}
-\affil[7]{University of Manchester, Faculty of Biology, Medicine and Health, Manchester, United Kingdom}
 
@@
 We thank Peter L\"onnerberg for preprocessing the human embryo data and making it available through EGA.
+We also thank Prof. Dr. Tatjana Sauka-Spengler and Dr. Ruth Williams for their great support with helping setting up the chicken embryo enhancer reporter assays.
+Computing was performed at the Vlaamse Supercomputer Center and VIB datacore. 
+This work was funded by the following grants to 
+S. Aerts , %TODO: Add funding
+PhD fellowship from FWO to S. De Winter (1191323N) and 
+Post doctoral fellowship from ... to C. Mannens (), %TODO: add funding
 
@@
   \item Conceptualization: S.D.W. \& S.A.
-  \item Experiments and sample preparation: S.D.W., V.C., L.H., C.M., R.M.W., \& R.V.P.
+  \item Experiments and sample preparation: S.D.W., V.C., L.H., C.M., \& R.V.P.
   \item Computational analysis: S.D.W. \& C.M.

### 2025-05-15 manuscript.tex
Changes troughout

@@
 facial mesenchyme enhancers. 
-Our neural tube enhancer compendium is available as a resource at: something.aertslab.org
+We provide tools for building enhancer code compendia through our genomic deep learning package, CREsted.
 
@@
 We compare this compendium to that of the head of a four weeks post conception human embryo.
-Through these comparisons we assess the enhancer code paradigm, 
-and also provide counter examples specifically related to the enhancer code underlying anterior-posterior patterning.
+Through these comparisons we assess the enhancer code paradigm.
 \par
@@
 neural crest states\cite{abdelfattahActuationEnhancesPatterning2021}. 
-To characterize the organoid-derived cell types to a greater level of detail at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and
-scRNA-seq) 
+To characterize the organoid-derived cell types at both transcriptomic as well as chromatin accessibility level we performed 10x single-cell multiomics (combined scATAC-seq and scRNA-seq) 
 at four time points of differentiation (day 4-5, day 6-7, day 8-9 and day 10-11; \textbf{Fig. 1a}), 
@@
 we identified sets of co-accessible regions in the organoid model using topic modeling 
-and identified equivalent cells in the embryo where the same regions were also co-accessible.
+and identified equivalent cells in the embryo (defined as cells where the same regions were also co-accessible).
 We observed similar patterns of marker gene expression for the same cell state across the organoid and embryo (\textbf{Fig. 1i}) 
@@
 we trained two sequence-to-function model convolutional neural networks:
-"$DeepNeuralTube^o$" and "$DeepNeuralTube^e$", on the organoid and embryo atlases respectively. 
+"$DeepNeuralTube^o$" and "$Deep$-$NeuralTube^e$", on the organoid and embryo atlases respectively. 
 The models use the DNA sequence of genomic regions as input and classify regions based on their cell type- or state-specific chromatin accessibility (in a multi-label multi-class fashion; \textbf{Fig.
@@
     colored by: cluster identity (\textbf{c}), model of origin 
-    (\textbf{d}; only showing \textit{de novo} motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
+    (\textbf{d}; only showing known motifs in this case) and whether the motif is found \textit{de novo} (Deep Learning) or by motif enrichment (cisTarget; \textbf{e}).
     \textbf{f}, Example representative motifs from clusters in b, showing motifs from motif enrichment (left) and found
@@
 identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
-From ventral to dorsal this is topic 36, 38, 33, 54 and 48 for the organoids and
+From ventral to dorsal this is topic 33, 38, 36, 54 and 48 for the organoids and
 topic 34, 38, 79, 88 and 58 for the embryo. 
@@
 floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
-\textit{SHH} is most strongly expressed in the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
+\textit{SHH} is most strongly expressed in the cells of the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
 SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
@@
 Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
-Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig.
+Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig.
 3i}).
@@
     green is GFP driven by chr18:77367159-77367659; hg38 enhancer. Scale bar represents \SI{500}{\micro\meter}. 
+    O. organoid. E. Embryo.
     Also see Figure S3 and S4.
@@
 \subsection*{
-  Homotypic clusters of distinct homeodomain transcription factor binding sites encode anterior-posterior patterning
+  Distinct homeodomain transcription factor binding sites encode anterior-posterior patterning
 }
@@
 for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
-From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
+From anterior to posterior this is \textit{TAATTAG}, \textit{TAATCCCT}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
 Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
@@
   \textbf{c}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis). 
-  \textbf{d}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for all anterior-posterior topic (x-axis). 
+  \textbf{d}, Heatmap showing the log of the number of seqlets for each identified pattern (y-axis) for each anterior-posterior topic (x-axis). 
   \textbf{e}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in embryo
@@
   \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
-  U.: UMAP. 
+  U.: UMAP. AP: anterior-posterior, DV: dorso-ventral, iter: iteration. 
   Also see Figure S5.
@@
   \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) neural crest cells colored by cell-topic probabilities in RGB color scales for neural crest related topics. 
-  \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic.
-Contribution scores for $DeepNeuralTube^o$ are not outlined.
-Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
+  \textbf{b}, Code table showing average contribution score of pattern instances for each neural crest related topic.  Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
   \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neural crest related topics. 
@@
 The location of the SNP is indicated with a dotted line.
-  U: UMAP.
+  U: UMAP. O.: organoid. E.: Embryo. bps: base pairs. 
   Also see Figure S6 and S7.
@@
 
-During early neural differentiation, we identified several topics for different stages of neural fate
-acquisition in both organoid and embryo cells (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
+For the cells representing early neural differentiation, we identified several topics for different stages of neural fate
+acquisition in both organoid and embryo (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
 $DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
@@
   \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics. 
-  \textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic.
-Contribution scores for $DeepNeuralTube^o$ are not outlined.
-Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
+  \textbf{b}, Code table showing average contribution score of pattern instances for each neuron related topic. Contribution scores for $DeepNeuralTube^e$ are outlined and negated. 
   \textbf{c}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) of neuron related topics. 
@@
   \textbf{e}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns.
+  U.: UMAP. O.: organoid. E.: embryo.
   Also see Figure S8.
@@
 Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neural progenitors 
-in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
+in the embryo. \par
+
+To get insight into the minimal set of TFBS that are needed to encode these cell states we trained
 logistic regression models to classify genomic regions as belonging to the topic representing each state or not,
@@
 models classifying genomic regions to each topic using motif scores as features (top) and patterns with positive
-coefficient for each logistic regression model along with the number of each patterns used for enhancer design
-represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
-each topic (bottom). Cell type-classes are indicated by colors. 
+coefficient for each logistic regression model along with the number of each patternsused for enhancer design
+represented by the size of each dot (middle)
 (Continues next page)}
@@
   (Continued) 
+   and boxplots (n = 200) showing prediction score of designed enhancers for each topic (bottom). Cell type-classes are indicated by colors. 
   \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
@@
 that enhancers consist of clusters of 4-5 different TFBS. 
-Homotypic clusters of 3-4 homeobox instances were sufficient to both create and identify enhancers specific to a certain AP domain. 
+Homotypic clusters of 3-4 homeobox instances were sufficient to design enhancers specific to a certain AP domain. 
 Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. 
@@
 interpret and design enhancers underlying early human neural and neural crest development.
+Alongside with these models, we implemented new tools in our genomic deep learning framework CREsted\cite{kempynckCREstedModelingGenomic2025}, specifically for the analysis of enhancer codes underlying complex biological systems.
 
@@
 (reviewer token: ). %TODO: add before submission.
-Raw sequencing data of the human embryo are available through the European Genome Phenome Archive 
-(). %TODO: Add link to EGA
+Raw sequencing data of the human embryo will be made available through the European Genome Phenome Archive. %TODO: Add link to EGA
 scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks and cell-and region topic
 probabilities of the human embryos are available through Zenodo (\href{https://doi.org/10.5281/zenodo.15281826}{10.5281/zenodo.15281826}). 
-$DeepNeuralTube_O$ and $DeepNeuralTube_E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
+$DeepNeuralTube_O$ and $DeepNeural$- $Tube_E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
 Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the UCSC genome browser: 
@@
 \end{itemize}
-The enhancer compendium generated in this study is available at something.aertslab.org. %TODO: generate landing page
+
+Code and tutorials to perform enhancer code analysis as presented in this manuscript is available in \href{https://github.com/iaertslab/crested}{CREsted}, see \href{https://crested.readthedocs.io/en/stable/}{crested.readthedocs.io}.
 
@@
 \end{description}
-
-\newpage
-
-\section*{MAIN FIGURE TITLES AND LEGENDS}
-
-%%%  At final submission, figure files MUST be 
-%%%  provided separately as high-resolution image 
-%%%  files. All of the panels for a figure should 
-%%%  be in the same file. Figures should have 
-%%%  clear labels/file names (Figure 1, Figure 2, 
-%%%  etc.).
-
-%%%  Figure titles and legends should be placed 
-%%%  at the end of the main text. You do not 
-%%%  need to place the figures, nor their titles 
-%%%  and legends, within the main text. While 
-%%%  typesetting your article, our team will 
-%%%  place each figure in the best location 
-%%%  based on the final layout and on your 
-%%%  figure citations, e.g., (Figures 1A and 1B).
-
-%%%  Please review the figure guidelines before 
-%%%  submitting your final materials: 
-%%%  https://www.cell.com/figureguidelines.
-
-\newpage
-
-\section*{MAIN TABLES, INCLUDING TITLES AND LEGENDS}
-
-%%%  Whenever possible, we encourage you to submit 
-%%%  your main-text tables as Microsoft Word documents, 
-%%%  using Word's table function. This will ensure 
-%%%  the best results during conversion. Tables 
-%%%  should be numbered Table 1, Table 2, etc. and 
-%%%  should not include subpanels (do not use Table 1A, 
-%%%  1B, etc.). Give each table a brief descriptive 
-%%%  title. Table legends are optional but encouraged. 
-%%%  Footnotes (superscript lowercase letters) should 
-%%%  be used where necessary to indicate some feature 
-%%%  of the data; please do not use bold, italic, 
-%%%  colored text, or shading for this purpose. Use 
-%%%  separate cells, not line breaks or spaces, for 
-%%%  all discrete data elements. Small embedded 
-%%%  graphics with color are OK.
-
-%%%  Like figures, all tables must be cited within 
-%%%  the main text, and our typesetting team will 
-%%%  place the tables within the typeset paper at 
-%%%  the appropriate locations.
-
-%\subsection*{Table 1. A table with clear organization of data}
-
-%\begin{tabular}{|l | l | l | l|} 
-% \hline
-% \textbf{Column 1} & \textbf{Column 2} & \textbf{Column 3} & \textbf{Column 4} \\ [1ex] 
-% \hline
-% Row A\textsuperscript{a} & 6 & 87,837 & 787 \\ [1ex] 
-% \hline
-% Row B & 7 & 78 & 5,415 \\ [1ex] 
-% \hline
-% Row C & 545 & 778 & 7507 \\ [1ex] 
-% \hline
-% Row D & 545 & 18,744 & 7,560 \\ [1ex] 
-% \hline
-% Row E & 88 & 788 & 6,344 \\ [1ex] 
-% \hline
-%\end{tabular}
-
-\bigskip
-
-%The table legend (optional) follows the table itself. The legend should be used to provide additional info that relates to the table as a whole.
-
-%textsuperscript{a}Footnotes can be used to provide additional info on specific content within the table, such as this footnote to the first row (row A). Do not use footnotes in the table title.
-
-%\textsuperscript{b}More footnotes
-%
-%NOTE: For the journal Cell, supplemental figures should instead be uploaded as separate files, the same process as the main figures; their titles and legends should be placed in the main text, directly after the main figure legends and any tables, under a “supplemental figures” heading
 
@@
 The fragment size of the libraries were analyzed using the Bioanalyzer high-sensitivity chip. 
-All embryonal libraries were first sequenced on the NextSeq2000 instrument (Illumina) before being sequenced deeper on 
+All embryonal libraries were first sequenced on the NextSeq- 2000 instrument (Illumina) before being sequenced deeper on 
 the NovaSeq6000 instrument (lllumina). 
@@
 \begin{itemize}
-  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix (IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG).
-  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG).
-  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix (6-7\_1-2-3\_1 and 6-7\_1-2-3\_2).
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{1.5}{\percent} PEG hydrogel matrix:\\ IPS\_\_081787\_\_575955\_\_10x\_multiome\_neural\_tube\_organoids\_1\_5\_PEG.
+  \item A single library containing a mix of organoids at 1, 3, 5, 7, 9 and 11 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix:\\ IPS\_\_c13535\_\_a58829\_\_10x\_multiome\_neural\_tube\_organoids\_2\_PEG.
+  \item Two libraries containing a mix of organoids at 1, 2, 3, 6 and 7 days of differentiation cultured in \SI{2}{\percent} PEG hydrogel matrix:\\ 6-7\_1-2-3\_1 and 6-7\_1-2-3\_2.
   \item A single library containing a mix of organoids at 6 and 7 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix (6-7).
+    PEG hydrogel matrix:\\ 6-7.
   \item A single library containing a mix of organoids at 4, 5, 8 and 9 days of differentiation cultured in
-    \SI{2}{\percent} PEG hydrogel matrix (8-9\_4-5).
+    \SI{2}{\percent} PEG hydrogel matrix:\\ 8-9\_4-5.
   \item A single library containing a mix of organoids at 8 and 9 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix (8-9).
+    PEG hydrogel matrix:\\ 8-9.
   \item A single library containing a mix of organoids at 10 and 11 days of differentiation cultured in \SI{2}{\percent}
-    PEG hydrogel matrix (10-11).
+    PEG hydrogel matrix:\\ 10-11.
 \end{itemize}
@@
 cell types (see single-cell RNA cell type annotation) using the export\_pseudobulk function, with default parameters, from pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023}
-(v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618}) followed by peak calling using MACS\cite{zhangModelbasedAnalysisChIPSeq2008} (v2.2.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_013291/resolver}{RRID:SCR\_013291} ) using the pycisTopic function peak\_calling, using default parameters, and consensus peak generation using the pycisTopic function get\_consensus\_peaks, using a peak width of 500 bp and default parameters.
+(v1.0.3.dev20+g8955c76, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID:SCR\_026618}) followed by peak calling using MACS\cite{zhangModelbasedAnalysisChIPSeq2008} (v2.2.9.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_013291/resolver}{RRID:SCR\_013291} ) using the pycisTopic function peak\_calling, using default parameters, and consensus peak generation using the pycisTopic function get\_consensus\_peaks, using a peak width of 500 bp and default parameters.
 This resulted in 488,211 consensus peaks.
@@
 Sample and barcode level quality control data was calculated using the compute\_qc\_stats function from
-pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{}{RRID:SCR\_026618})
+pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/data/record/nlx_144509-1/SCR_026618/resolver}{RRID- :SCR\_026618})
 cells were filtered based on the number of unique fragments per cell, Fraction of Reads in Peaks (FRIP) and
@@
 the subset of cells annotated as neuron (Neuron cells and neural progenitor early neuron cells) and one on the subset of cells annotated as neural crest.
-Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; RRID:SCR\_026708) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/datmccallumMALLETMachineLearning2002618/resolver}{RRID:SCR\_026618}) function run\_cgs\_models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
+Topic modeling was performed using Mallet\cite{mccallumMALLETMachineLearning2002} (v202108; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026708/resolver}{RRID- :SCR\_026708}) using the pycisTopic\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0.3.dev20+g8955c76, \href{https://rrid.site/datmccallumMALLETMachineLearning2002618/resolver}{RRID:SCR\_026618}) function run\_cgs\_- models\_mallet using 500 iterations, alpha equal to 50, eta equal to 0.1, alpha\_by\_topic equal to True and eta\_by\_topic equal to False and using random seed 555.
 Topic modeling was performed separately for all cells and for each subset of cells.
@@
 A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
-CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
+CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:  SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
 1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
@@
 A deepTopic model\cite{minnoyeCrossspeciesAnalysisEnhancer2020}, now implemented in
-CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
+CREsted\cite{kempynckCREstedModelingGenomic2025} (\href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID: SCR\_026617}), containing 5 convolutional and max pooling layers: dimensions (500,
 1,024), (125, 512), (32, 512), (8, 512) and (2, 512) a fully connected dense layer of dimension 1,024 and a dense output
@@
 
-A cistarget motif ranking database was created by first getting the genomic sequence of each organoid consensus peak with 1,000 bp background padding using the create\_fasta\_with\_padded\_bg\_from\_bed script from the create\_cisTarget\_databases repository (\href{https://github.com/aertslab/create_cisTarget_databases}{https://github.com/aertslab/create\_cisTarget\_databases}) and scoring each DNA sequence using Cluster-Buster\cite{frithClusterBusterFindingDense2003}(\href{https://rrid.site/data/record/nlx_144509-1/SCR_024810/resolver?}{RRID:SCR\_024810}) using the create\_cisTarget\_databases command create\_cistarget\_motif\_databases followed by the generation of the ranking databases using the command convert\_motifs\_or\_tracks\_vs\_regions\_or\_genes\_scores\_to\_rankings\_cistarget\_dbs.
-Next, the SCENIC+\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0a2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026702/resolver?}{RRID:SCR\_026702}) command scenicplus grn\_inference motif\_enrichment\_cistarget command was used to perform cisTarget\cite{potierUsingCisTargetXPredict2012} motif enrichment analysis for each organoid and embryo topic.
+A cistarget motif ranking database was created by first getting the genomic sequence of each organoid consensus peak with 1,000 bp background padding using the create\_fasta\_with\_padded- \_bg\_from\_bed script from the create\_cisTarget\_databases repository (\href{https://github.com/aertslab/create_cisTarget_databases}{https://github.com/aertslab/  create\_cisTarget\_databases}) and scoring each DNA sequence using Cluster-Buster\cite{frithClusterBusterFindingDense2003}(\href{https://rrid.site/data/record/nlx_144509-1/SCR_024810/resolver?}{RRID: SCR\_024810}) using the create\_cisTarget\_databases command create\_cistarget\_motif\_databases followed by the generation of the ranking databases using the command convert\_motifs\_or\_tracks- \_vs\_regions\_or\_genes\_scores\_to\_rankings\_cistarget\_dbs.
+Next, the SCENIC+\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023} (v1.0a2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026702/resolver?}{RRID: SCR\_026702}) command scenicplus grn\_inference motif- \_enrichment\_cistarget command was used to perform cisTarget\cite{potierUsingCisTargetXPredict2012} motif enrichment analysis for each organoid and embryo topic.
 
@@
 min\_ic\_windowsize = 6 and 
-ppm\_pseudocount= 0.001.
+ppm- \_pseudocount= 0.001.
 Next, similar patterns were clustered together by first calculating a pairwise pattern-to-pattern similarity metric
-using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}).
-The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID:SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID:SCR\_008058}).
+using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID: SCR\_026620}).
+The Pearson correlation of one minus the E-value (product of the TomTom p-value and the number of patterns), calculated using the corrcoef function from NumPy\cite{harrisArrayProgrammingNumPy2020} (v2.0.1; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008633/resolver}{RRID: SCR\_008633}), was used for hierarchical clustering using the hierarchy.linkage (using average as method) and fcluster (using maxclust as criterion and a manual defined number of clusters) functions from Scipy\cite{virtanenSciPy10Fundamental2020} (v1.15.2; \href{https://rrid.site/data/record/nlx_144509-1/SCR_008058/resolver}{RRID: SCR\_008058}).
 
@@
 A pairwise similarity matrix between all enriched motifs (see Motif enrichment analysis) and \textit{de novo} motifs was
-calculated using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}).
-The Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.11.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID:SCR\_018139}) functions pca, neighbors, tsne and leiden were used to generate a tSNE-embedding of the motifs based on 50 principle components and to cluster the motifs at a leiden resolution of 2.
+calculated using TomTom\cite{guptaQuantifyingSimilarityMotifs2007} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID: SCR\_026620}).
+The Scanpy\cite{wolfSCANPYLargescaleSinglecell2018} (v1.11.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_018139/resolver}{RRID: SCR\_018139}) functions pca, neighbors, tsne and leiden were used to generate a tSNE-embedding of the motifs based on 50 principle components and to cluster the motifs at a leiden resolution of 2.
 
@@
 
-Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID:SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
+Enhancer candidates were ordered from Twist Bioscience (\href{https://rrid.site/data/record/nlx_144509-1/SCR_025817/resolver}{RRID: SCR\_025817}) with 5\lq 5\lq-GGTACCGAGCTCGAG-3\lq and 3\lq 5\lq-GGGCTCGAGATCTGC-3\lq adaptors.
 Plasmid pTK BsmBI LacZ Citrine Nanotag 24 (\href{https://rrid.site/data/record/nif-0000-11872-1/Addgene_130522/resolver}{RRID:Addgene\_130522})\cite{williamsReconstructionGlobalNeural2019} was linarized using forward primer 5\lq-ggAACTCGAGCTCGGTACCT-3\lq and reverse primer 5\lq-GGGCTCGAGATCTGCGATC-3\lq using PCR.
@@
 \subsubsection*{Anterior-posterior synthetic enhancer design}
-5\lq-GGATTAG-3\lq and 5\lq-CTAATTAG-3\lq were used as motifs in the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID:SCR\_026617}) function enhancer\_design\_motif\_insertion to design 200 synthetic enhancers of length 500 for both topic 31 and 61 using up to 5 motif instances.
+5\lq-GGATTAG-3\lq and 5\lq-CTAATTAG-3\lq were used as motifs in the CREsted\cite{kempynckCREstedModelingGenomic2025} (v1.3.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026617/resolver}{RRID: SCR\_ 026617}) function enhancer\_design\_motif\_insertion to design 200 synthetic enhancers of length 500 for both topic 31 and 61 using up to 5 motif instances.
 
@@
 Motif scores for all patterns (convered to a PWM) belonging to cluster 4 or 8 of anterior-posterior \textit{de novo} patterns hits
-were calculated on the top, based on region-topic probabilities, 3,000 regions for topic 31 and topic 61 using  FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID:SCR\_026620}) using threshold parameter set to 0.5 to make sure each region gets a score for at least one pattern per cluster.
+were calculated on the top, based on region-topic probabilities, 3,000 regions for topic 31 and topic 61 using  FIMO\cite{grantFIMOScanningOccurrences2011} implemented in tangermeme (v0.4.0; \href{https://rrid.site/data/record/nlx_144509-1/SCR_026620/resolver}{RRID: SCR\_026620}) using threshold parameter set to 0.5 to make sure each region gets a score for at least one pattern per cluster.
 For each sequence and cluster the maximum score, based on the $-log_{10}(p-value)$ of the hit, was retained.
@@
 \subsubsection*{logistic regression models for each cell states}
-For each cell type, neural progenitors dorso-ventral patterning, neural progenitors anterior-posterior patterning, neurons,
+For each cell type, neural progenitors dorso-ventral patterning, neural progenitors anterior- posterior patterning, neurons,
 and neural crest, motif scores for all patterns (converted to a PWM) patterns belonging to clusters of the cell type\lq s
@@
     variable genes) across cell types and organoid and embryo. Number top left indicates Pearson correlation
-    coefficient. \textbf{d}, d histogram showing distribution of Pearson correlation coefficients of average gene
+    coefficient. \textbf{d}, histogram showing distribution of Pearson correlation coefficients of average gene
     expression profile per topic, topics representing
@@
     \textbf{Validation of DeepNeuralTube and profiles of selected enhancers}\\
-    \textbf{a-b}, precision recall curves for neural tube enhancers (left) and facial mesenchyme enhancers (right) for
-    organoid (\textbf{a}) and embryo (\textbf{b}). \textbf{c}, Chromatin accessibility profile of dorso-ventral topics for organoid
+    \textbf{a-b}, precision recall curves for classifying neural tube enhancers (left) and facial mesenchyme enhancers (right) using organoid (\textbf{a}) and embryo models (\textbf{b}).
+\textbf{c}, Chromatin accessibility profile of dorso-ventral topics for organoid
     (top) and embryo (bottom) of region chr7:42,058,384-42,246,450; hg38. Location of HS111 enhancer is indicated. Values are
@@
     $DeepNeuralTube_E$ (bottom) for each topic for neural crest enhancer chr1:161200798-161201298; hg38.
+    Org. Organoid. Emb embryo.
     Related to Figure 2.
@@
     \textbf{c}, Line shart, showing $-log_{10}(p value)$ of FOXA2 motif hits calculated using FIMO\cite{grantFIMOScanningOccurrences2011,schreiberTangermemeBiologicalSequence2025} for general regions and specific regions stratified by the number of FOXA2 instances per regions.
-* indicates two-sided p-value (Mann-Whitney U test) < 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
+* indicates two-sided p-value (Mann-Whitney U test) $<$ 1e-6 calculated between the distribution of motif scores for specific and general regions stratified by the number of FOX instances per region.
     Related to Figure 4.

### 2025-05-22 manuscript.tex
Changes S.A.

@@
 
-By encoding clusters of transcription factor (TF) binding sites, genomic enhancers act as hubs for combinatorial TF
-recruitment. 
-Enhancer code studies have been limited to specific cell types in model organisms. 
-Here we set out to create a compendium of enhancer codes for all cell types of the developing human neural tube and neural crest. 
-To this end, we generated a single-cell multiome atlas of human neural tube organoids and of the head of a four weeks post conception human embryo. 
-Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared between the organoid 
-model and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
-We confirm the paradigm that enhancers are encoded using heterotypic clusters of 4-5 TF binding sites in a nucleosome depleted region, using soft syntactic rules. 
-We also provide counter examples to this rule, with enhancers encoded as homotypic clusters and by dimer motifs. 
-Finally, we validate DeepNeuralTube using enhancer reporter assays and assess the role of human genetic variants on 
-facial mesenchyme enhancers. 
-We provide tools for building enhancer code compendia through our genomic deep learning package, CREsted.
+Genomic enhancer code studies have been limited to specific cell types in model organisms. 
+We created a compendium of enhancer codes for all cell types of the human neural tube and neural crest. 
+We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week-post conception 
+human embryo. Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared 
+between the organoid and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
+We confirm the paradigm that enhancers are encoded as heterotypic clusters of 4-5 TF binding sites within one 
+nucleosome depleted region, using soft syntactic rules. We also provide counter examples to this rule, with enhancers 
+encoded as homotypic clusters and by dimer motifs. Finally, we validate DeepNeuralTube using enhancer reporter assays 
+and by interpreting genetic variation associated with facial features.
 
@@
 In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
-a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}.
-We compare this compendium to that of the head of a four weeks post conception human embryo.
-Through these comparisons we assess the enhancer code paradigm.
-\par
+a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
+in direct comparison to an enhancer compendium we generate in parallel for the head of a four-week post conception human 
+embryo. Through these comparisons we assess the enhancer code paradigm and construct an enhancer annotation and code 
+resource that is a consensus between the organoid model and the human embryo.\par
 
@@
 
-\subsection*{Human neural tube organoids recapitulate early neural and neural crest development}
-
-We generated human neural tube organoids to model early human neural and neural crest development. 
+\subsection*{Human neural tube organoids recapitulate early neuronal and neural crest development}
+
+We generated human neural tube organoids to model early human neuronal and neural crest development. 
 To this end we embedded single cells from two induced pluripotent stem cell (iPSC) lines (Sigma and BJ1) and one embryonic stem cell line (ESC; H9) into separate synthetic poly-ethylene glycol (PEG) based extracellular matrices.
@@
 To evaluate the biological relevance of the cell types and states obtained from the organoid, 
-we also performed 10x single cell multiomics on the head of a single four weeks post conception human embryo, 
+we also performed 10x single cell multiomics on the head of a single four-weeks post conception human embryo, 
 resulting in 25,785 high quality cells (\textbf{Fig. 1c}). 
@@
 Next, we sub-clustered the neural progenitor, early neuron, and neural crest cells (\textbf{Fig. 1d-f}). 
-Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorso-ventral patterning, evident from the expression of dorso-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
+Both organoid and embryo neural progenitors show a gradient of different cell states reflecting dorsal-ventral patterning, evident from the expression of dorsal-ventral marker TFs\cite{delasRepressiveInteractionsGene2020}:
 \textit{ARX}, \textit{NKX2-2}, \textit{PAX6}, \textit{IRX3}, \textit{ZIC1} and \textit{GDF7} (\textbf{Fig. 1d}). 
@@
 
-Scoring cells of both organoid and embryo origin for the expression of cell cylce marker genes, 
+Scoring cells of both organoid and embryo origin for the expression of cell cycle marker genes, 
 we observed that both early differentiating peripheral neurons and to a greater extent early differentiating neurons of the central nervous system are in cell cycle arrest (\textbf{Fig.
-1h}), an important hallmark of neural differentiating\cite{
+1h}), an important hallmark of differentiating neurons\cite{
 braunComprehensiveCellAtlas2023}, while other cell types are still in different phases of the cell cycle (\textbf{Fig. 1h})\par
@@
   \caption{
-    \textbf{Human neural tube organoids recapitulate early neural and neural crest development.} \\
+    \textbf{Human neural tube organoids recapitulate early neuronal and neural crest development.} \\
     \textbf{a}, Schematic of organoid differentiating protocol and human embryo experiment 
@@
 \subsection*{
-  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neural and neural crest 
+  DeepNeuralTube: sequence-to-function models that reveal the enhancer code underlying early neuronal and neural crest 
   development
@@
 The classes of the models are defined as the topics from topic modeling. 
-To get a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl.
+To obtain a fine-grained definition of cell states, we performed topic modeling separately for the subset of neural progenitors, neural crest (incl.
 facial mesenchyme and peripheral neurons) and early differentiating neurons for both organoid (to train $DeepNeuralTube^o$) and embryo (to train $DeepNeuralTube^e$).\par
@@
 and auPR of 0.17 for both models (\textbf{Fig. S2a-b})). 
-Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) are predicted to be specific to pre-migratory neural crest by
-both models. Given that the VISTA enhancers are annotated based on localization of enhancer activity only, and given
-that it is difficult to visually distinguish between the neural tube and pre-migratory neural crest, this result 
-is not surprising.\par
+Of note, a fraction of neural tube enhancers (\SI{24}{\percent}) is predicted to be specific to pre-migratory neural crest by
+both models. Given that the VISTA enhancers are annotated based on spatial localization of enhancer activity by 
+\textit{in situ} hybridization, and given that it is difficult to visually distinguish between the neural tube and pre-
+migratory neural crest, this result is not surprising.\par
 
@@
 More importantly, most clusters (\SI{70}{\percent}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model (\textbf{Fig. 2f-g} and \textbf{Table S2}). 
-Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ use known-motifs as features to classify regions based
+Thus, $DeepNeuralTube^o$ and $DeepNeuralTube^e$ learn to use previously known motifs as features to classify regions based
 on cell type-specific chromatin accessibility.\par
@@
 Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
-and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
-Next, we tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
+and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and TFAP2 (\textbf{Fig. 2i}). 
+We tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
 Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par
@@
 \subsection*{
-  The enhancer code underlying dorso-ventral patterning of neural progenitors
-}
-
-Focusing on neural progenitors, we identified topics representing cells and regions of different dorso-ventral
+  The enhancer code underlying dorsal-ventral patterning of neural progenitors
+}
+
+Focusing on neural progenitors, we identified topics representing cells and regions of different dorsal-ventral
 identities in both the organoid and the embryo data (\textbf{Fig. 3a}). 
-From ventral to dorsal this is topic 33, 38, 36, 54 and 48 for the organoids and
-topic 34, 38, 79, 88 and 58 for the embryo. 
-As an example we first investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
+From ventral to dorsal these are topics 33, 38, 36, 54 and 48 for the organoids and
+topics 34, 38, 79, 88 and 58 for the embryo. 
+As an example, we investigated two enhancers of \textit{SHH} that are strongly conserved and specific to the 
 floorplate: SFPE1 and SFPE2\cite{jeongDistinctRegulatorsShh2003}. 
 \textit{SHH} is most strongly expressed in the cells of the most ventral topic (33 and 34 resp. for the organoids and the embryo; \textbf{Fig. 3b}).
-SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed most accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
+SFPE1 is located upstream of \textit{SHH} and SFPE2 in its second intron. Both enhancers are indeed differentially  accessible in the cells belonging to the ventral topic (\textbf{Fig. 3c}). 
 When we tested the enhancer activity of SFPE1 in chicken embryos we could confirm its floor plate specific activity
@@
 
-To investigate the enhancer code underlying dorso-ventral patterning of neural progenitors further, we clustered 
-and aligned \textit{de novo} motifs, identified in dorso-ventral regions, based on pairwise motif similarities. 
+To investigate the enhancer code underlying dorsal-ventral patterning of neural progenitors further, we clustered 
+and aligned \textit{de novo} motifs, identified in dorsal-ventral regions, based on pairwise motif similarities. 
 Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
-resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3f}).
-Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorso-ventral states and the 
+resulting in a dorsal-ventral enhancer code-table (\textbf{Fig. 3f}).
+Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorsal-ventral states and the 
 importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
@@
 ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
-For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
+For all dorsal-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
 SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
-Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
+Similarly, ZEB instances have a negative importance to the accessibility of regions in dorsal-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
 Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig.
@@
 
-To assess the ability of DeepNeuralTube for prioritizing enhancer with cell type-specific activity we selected a genomic
-region that was predicited to be specific for ventral neural progenitors (\textbf{Fig. 3k}). Explaining this region,
+To assess the ability of DeepNeuralTube for prioritizing enhancers with cell type-specific activity we selected a genomic
+region that was predicted to be specific for ventral neural progenitors (\textbf{Fig. 3k}). Explaining this region,
 reveals multiple FOX instances and a SOX instance by both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ (\textbf{Fig. 3j}). 
@@
 For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
-We classified the regions that overlap with any of the ChIP-seq datasets as regions that are generally bound by FOXA2 and those that did not as floorplate specific.
-Quantifying both the number of FOX instances per region and
-the affinity of each individual instance (making use of the FOX motif score calculated using
-FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) reveals that floorplate 
-specific FOXA2 target regions have fewer FOX instances compared to general FOXA2 target regions (\textbf{Fig. S4b}). 
-Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOX instances in
-specific regions (\textbf{Fig. S4c}). These results suggest that the overall affinity of general FOXA2 target regions 
-is higher (either by having a greater number of binding sites or by having binding sites with higher affinity).
+In the FOXA2 target enhancers of the floorplate that are also bound by FOXA2 in other systems (HepG2 and A549 cell lines), 
+we identified a higher number of FOXA2 sites using FIMO\cite{
+grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025
+} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
+Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOXA2 instances 
+in specific regions (\textbf{Fig. S4c}). This suggests that FOXA2 by itself can bind to a subset of genomic regions 
+independently of the cell type (i.e., indepdent of cooperating TFS), but mainly to regions with multiple sites, and 
+high-affinity sites.
 
@@
   \caption{
-    \textbf{Deciphering the enhancer code underlying dorso-ventral patterning of neural progenitors.} \\
-    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neural progenitors colored by cell-topic probabilities in RGB color scales of dorso-ventral topics. 
+    \textbf{Deciphering the enhancer code underlying dorsal-ventral patterning of neural progenitors.} \\
+    \textbf{a}, UMAP of organoid (top) and embryo (bottom) neural progenitors colored by cell-topic probabilities in RGB color scales of dorsal-ventral topics. 
     (Continues next page).
@@
     (Continued) 
-    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorso-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
+    \textbf{b}, Boxplots of the expression of \textit{SHH} in organoid (top) and embryo (bottom) across different dorsal-  ventral topics ($n_{topic 33}$ = 4,571, $n_{topic 38}$ = 4,521, $n_{topic 36}$ = 6,120, $n_{topic 54}$ = 2,808, and $n_{topic 48}$ = 5,390 for organoid cells;
       and $n_{topic 34}$ = 637, $n_{topic 38}$ = 1,093, $n_{topic 79}$ = 3,939, $n_{topic 88}$ = 1,861, and $n_{topic 58}$ = 1,773 for embryo cells).  With n representing the number of cells.
-    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorso-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
+    \textbf{c}, Chromatin accessibility in organoid (top) and embryo (bottom) across dorsal-ventral topics for locus chr7:155,799,664-155,827,483 ; hg38. SFPE2 and SFPE1 are highlighted. 
     \textbf{d}, Enhancer reporter assay in chicken embryo. Blue is DAPI, pink is constitutive electroportion control,
@@
     \textbf{e}, Nucleotide contribution scores for SFPE2 (left) and SFPE1 (right) for organoid topic 33 (top) and embryo topic 34 (bottom). 
-    \textbf{f}, Code table showing average contribution score of pattern instances for each dorso-ventral topic.
+    \textbf{f}, Code table showing average contribution score of pattern instances for each dorsal-ventral topic.
 Contribution scores for $DeepNeuralTube^o$ are not outlined.
 Contribution scores for $DeepNeuralTube^e$ are outlined and negated.
-    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorso-ventral topics. 
-    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorso-ventral topics. 
+    \textbf{g}, Heatmap showing the maximum normalized contribution score for each of the top 1,000 regions per topic (y-axis) and pattern (x-axis) for dorsal-ventral topics. 
+    \textbf{h}, Heatmap showing scaled expression of transcription factors corresponding to identified patterns in organoid (left) and embryo (right) dorsal-ventral topics. 
     \textbf{i}, Heatmap showing Jaccard index of regions based on the presence of instances of the identified patterns. 
@@
 We first asked if the AP-program is distinct from the DV program. 
-In other words, is AP patterning encoded in the same or in different genomic regions as DV patterning? 
+In other words, is AP patterning encoded by the same or different genomic regions as DV patterning? 
 For this purpose, we calculated the pairwise correlation between region-topic probabilities of all AP and DV topics (\textbf{Fig. 4b}). 
@@
 with the exception of AP Topic 59 and 31 that correlate with the region-topic probability of DV Topic 38 
-(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b})\par
+(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b}).\par
 
@@
 cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
-Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} 
-and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
+Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2}; 
+and the presence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
 In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
@@
 4h}), even though the prediction score was optimized towards the opposite class.\par
+%TODO: S.A.: Is this really needed here?
 
@@
 anterior-posterior topics. 
-  \textbf{b}, Heatmap showing pairwise Pearson correlation coefficients of region-topic probabilities for dorso-ventral and anterior-posterior topics. 
+  \textbf{b}, Heatmap showing pairwise Pearson correlation coefficients of region-topic probabilities for dorsal-ventral and anterior-posterior topics. 
   \textbf{c}, Heatmap showing prediction scores of $DeepNeuralTube^e$ on the topic 3,000 regions per topic (y-axis) for each anterior-posterior topics (x-axis). 
@@
   \textbf{i}, receiver-operator curve (ROC) (top) and precision-recall curve (bottom) for logistic regression classifier classifying genomic regions as either Topic 61 or Topic 31 based on motif scores for \textit{CTAATTAG} and \textit{CTAATCCCT} motifs.
-  U.: UMAP. AP: anterior-posterior, DV: dorso-ventral, iter: iteration. 
+  U.: UMAP. AP: anterior-posterior, DV: dorsal-ventral, iter: iteration. 
   Also see Figure S5.
@@
 Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
-instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) 
-the importance of TEAD instances correlates with \textit{TEAD3/4} expression, 
-GRHL with \textit{GRHL1/1} 
+instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}). 
+The importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} 
 and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Fig. S6). 
-Note that ZEB instances have a positive contribution to the pre-migratory while it has a 
+Note that ZEB instances have a positive contribution to the pre-migratory while they have a
 negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
-and moreover a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
+and moreover, a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
 The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Fig. S6). 
-This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
+This is in line with the previously reported role of ZEB as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par
 
@@
 Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). 
-We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. 
+We could not link the nuclear receptor instances to a specific TF based on their correlation with TF expression. 
 The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), 
@@
 Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. 
-Instances of a FOX patterns and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). 
+Instances of a FOX patterns and a dimer consisting of a homeobox and EBOX motif are important for this state (\textbf{Fig. 5b}). 
 The importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and 
 that of the dimer motif with the expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Fig. S6). 
-The latter is consistent with a recent study where it has been shown that TWIST1 and homeodomain TFs cooperatively
+The latter is consistent with a recent study that showed that TWIST1 and homeodomain TFs cooperatively
 bind this specific dimer motif regulating the expression of genes involved in
@@
 index of genomic regions based on the presence of an instance of each identified pattern (\textbf{Fig. 5e}). Using this
-method we found three sets of regions that are predicted to contain instances of different sets of TFs. One set where
+method, we found three sets of regions that are predicted to contain instances of different sets of TFs. One set where
 instances of TEAD, GRHL and ZEB co-occur, one where instances of Nuclear Receptor, AP2 and ZIC
@@
 Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
-we asked if, using $DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity.
+we asked whether $DeepNeuralTube$ could explain the effect of some of these variants on enhancer activity.
 For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$
@@
 Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Fig. S7}). 
-As an example we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
+As an example, we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
 The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
@@
 \subsection*{
-  The enhancer code underlying early neural differentiation.
-}
-
-For the cells representing early neural differentiation, we identified several topics for different stages of neural fate
-acquisition in both organoid and embryo (\textbf{Fig. 6a}). Again we made use of $DeepNeuralTube^o$ and
+  The enhancer code underlying early neuronal differentiation.
+}
+
+For the cells representing early neuronal differentiation, we identified several topics for different stages of neuronal fate
+acquisition in both organoid and embryo (\textbf{Fig. 6a}). Again, we made use of $DeepNeuralTube^o$ and
 $DeepNeuralTube^e$ to generate a neuron code-table (\textbf{Fig. 6b}). 
@@
   \captionsetup{justification=raggedright}
-  \caption{ \textbf{Deciphering the enhancer code of early neural differentiation.} \\
+  \caption{ \textbf{Deciphering the enhancer code of early neuronal differentiation.} \\
   \textbf{a}, UMAP of organoid (top two) and embryo (bottom two) differentiating neuron cells colored by cell-topic probabilities in RGB color scales for neuron related topics. 
@@
 Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in 
-respectively the neural tube organoids and the head of a four weeks post conception human embryo with 
+respectively the neural tube organoids and the head of a four-weeks post conception human embryo with 
 almost one-to-one correspondence based on TFBS importances and TF expression. 
@@
 
-These results indicate that enhancers specific to DV-patterning, neural crest states and early neural differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
+These results indicate that enhancers specific to DV-patterning, neural crest states and early neuronal differentiation are encoded in the human genome using around four to five distinct TFBS (\textbf{Fig.
 7a})
@@
 
-We built a compendium of enhancer codes underlying early human neural and neural crest development. 
+We built a compendium of enhancer codes underlying early human neuronal and neural crest development. 
 Providing enhancer codes of neural progenitor DV- and AP-patterning, 
@@
 To create this compendium, we generated a single-cell multiome (combined scATAC-seq and scRNA-seq of the same cell) 
-dataset of the head of a four weeks post conception human embryo and we showed that the embryonic cell states can 
-effectively be reproduced using human neural tube organoids on both transcriptomic level and on the level of the 
+dataset of the head of a four-weeks post conception human embryo and we showed that the embryonic cell states can 
+effectively be reproduced using human neural tube organoids both at transcriptomic level and at the level of 
 enhancer codes. Specifically, we identified a total of fifteen distinct cell states with one-to-one matches between 
-the organoid and embryo. The fact that cell states were defined in a data driven manner by making use of topic 
+the organoid and the embryo. The fact that cell states were defined in a data driven manner by making use of topic 
 modeling and that each of these states were independently identified in two separate datasets increases the confidence 
-of their biological relevance. Furthermore, this high similarity between the organoid and the embyro derived cell 
+of their biological relevance. Furthermore, this high similarity between organoid and embyro derived cell 
 states validates neural tube organoids as a biological model for early human neural tube and neural crest development, 
@@
 Nonetheless, there are recurrent patterns that are shared among enhancers of different cell types. 
-In particular, enhancers underlying DV patterning, early neural fate acquisition and neural crest 
+In particular, enhancers underlying DV patterning, early neuronal fate acquisition and neural crest 
 tend to consist of heterotypic clusters of 4-5 different TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located within one NDR (+/-147bp). 
 That is to say, cooperative binding of multiple TFs underlies the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. 
-Here, we see most evidence for a model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer model\cite{arnostiTranscriptionalEnhancersIntelligent2005}).
+Here, we see most evidence for a model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (
+i.e., "soft syntax"\cite{zeitlingerSevenMythsHow2020, dewinterModellingDesignTranscriptional2025} or flexible billboard 
+enhancer model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). 
 However, for multiple cell types we also observed evidence of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. 
@@
 and SOX10 homodimers in neural crest. 
-For all these TFs we found dimer motifs with a fixed number of nucleotides in between the two TFBS instances. 
-Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bp spacing,
-with instances with 4 bp spacing being the most accessible (\textbf{Fig. 10}). 
+For all these TFs we found dimer motifs with a fixed length of the spacer between two TFBS instances. 
+Of note, for SOX10 we observed three different dimer motifs with either 3, 4 or 5 bp spacing,
+whereby, instances with a 4 bp spacer were most accessible (\textbf{Fig. 10}). 
 Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the 
@@
 dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. 
-Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites 
+Accordingly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites 
 and/or binding sites of lower affinity. 
@@
 %TODO: S.A.: But retain strongest footprint ?!
-Our results indicate that enhancers underlying neural progenitor AP-patterning form an exception to the rule 
-that enhancers consist of clusters of 4-5 different TFBS. 
-Homotypic clusters of 3-4 homeobox instances were sufficient to design enhancers specific to a certain AP domain. 
+Our results indicate that enhancers underlying neural progenitor AP-patterning form an exception to the heterotypic rule 
+of 4-5 different TFBS per enhancer. 
+Homotypic clusters of 3-4 homeobox instances were sufficient to (\textit{in silico}) design enhancers specific to a certain AP domain. 
 Similar findings were already reported in fly\cite{crockerLowAffinityBinding2015}. 
@@
 In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} 
-and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018} 
+and thus many of the 1,639 known human TFs recognize similar motifs\cite{lambertHumanTranscriptionFactors2018}, 
 there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} 
@@
 
-In conclusion, we provide an enhancer code compendium of early human neural and neural crest development. 
-We confirm the paradigm that enhancers consist out of heterotypic clusters of TFBS with soft syntactic rules 
-but also provide a counter example to this paradigm. 
-Furthermore, we validate the biological relevance of neural tube organoids. 
+In conclusion, we provide an enhancer code compendium of early human neuronal and neural crest development. 
+We confirm the paradigm that most human enhancers consist of heterotypic clusters of TFBS with soft syntactic rules 
+but also provide counter exampled to this paradigm. 
+Furthermore, we validate the biological relevance of neural tube organoids, 
+having representative transcriptomes and genomic regulatory code activation compared to embryo development. 
 Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, 
 freely available to the research community, 
-which can be used to predict, 
-interpret and design enhancers underlying early human neural and neural crest development.
-Alongside with these models, we implemented new tools in our genomic deep learning framework CREsted\cite{kempynckCREstedModelingGenomic2025}, specifically for the analysis of enhancer codes underlying complex biological systems.
+that can be used to predict, 
+interpret and design enhancers underlying early human neuronal and neural crest development.
+Alongside these models, we implemented new tools in our genomic deep learning framework CREsted\cite{kempynckCREstedModelingGenomic2025}, specifically for the analysis of compendia of enhancer codes underlying complex biological systems.
 
@@
 
-In this study we focused on predicting TF binding to enhancers, but not so much on their effector function. 
-The effect a TF has on target gene expression depends on its effector domain and interactions with
-co-factors\cite{lambertHumanTranscriptionFactors2018} and this may depend on the sequence context of
+In this study we focused on predicting TF binding to enhancers, without much attention to their effector function.
+The effect of a TF on target gene expression depends on its effector domain and interactions with
+co-factors\cite{lambertHumanTranscriptionFactors2018} and this may also depend on the sequence context of
 the enhancer itself\cite{jacobsWidespreadRegulatorySpecificities2023}. 
@@
 These studies may especially be important to infer repressive roles of TFs on gene expression, 
-an effect that is currently difficult to study with state of the art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
+an effect that is currently difficult to study with state-of-the-art methods\cite{bravogonzalez-blasSCENICSinglecellMultiomic2023, kempynckCREstedModelingGenomic2025,
 dewinterModellingDesignTranscriptional2025}. 
-In particular for the neural tube, it is known that repressive interactions in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. 
-In this study we could only predict a strong repressive effect of ZEB2, a chromatin closing repressor that is used across a variety of biological
+For the neural tube, it is known that repressive interactions in its core GRN are important\cite{delasRepressiveInteractionsGene2020}. 
+In this study we could predict a strong repressive effect of ZEB1/2, a chromatin closing repressor that is used across a variety of biological
 systems\cite{minnoyeCrossspeciesAnalysisEnhancer2020, taskiranCelltypedirectedDesignSynthetic2024}
-(\textbf{Fig. S11}).\par
-
-Furthermore, the expression of a gene is in general regulated by multiple enhancers in the gene's locus\cite{
+(\textbf{Fig. S11}),
+but we anticipate that additional repressor TFs can be added to gene-regulatory models with improved experimental and/or computational techniques in the future.
+\par
+
+Furthermore, the expression of a gene is generally regulated by multiple enhancers in the gene's locus\cite{
   perryShadowEnhancersFoster2010,
@@
 loubiereDevelopmentalHousekeepingTranscriptional2024, thomasEnhancerCooperativityCan2025}.
-In this study we focused on the enhancer code of individual enhancers 
-but not how the signal of multiple enhancers is integrated to control target gene expression. 
+In this study we focused on the enhancer code of individual enhancers without addressing 
+how the signal of multiple enhancers is integrated to control target gene expression. 
 Currently efforts are ongoing to build large scale models for modeling gene expression making use of the
@@
 
-To conclude, we suspect that future models will need to take into account both the effector function of TFs through
+To conclude, we suspect that future models will take into account both the effector function of TFs through
 modeling co-factor interactions and the entire gene's locus (including all enhancers and promoter(s)) to effectively
@@
 \begin{itemize}
-  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorso-ventral UCSC session}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorsal-ventral UCSC session}
   \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}
@@
 \subsubsection*{logistic regression models for each cell states}
-For each cell type, neural progenitors dorso-ventral patterning, neural progenitors anterior- posterior patterning, neurons,
+For each cell type, neural progenitors dorsal-ventral patterning, neural progenitors anterior- posterior patterning, neurons,
 and neural crest, motif scores for all patterns (converted to a PWM) patterns belonging to clusters of the cell type\lq s
@@
     \textbf{a-b}, precision recall curves for classifying neural tube enhancers (left) and facial mesenchyme enhancers (right) using organoid (\textbf{a}) and embryo models (\textbf{b}).
-\textbf{c}, Chromatin accessibility profile of dorso-ventral topics for organoid
+\textbf{c}, Chromatin accessibility profile of dorsal-ventral topics for organoid
     (top) and embryo (bottom) of region chr7:42,058,384-42,246,450; hg38. Location of HS111 enhancer is indicated. Values are
@@
   \caption{
-    \textbf{Linking motif to TF in dorso-ventral neural progenitors}.\\
+    \textbf{Linking motif to TF in dorsal-ventral neural progenitors}.\\
     \textbf{a-h} Scatter plot showing Pearson correlation coefficient of average TF expression and motif importance per
@@
     \textbf{a-b}, Scatter plot showing performance of the logistic regression models as  area under the (AUC) precision recall (PR) curve (\textbf{a}) and receiver operator curve (ROC) (\textbf{b}) for different cell types for organoid (O) and embryo (E) represented by topic numbers.
-Performance is shown to classify the contrast between topics of the same major class (i.e., neuron progenitor dorso-ventral, neuron progenitor anterior-posterior, neural crest and neuron; x-axis) and genome wide classification of consensus peaks as belonging to a certain topic or not.
+Performance is shown to classify the contrast between topics of the same major class (i.e., neuron progenitor dorsal-ventral, neuron progenitor anterior-posterior, neural crest and neuron; x-axis) and genome wide classification of consensus peaks as belonging to a certain topic or not.
     Related to Figure 7.

### 2025-05-22 manuscript.tex
Text changes

@@
 
+% set T1 font encoding so underscores are rendered correctly
+\usepackage[T1]{fontenc}
+
 \makeatletter
@@
 We created a compendium of enhancer codes for all cell types of the human neural tube and neural crest. 
-We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week-post conception 
+We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week post conception 
 human embryo. Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared 
@@
 In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
-a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
+a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
 in direct comparison to an enhancer compendium we generate in parallel for the head of a four-week post conception human 
@@
 Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
-and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and TFAP2 (\textbf{Fig. 2i}). 
+and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
 We tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
@@
 S3).
-In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
+In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{Fig. 3f-h}); 
 PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
@@
 we identified a higher number of FOXA2 sites using FIMO\cite{
-grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025
-} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
+grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
 Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOXA2 instances 
 in specific regions (\textbf{Fig. S4c}). This suggests that FOXA2 by itself can bind to a subset of genomic regions 
-independently of the cell type (i.e., indepdent of cooperating TFS), but mainly to regions with multiple sites, and 
+independently of the cell type (i.e., independent of cooperating TFs), but mainly to regions with multiple sites, and 
 high-affinity sites.
@@
 4a}).
-We first asked if the AP-program is distinct from the DV program. 
+We first asked if the AP-program is distinct from the DV-program. 
 In other words, is AP patterning encoded by the same or different genomic regions as DV patterning? 
@@
 From anterior to posterior this is \textit{TAATTAG}, \textit{TAATCCCT}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
-Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
 Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
-cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
+cooperative TF-binding\cite{mannChapter3Hox2009}. 
 Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2}; 
-and the presence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
+and the presence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig.
+4d-e} and Fig.
+S5), which resembles cooperative binding of Hox proteins
+in \textit{Drosophila}\cite{mannChapter3Hox2009}.
 In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
@@
 with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
-topic (\textbf{Fig. 4d-e} and Fig. S5)\par
+topic (\textbf{Fig. 4d-e} and Fig. S5). The binding specificity of these TFs might be modulated by another co-factor, 
+analogous to Hox-Exd co-binding in fly (termed "latent specificity")\cite{slatteryCofactorBindingEvokes2011}. PBX1-3 are the human orthologs of Exd.\par
 %TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)
@@
 instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}). 
-The importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} 
+The importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/2} 
 and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Fig. S6). 
@@
 
-Recently, a genome wide association study (GWAS) identified 203 single-nucleotide polymorphisms (SNPs)
-that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
+Recently, a genome wide association study (GWAS) identified single-nucleotide polymorphisms (SNPs) that are associated 
+with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
 Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
-we asked whether $DeepNeuralTube$ could explain the effect of some of these variants on enhancer activity.
-For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$
-and calculated the difference in prediction score between both alleles (delta prediction score). 
-At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
-Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
-Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Fig. S7}). 
+we asked whether $DeepNeuralTube$ could explain the effect of some of these variants. 
+We scored both the reference and alternative allele(s) using both $DeepNeuralTube$ models and calculated the difference 
+in prediction score between alleles (delta prediction score). 
+At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs, out a total of 23,701 SNPs associated 
+with facial variation, that are explainable by either of the two models (\textbf{Fig. 5f}). 
+Most of the "explainable" SNPs have a high delta prediction score for either the migratory neural crest class 
+(organoid Topic 59) or the facial mesenchyme class (organoid Topic 58; \textbf{Fig. S7}).
 As an example, we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
@@
 Finally, in the organoids we could identify an additional topic with instances of GATA
-(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Fig. S8).\par
+(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Fig. S8), a marker 
+gene for the induction of V2 interneurons\cite{zhouGATA2RequiredGeneration2000, delbarrioRegulatoryNetworkInvolving2007,
+sagnerEstablishingNeuronalDiversity2019}.\par
 %TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate.
@@
 Thus, additional nucleotides, next to the core \textit{TAAT}, form a TFBS that is both necessary and sufficient for genome wide cell type specificity. 
+This observation is similar to Hox-Exd co-binding, where sequence specificity of HOX proteins is modulated by another 
+co-factor\cite{slatteryCofactorBindingEvokes2011}.
 In general, although there are only around 100 different DNA binding domains across human TFs\cite{lambertHumanTranscriptionFactors2018} 
@@
 there are slight variations in preferential TFBS of TFs of the same family\cite{lambertHumanTranscriptionFactors2018, jolmaDNABindingSpecificitiesHuman2013, dealmeidaDeepSTARRPredictsEnhancer2022} 
-that could be important for cell type specific binding.\par
+that could be important for cell type-specific binding.\par
 
@@
 We confirm the paradigm that most human enhancers consist of heterotypic clusters of TFBS with soft syntactic rules 
-but also provide counter exampled to this paradigm. 
+but also provide counter example to this paradigm. 
 Furthermore, we validate the biological relevance of neural tube organoids, 
@@
 probabilities of the human embryos are available through Zenodo (\href{https://doi.org/10.5281/zenodo.15281826}{10.5281/zenodo.15281826}). 
-$DeepNeuralTube_O$ and $DeepNeural$- $Tube_E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
+$DeepNeuralTube^O$ and $DeepNeural$- $Tube^E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
 Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the UCSC genome browser: 
 \begin{itemize}
-  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorsal-ventral UCSC session}
-  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}
-  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neuron}{neuron UCSC session}
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorsal-ventral UCSC
+    session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural- \%20tube\%20doro\%2Dventral
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural- \\\%20tube\%20neural\%20crest
+  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neuron}{neuron UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural\%20- tube\%20neuron
   \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20anterior%2Dposterior}{anterior-posterior
-      UCSC session}
+      UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neur- al\%20tube\%20anterior\%2Dposterior
 \end{itemize}
@@
 We thank Peter L\"onnerberg for preprocessing the human embryo data and making it available through EGA.
-We also thank Prof. Dr. Tatjana Sauka-Spengler and Dr. Ruth Williams for their great support with helping setting up the chicken embryo enhancer reporter assays.
+We also thank Prof. Dr. Tatjana Sauka-Spengler and Dr. Ruth Williams for their great support with helping to set up the 
+chicken embryo enhancer reporter assays.
 Computing was performed at the Vlaamse Supercomputer Center and VIB datacore. 
-This work was funded by the following grants to 
-S. Aerts , %TODO: Add funding
-PhD fellowship from FWO to S. De Winter (1191323N) and 
-Post doctoral fellowship from ... to C. Mannens (), %TODO: add funding
+This work was funded by the following grants to S. Aerts: ERCAdG (101054387); CZI (DI2-0000000068); SBO (S005024N); FWO 
+(G0I2722N EOS ID 40007513, G094121N, G044124N); and Foundation Against Cancer (2020-1396 \& 2024-140), 
+the following grants to S. Linnarson: Erling-Persson Foundation Human Developmental Cell Atlas grant, Knut and Alice
+Wallenberg Foundation grants 2015.0041, 2018.0172 and 2018.0220, Swedish Foundation for Strategic Research SB16-0065, 
+and EU Horizon2020 BRAINTIME project 874606 and 
+PhD fellowship from FWO to S. De Winter (1191323N).
 

### 2025-05-22 scripts/figure_2/plot_figure.py
Update scripts.

@@ -196,10 +196,8 @@ class Grouper:
                 self._groups.add(_id)
             else:
                 self._grp_to_idx[_id].append(i)
-
     def __repr__(self) -> str:
         return f"Grouper with {len(self._groups)} elements."
-
     def get_groups(self, verbose=True):
         for group in tqdm(self._groups, disable=not verbose):
             yield group, self._grp_to_idx[group]
@@ -337,7 +335,7 @@ def get_start_end_max_stretch(arr):
 
 @cache
 def prepare_vista_data():
-    genome_dir = "/data/projects/c20/sdewin/resources/"
+    genome_dir = "../../../../../resources/"
     hg38 = crested.Genome(
         pathlib.Path(os.path.join(genome_dir, "hg38/hg38.fa")),
         pathlib.Path(os.path.join(genome_dir, "hg38/hg38.chrom.sizes")),
@@ -637,7 +635,7 @@ def prepare_ref_alt_data_vista():
     classes_to_consider = [
         topic_name_to_model_index_organoid(x) for x in neural_tube_topics_organoid
     ]
-    genome_dir = "/data/projects/c20/sdewin/resources/"
+    genome_dir = "../../../../../resources"
     hg38 = crested.Genome(
         pathlib.Path(os.path.join(genome_dir, "hg38/hg38.fa")),
         pathlib.Path(os.path.join(genome_dir, "hg38/hg38.chrom.sizes")),
@@ -781,9 +779,7 @@ def plot_w_cache(key):
             else:
                 ax = plot_func(fig, gs, y, x)
                 ax_cache[key] = ax
-
         return wrapper
-
     return decorator
 
 
@@ -1103,6 +1099,27 @@ def plot_PR_vista_neural_tube(
     ax.set_ylabel("Precision")
     ax.legend()
     return ax
+
+(
+    prediction_scores_organoid_max,
+    prediction_scores_embryo_max,
+    neural_tube_pos_enhancers,
+    facial_mesenchyme_pos_enhancers,
+) = prepare_vista_data()
+
+organoid_topics = np.array([
+    6, 4, 23, 24, 13, 2, 33, 38, 36, 54, 48, 62, 60, 65, 59, 58
+]) - 1
+
+embryo_topics = np.array(
[…]

### 2025-05-22 scripts/figure_2/plot_pr.py

@@ -1,0 +1,220 @@
+
+import os
+import tensorflow as tf
+from sklearn import metrics
+import pickle
+import gzip
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import json
+
+path_to_organoid_model = "../data_prep_new/organoid_data/MODELS/"
+organoid_model = tf.keras.models.model_from_json(
+    open(os.path.join(path_to_organoid_model, "model.json")).read(),
+    custom_objects={"Functional": tf.keras.models.Model},
+)
+organoid_model.load_weights(
+    os.path.join(path_to_organoid_model, "model_epoch_23.hdf5")
+)
+
+path_to_embryo_model = "../data_prep_new/embryo_data/MODELS/"
+embryo_model = tf.keras.models.model_from_json(
+    open(os.path.join(path_to_embryo_model, "model.json")).read(),
+    custom_objects={"Functional": tf.keras.models.Model},
+)
+embryo_model.load_weights(os.path.join(path_to_embryo_model, "model_epoch_36.hdf5"))
+
+
+organoid_non_augmented_data = pickle.load(
+    open("../data_prep_new/organoid_data/MODELS/nonAugmented_data_dict.pkl", "rb")
+)
+
+embryo_non_augmented_data = pickle.load(
+    gzip.open("../data_prep_new/embryo_data/MODELS/nonAugmented_data_dict.pkl.gz", "rb")
+)
+
+y_score_organoid = organoid_model.predict(organoid_non_augmented_data["test_data"])
+y_true_organoid = organoid_non_augmented_data["y_test"]
+
+y_score_embryo = embryo_model.predict(embryo_non_augmented_data["test_data"])
+y_true_embryo = embryo_non_augmented_data["y_test"]
+
+color_dict = json.load(open("../color_maps.json"))
+
+organoid_topics = np.array(
+    [
+        6, 4, 23, 24, 13, 2,
+        33, 38, 36, 54, 48,
+        62, 60, 65, 59, 58
+    ]
+) - 1
+
+colors_organoid = [
+    *np.repeat(color_dict["cell_type_classes"]["neuron"], 6),
+    *np.repeat(color_dict["cell_type_classes"]["progenitor_dv"], 5),
[…]

### 2025-05-22 scripts/figure_3/FOX_general.py

@@ -1,0 +1,831 @@
+
+"""
+
+bedtools intersect \
+    -b ../data_prep_new/validation_data/FOXA2_ChIP/HEPG2_FOXA2_ENCFF466FCB.bed.gz \
+    -a ../data_prep_new/organoid_data/ATAC/topics/training_data/Topic_33.bed -F 0.4 -wa \
+  > organoid_topic_33_w_HEPG2.bed
+
+bedtools intersect \
+    -b ../data_prep_new/validation_data/FOXA2_ChIP/A549_FOXA2_ENCFF686MSH.bed.gz \
+    -a ../data_prep_new/organoid_data/ATAC/topics/training_data/Topic_33.bed -F 0.4 -wa \
+  > organoid_topic_33_w_A549.bed
+
+bedtools intersect \
+    -b ../data_prep_new/validation_data/FOXA2_ChIP/HEPG2_FOXA2_ENCFF466FCB.bed.gz \
+    -a ../data_prep_new/embryo_data/ATAC/topics/training_data/Topic_34.bed -F 0.4 -wa \
+  > embryo_topic_34_w_HEPG2.bed
+
+bedtools intersect \
+    -b ../data_prep_new/validation_data/FOXA2_ChIP/A549_FOXA2_ENCFF686MSH.bed.gz \
+    -a ../data_prep_new/embryo_data/ATAC/topics/training_data/Topic_34.bed -F 0.4 -wa \
+  > embryo_topic_34_w_A549.bed
+
+"""
+
+import numpy as np
+from dataclasses import dataclass
+import h5py
+from typing import Self
+from tqdm import tqdm
+import os
+import pandas as pd
+from crested._genome import Genome
+from crested.utils._seq_utils import one_hot_encode_sequence
+from tangermeme.tools.fimo import fimo
+import matplotlib.pyplot as plt
+from scipy.stats import mannwhitneyu
+import matplotlib
+import modiscolite
+import logomaker
+import seaborn as sns
+
+organoid_dv_topics = np.array([8, 16, 13, 9, 11, 19, 25, 1, 29, 23, 3]) + 25
+
+embryo_dv_topics = np.array([4, 44, 57, 8, 49, 21, 58, 28]) + 30
+
+
+@dataclass
+class Seqlet:
+    contrib_scores: np.ndarray
+    hypothetical_contrib_scores: np.ndarray
+    ppm: np.ndarray
+    start: int
+    end: int
+    region_name: str
+    region_one_hot: np.ndarray
[…]

### 2025-05-22 scripts/figure_3/enhancer_a2.py

@@ -1,0 +1,139 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import logomaker
+import pathlib
+import os
+import crested
+import tensorflow as tf
+import pandas as pd
+import json
+
+genome_dir = "../../../../../resources/"
+hg38 = crested.Genome(
+    pathlib.Path(os.path.join(genome_dir, "hg38/hg38.fa")),
+    pathlib.Path(os.path.join(genome_dir, "hg38/hg38.chrom.sizes")),
+)
+
+path_to_embryo_model = "../data_prep_new/embryo_data/MODELS/"
+embryo_model = tf.keras.models.model_from_json(
+    open(os.path.join(path_to_embryo_model, "model.json")).read(),
+    custom_objects={"Functional": tf.keras.models.Model},
+)
+embryo_model.load_weights(os.path.join(path_to_embryo_model, "model_epoch_36.hdf5"))
+
+path_to_organoid_model = "../data_prep_new/organoid_data/MODELS/"
+organoid_model = tf.keras.models.model_from_json(
+    open(os.path.join(path_to_organoid_model, "model.json")).read(),
+    custom_objects={"Functional": tf.keras.models.Model},
+)
+organoid_model.load_weights(os.path.join(path_to_organoid_model, "model_epoch_23.hdf5"))
+
+enhancer_A2_region = "chr18:77367159-77367659"
+
+attr_a2_organoid, oh_a2_organoid = crested.tl.contribution_scores(
+    input=enhancer_A2_region,
+    target_idx=32,
+    model=organoid_model,
+    genome=hg38
+)
+attr_a2_embryo, oh_a2_embryo = crested.tl.contribution_scores(
+    input=enhancer_A2_region,
+    target_idx=33,
+    model=embryo_model,
+    genome=hg38
+)
+
+p_embryo = crested.tl.predict(enhancer_A2_region, embryo_model, hg38)
+p_organoid = crested.tl.predict(enhancer_A2_region, organoid_model, hg38)
+
+organoid_progenitor_topics_to_show = np.array([33, 38, 36, 54, 48]) - 1
+embryo_progenitor_topics_to_show = np.array([34, 38, 79, 88, 58]) - 1
+
+
+N_PIXELS_PER_GRID = 50
+
+plt.style.use(
[…]

### 2025-05-22 scripts/figure_3/plot_figure.py

@@ -156,7 +156,6 @@ embryo_progenitor_cell_topic.index = [
     for x in embryo_progenitor_cell_topic.index
 ]
 
-
 def rgb_scatter_plot(
     x,
     y,
@@ -218,6 +217,21 @@ def rgb_scatter_plot(
     ax.text(
         0.8, 0.78, b_name, color="blue", ha="left", va="top", transform=ax.transAxes
     )
+    if r_vmin is None:
+        r_vmin = r_values.min()
+    if r_vmax is None:
+        r_vmax = r_values.max()
+    if g_vmin is None:
+        g_vmin = g_values.min()
+    if g_vmax is None:
+        g_vmax = g_values.max()
+    if b_vmin is None:
+        b_vmin = b_values.min()
+    if b_vmax is None:
+        b_vmax = b_values.max()
+    print(f"R: {r_vmin, r_vmax}\nG: {g_vmin, g_vmax}\nB: {b_vmin, b_vmax}")
+
+
 
 
 cell_topic_bin_organoid = pd.read_table(
@@ -754,6 +768,18 @@ organoid_embryo = [
     (54, 88),
     (48, 58),
 ]
+
+from scipy import stats
+
+corrs_patterns = []
+for cluster in selected_clusters:
+    corrs_patterns.append(
+        stats.pearsonr(
+            [cluster_to_topic_to_avg_pattern_organoid[cluster][o].sum() for o, _ in organoid_embryo],
+            [cluster_to_topic_to_avg_pattern_embryo[cluster][e].sum() for _, e in organoid_embryo]
+        ).statistic
+    )
+print(np.mean(corrs_patterns))
 
 n_clusters = len(selected_clusters)
 
@@ -2736,6 +2762,7 @@ for i, cluster in enumerate(tqdm(selected_clusters)):
         if j == 0:
             _ = ax.set_ylabel(f"cluster_{cluster}")
         axs.append(ax)
+    print(f"{YMIN}, {YMAX}")
     for ax, (organoid_topic, embryo_topic) in zip(axs, organoid_embryo):
         _ = ax.set_ylim(YMIN, YMAX)
         _ = ax.set_axis_off()
@@ -3109,6 +3136,28 @@ ax_FOX_logo_organoid.set_xticks(
 ax_FOX_logo_organoid.set_xlim(-0.5, MX + 0.5)
[…]

### 2025-05-22 scripts/figure_4/plot_figure.py

@@ -93,6 +93,20 @@ def rgb_scatter_plot(
     ax.text(
         0.8, 0.78, b_name, color="blue", ha="left", va="top", transform=ax.transAxes
     )
+    if r_vmin is None:
+        r_vmin = r_values.min()
+    if r_vmax is None:
+        r_vmax = r_values.max()
+    if g_vmin is None:
+        g_vmin = g_values.min()
+    if g_vmax is None:
+        g_vmax = g_values.max()
+    if b_vmin is None:
+        b_vmin = b_values.min()
+    if b_vmax is None:
+        b_vmax = b_values.max()
+    print(f"R: {r_vmin, r_vmax}\nG: {g_vmin, g_vmax}\nB: {b_vmin, b_vmax}")
+
 
 
 def topic_name_to_model_index_embryo(t) -> int:
@@ -812,6 +826,11 @@ precision, recall, threshold = precision_recall_curve(
 fpr, tpr, thresholds = roc_curve((y_test > 0).ravel(), reg.predict_proba(X_test)[:, 1])
 
 
+from sklearn.metrics import auc
+
+auc(recall, precision)
+auc(fpr, tpr)
+
 motif_embedding_results = pickle.load(open("draft/motif_embedding.pkl", "rb"))
 
 n_seq = 200
@@ -1252,6 +1271,7 @@ for i, cluster in enumerate(clusters_to_show):
     ax_pattern.set_xticks([])
     ax_pattern.set_yticks([])
 #
+
 ax_tf_gex = fig.add_subplot(gs[0:10, 39-3:])
 sns.heatmap(
     (

### 2025-05-22 scripts/figure_5/plot_figure.py

@@ -96,6 +96,20 @@ def rgb_scatter_plot(
     ax.text(
         0.8, 0.78, b_name, color="blue", ha="left", va="top", transform=ax.transAxes
     )
+    if r_vmin is None:
+        r_vmin = r_values.min()
+    if r_vmax is None:
+        r_vmax = r_values.max()
+    if g_vmin is None:
+        g_vmin = g_values.min()
+    if g_vmax is None:
+        g_vmax = g_values.max()
+    if b_vmin is None:
+        b_vmin = b_values.min()
+    if b_vmax is None:
+        b_vmax = b_values.max()
+    print(f"R: {r_vmin, r_vmax}\nG: {g_vmin, g_vmax}\nB: {b_vmin, b_vmax}")
+
 
 
 def topic_name_to_model_index_organoid(t: str) -> int:
@@ -671,6 +685,26 @@ cluster_to_name = {
 }
 
 organoid_embryo = [(62, 103), (60, 105), (65, 94), (59, None), (58, 91)]
+
+from scipy import stats
+
+corrs_patterns = []
+for cluster in selected_clusters:
+    corrs_patterns.append(
+        stats.pearsonr(
+            [
+                cluster_to_topic_to_avg_pattern_organoid[cluster][o].sum() for o, e in organoid_embryo
+                if not (o is None or e is None)
+            ],
+            [
+                cluster_to_topic_to_avg_pattern_embryo[cluster][e].sum() for o, e in organoid_embryo
+                if not (o is None or e is None)
+            ]
+        ).statistic
+    )
+print(np.mean(corrs_patterns))
+
+
 
 motifs = {
     n: pattern.ppm[range(*pattern.ic_trim(0.2))].T
@@ -1392,9 +1426,11 @@ for i, cluster in enumerate(tqdm(selected_clusters)):
         if j == 0:
             _ = ax.set_ylabel(f"cluster_{cluster}")
         axs.append(ax)
+    print(f"{YMIN}, {YMAX}")
     for ax, (organoid_topic, embryo_topic) in zip(axs, organoid_embryo):
         _ = ax.set_ylim(YMIN, YMAX)
         _ = ax.set_axis_off()
+
 ax_hit_heatmap = fig.add_subplot(gs[0:16, 23:29])
 sns.heatmap(
     hits_merged_organoid_subset_per_seq_and_cluster_max_scaled.loc[
[…]

### 2025-05-22 scripts/figure_6/plot_figure.py

@@ -95,6 +95,19 @@ def rgb_scatter_plot(
     ax.text(
         0.8, 0.78, b_name, color="blue", ha="left", va="top", transform=ax.transAxes
     )
+    if r_vmin is None:
+        r_vmin = r_values.min()
+    if r_vmax is None:
+        r_vmax = r_values.max()
+    if g_vmin is None:
+        g_vmin = g_values.min()
+    if g_vmax is None:
+        g_vmax = g_values.max()
+    if b_vmin is None:
+        b_vmin = b_values.min()
+    if b_vmax is None:
+        b_vmax = b_values.max()
+    print(f"R: {r_vmin, r_vmax}\nG: {g_vmin, g_vmax}\nB: {b_vmin, b_vmax}")
 
 
 def topic_name_to_model_index_organoid(t: str) -> int:
@@ -657,6 +670,26 @@ cluster_to_name = {
 
 organoid_embryo = [(6, 10), (4, 8), (23, 13), (24, 24), (13, 18), (2, 29)]
 
+from scipy import stats
+
+corrs_patterns = []
+for cluster in selected_clusters:
+    corrs_patterns.append(
+        stats.pearsonr(
+            [
+                cluster_to_topic_to_avg_pattern_organoid[cluster][o].sum() for o, e in organoid_embryo
+                if not (o is None or e is None)
+            ],
+            [
+                cluster_to_topic_to_avg_pattern_embryo[cluster][e].sum() for o, e in organoid_embryo
+                if not (o is None or e is None)
+            ]
+        ).statistic
+    )
+print(np.mean(corrs_patterns))
+
+
+
 motifs = {
     n: pattern.ppm[range(*pattern.ic_trim(0.2))].T
     for n, pattern in zip(all_pattern_names, all_patterns)
@@ -1080,6 +1113,7 @@ for i, cluster in enumerate(tqdm(selected_clusters)):
         if j == 0:
             _ = ax.set_ylabel(f"cluster_{cluster}")
         axs.append(ax)
+    print(YMIN, YMAX)
     for ax, (organoid_topic, embryo_topic) in zip(axs, organoid_embryo):
         _ = ax.set_ylim(YMIN, YMAX)
         _ = ax.set_axis_off()
@@ -1155,5 +1189,7 @@ sns.heatmap(
     xticklabels=True,
 )
 fig.tight_layout()
+
+
 fig.savefig("Figure_6.png", transparent=False)
 fig.savefig("Figure_6.pdf")

### 2025-05-22 scripts/figure_7/genome_wide.py

@@ -1,7 +1,4 @@
 import os
-import matplotlib.pyplot as plt
-import matplotlib
-import scanpy as sc
 import pandas as pd
 import numpy as np
 from dataclasses import dataclass
@@ -9,21 +6,20 @@ import h5py
 from typing import Self
 from tqdm import tqdm
 import pickle
-import logomaker
 from functools import reduce
 import torch
-import seaborn as sns
-from scipy.stats import binomtest
 import json
-from matplotlib.lines import Line2D
 import pysam
 from crested.utils._seq_utils import one_hot_encode_sequence
 from sklearn.model_selection import train_test_split
 from sklearn.linear_model import LogisticRegression
 from sklearn.metrics import precision_recall_curve, roc_curve, auc
-from scipy.stats import norm
 from tangermeme.tools.fimo import fimo
 from tangermeme.utils import extract_signal
+import matplotlib.pyplot as plt
+from adjustText import adjust_text
+import matplotlib
+
 
 ##########################################################################################
 #
@@ -32,6 +28,76 @@ from tangermeme.utils import extract_signal
 #
 #
 #########################################################################################
+
+def topic_name_to_model_index_organoid(t: str) -> int:
+    cell_type = t.split("_Topic")[0]
+    topic = int(t.split("_")[-1])
+    if cell_type == "neuron":
+        return topic - 1
+    elif cell_type == "progenitor":
+        return topic - 1 + 25
+    elif cell_type == "neural_crest":
+        return topic - 1 + 55
+    else:
+        raise ValueError(f"{t} unknown.")
+
+
+def model_index_to_topic_name_organoid(i: int) -> str:
+    if i >= 0 and i < 25:
+        cell_type = "neuron"
+        topic_number = i + 1
+    elif i >= 25 and i < 55:
+        cell_type = "progenitor"
+        topic_number = i + 1 - 25
+    elif i >= 55 and i < 65:
+        cell_type = "neural_crest"
+        topic_number = i + 1 - 55
+    elif i >= 65 and i < 75:
+        cell_type = "pluripotent"
+        index_to_topic_mapping = {
+            65: 4,
+            66: 5,
+            67: 12,
+            68: 18,
+            69: 20,
[…]
